### 2018-09-21 Rmd/count_check_MS.Rmd
initial commit = most intro

---
title: "Tools for identifying implausible counts"
author: |
  | Richard J. Telford 
  | Department of Biological Sciences, University of Bergen, Postboks 7803, N-5020 Bergen, Norway
  | email: richard.telford@uib.no
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
output:
  bookdown::pdf_document2:
      latex_engine: xelatex
      number_sections: true
documentclass: article
classoption: a4paper
bibliography: extra/chironomid2.bib
csl: extra/journal-of-paleolimnology.csl
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = FALSE, message = FALSE)

loadd(chiron_clean)
loadd(chiron_air)
loadd(plot_base)
```

# Abstract{-}

**Keywords:** Quantitative reconstructions, transfer functions, performance statistics, chironomids

# Introduction

Six percent of papers published in _Molecular and Cellular Biology_ show evidence of inappropriate image duplication [@Bik2018] either due to error or, more rarely, misconduct. It would be reckless to assume that palaeoecological data does not have equivalent problems: the low rate of retractions in tha ecological and geological literature [@Grieneisen2012] may partially reflect our ability to detect errors and misconduct rather than their prevalence.

Databases

Various tools have been developed to identify questionable data. Deviations from the expected distribution of digits from the Newcomb-Benford law has been used to detect issues with scientific and financial data [@Barabesi2017]. Carlisle [-@Carlisle2017] identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data. Brown et al. [-@Brown2017] develop a simple test, GRIM, of whether means of integer data are consistent with the reported sample size. However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data.

A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often 50 for chironomids and several hundred for pollen and diatoms. This is important because larger count sums are associated with smaller uncertainties, both is in relative abundance of taxa and derived statistics such are transfer function reconstructions. However, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there is an insentive to misreport the minimum count sum. This paper develops a method, related to GRIM [@Brown2017], that can test if the count sums are probably as large as reported.

If the raw assemblage counts are archived, it is trivial to identify undercounts. Unfortunately, many palaeoecologists archive percent data without an indication of the count sum. The problem of estimating the count sum from just the percent data occupies most of this paper.

# Methods

All analyses were done in R version `r getRversion()` [@R]. Ordinations were fitted with vegan version `r packageVersion("vegan")` [@vegan] with square-root transformed assemblage data. Transfer functions were fitted with rioja version `r packageVersion("rioja")` [@rioja] using square-root transformed species data. Some diagnostics were performed using analogue version `r packageVersion("analogue")` [@analogue1].

Simulations ...

Code to replicate all the analyses shown above is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms).

# Results and discussion

# Conclusions

## Acknowledgements{-}

This work was partially funded by the Norwegian Research Council FriMedBio project palaeoDrivers (213607).

## References{-}

### 2019-02-06 Rmd/count_check_MS.Rmd
make it render

@@
 classoption: a4paper
-bibliography: extra/chironomid2.bib
+bibliography: extra/countMS2.bib
 csl: extra/journal-of-paleolimnology.csl
@@
 Six percent of papers published in _Molecular and Cellular Biology_ show evidence of inappropriate image duplication [@Bik2018] either due to error or, more rarely, misconduct.
-It would be reckless to assume that palaeoecological data does not have equivalent problems: the low rate of retractions in tha ecological and geological literature [@Grieneisen2012] may partially reflect our ability to detect errors and misconduct rather than their prevalence.
+It would be reckless to assume that the palaeoecological literature does not have equivalent problems: the low rate of retractions in tha ecological and geological literature [@Grieneisen2012] may partially reflect our ability to detect errors and misconduct rather than their prevalence.
 
-Databases
-
-Various tools have been developed to identify questionable data.
+Several numerical tools have been developed to identify questionable data.
 Deviations from the expected distribution of digits from the Newcomb-Benford law has been used to detect issues with scientific and financial data [@Barabesi2017].
@@
 A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often 50 for chironomids and several hundred for pollen and diatoms.
-This is important because larger count sums are associated with smaller uncertainties, both is in relative abundance of taxa and derived statistics such are transfer function reconstructions.
+This is important because larger count sums are associated with smaller uncertainties, both is in relative abundance of taxa and derived statistics such are transfer function reconstructions [@REFS].
 However, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there is an insentive to misreport the minimum count sum.
-This paper develops a method, related to GRIM [@Brown2017], that can test if the count sums are probably as large as reported.
 
-If the raw assemblage counts are archived, it is trivial to identify undercounts.
-Unfortunately, many palaeoecologists archive percent data without an indication of the count sum.
-The problem of estimating the count sum from just the percent data occupies most of this paper.
+If the genuine assemblage counts are archived, it is trivial to identify undercounts.
+Unfortunately, count data may be falsified so that it appears to meet the reported count sum.
+A more common problem is that, regretably, many palaeoecologists archive percent data without an indication of the count sum.
+This paper develops simple tests that can flag if the count data might have been misrepresented, or if the count sum of percent data is perhaps lower than was been reported.
+
+The key observation that there are many rare taxa in most communities [@Darwin18XX].
+A direct consequence of these rare taxa is that most community or assemblage samples will include taxa represented by a single individual [@Coddlington2009], hereafter singletons, unless the sampling effort is high relative to the species richness.
+Given this, we can estimate the count sum _N_ as $1/p * 100$ where _p_ is the percent abundance of the rarest taxon.
+
+<!-- Since it is not normally possible to estimate the count sum from percent data, we need to use some extra information.   -->
+
+<!-- This paper develops a method, related to GRIM [@Brown2017], that can test if the count sums are probably as large as reported. -->
+
+This paper first estimates the prevalance of microfossil assembages without singletons, and develops additional tests for samples without singletons.
+Next, these tests are applied to percent data where the count sums are known, and then to percent data where the count sums are not known.
+Finally, some caveats and problems are discussed.
 
@@
 
+Data sets were downloaded from the Palaeodata center, Neotoma and Pangaea and other sources.
+A range of microfossil data were sought to allow for difference in typical count sum and species richness.
+This paper does not attempt to be an exhaustive survey of all the data available.
+
 All analyses were done in R version `r getRversion()` [@R].
-Ordinations were fitted with vegan version `r packageVersion("vegan")` [@vegan] with square-root transformed assemblage data.
-Transfer functions were fitted with rioja version `r packageVersion("rioja")` [@rioja] using square-root transformed species data.
-Some diagnostics were performed using analogue version `r packageVersion("analogue")` [@analogue1].
-
-Simulations ...
-
 Code to replicate all the analyses shown above is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms).
 
-# Results and discussion
+# Results
+
+## Prevalence of singletons
+
+Dataset 1 consists of `r #nrow(d1)` chironomid samples from REF [-@REF].
+As the count sums are given for each sample, the data have been backtransformed to count data.
+Each sample `r #stopifnot(FALSE)` has at least one singleton (whole or half chironomid head capsule).
+The mean number of singletons is `r "##MEAN"`.
+Assuming that the number of singletons comes from an approximately Poisson distribution (Fig \@ref(fig:singleton-hist)), the expected proportion of samples without singletons is `r #estnos`.
+
+Dataset 2 is the `r #nrow(d2)` sample North American testate amoeba training set collated by REF [-@REF].
+`r #prop` of the assemblages have at least one singleton (mean `r #mean`).
+In the samples that lack singletons, the taxa are not multiples of the rarest taxon.
+
+Dataset 3 is a `r #nrow(d3)` pollen stratigraphy extracted from Neotoma.
+The associated paper (anonymsed here) reports that count sums are at least ###.
+All counts in dataset 3 sum to ###-###...
+No singletons.
+
+Dataset 1. reconstructing reported percent. Problem with halves.
+
+Dataset 4. Larocque-Tobler et al (2015) originally archived counts
+
+Dataset 5 diatoms sibera
+
+Dataset 6 diatoms Bianca
+
+Dataset 7??? Percent checker
+
+# Discussion
+
+relationship to GRIM
+
+Caveats
+
+decimal places -- precision
+
+Unused/unmerged samples - care taken here
+
+Low diversity
+
+Large counts
+
+Some taxonomic groups have microfossils that often come in groups of attached individuals. For example, for diatoms valves are counted, but these are commonly paired or in long chains.
+
+Assemblages vs communities
+
+Consequences
+
+Small undercounts in a few samples will have minimal impact on the precision of any palaeoenvironmental reconstruction or other statistics derived from the assemblage.
+Consequently, such undercounts can be regarded as bad practice that do not necessarily require correcting.
+
+Substantial or pervasive undercounts will potentially seriously effect the precision of the results.
+Such undercounts might either constitute a data handling error (for example, if samples with low counts were supposed to be merged but that step was forgotton, or using boilerplate method text), or falsification.
+Action to correct the literature is probably required.
+
+Some papers don't report count sums.
+When this important quality metric is omitted, the reader should be able to assume that the standard minimum count sum for the taxonomic group has been used (i.e.
+50 for chironomids, several hundred for pollen and diatoms).
+If the actual count sum is materially below this, then this potentially constitutes falsification by omission [@Fanelli2013].
 
@@
 
+non-negligable fraction of the literature
+
 ## Acknowledgements{-}
 
-This work was partially funded by the Norwegian Research Council FriMedBio project palaeoDrivers (213607).
-
 ## References{-}

### 2019-02-06 make_count_check_ms_with_drake.R

@@ -10,28 +10,29 @@ library("rjt.misc")
 #import scripts
 
 
-
+#drake configuration
+pkgconfig::set_config("drake::strings_in_dots" = "literals")
 
 #construct drake plan
 analyses <- drake_plan(
   
   #import & clean data
+  
 
   #make plots
   
-  #get bibliography
-  biblio = download.file("https://raw.githubusercontent.com/richardjtelford/Zabinskie/master/chironomid.bib", destfile = file_out("Rmd/extra/chironomid.bib")),
-  
   #add extra packages to bibliography
   biblio2 = package_citations(
-    packages = c("vegan", "rioja", "analogue", "palaeoSig"), 
-    old_bib = file_in("Rmd/extra/chironomid.bib"), 
-    new_bib = file_out("Rmd/extra/chironomid2.bib")),
+    packages = c("rioja"), 
+    old_bib = file_in("Rmd/extra/countMS.bib"), 
+    new_bib = file_out("Rmd/extra/countMS2.bib")),
   
   #knit manuscript
-  manuscript = rmarkdown::render(input = knitr_in("Rmd/count_check_MS.Rmd"), output_dir = "./output", output_file = file_out("output/count_check_MS.pdf")),
-  
-  strings_in_dots = "literals"
+  #knit manuscript
+  manuscript = target(
+    command = rmarkdown::render(input = knitr_in("Rmd/count_check_MS.Rmd"), output_dir = "./output", output_file = file_out("output/count_check_MS.pdf")), 
+    trigger = trigger(change =list(biblio2))
+  )
 )
 
 #configure and make drake plan
@@ -39,5 +40,7 @@ config <- drake_config(analyses)
 outdated(config)        # Which targets need to be (re)built?
 make(analyses)          # Build the right things.
 
-#voew dependency graph
+system("evince output/count_check_MS.pdf", wait = FALSE)#display pdf - only linux
+
+#view dependency graph
 vis_drake_graph(config, targets_only = TRUE)

### 2019-04-01 Rmd/count_check_MS.Rmd
US bird data

@@
 knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = FALSE, message = FALSE)
+options(digits = 1)
 
@@
 loadd(plot_base)
+
+loadd(testate_counts)
+loadd(testate_summary)
 ```
@@
 Six percent of papers published in _Molecular and Cellular Biology_ show evidence of inappropriate image duplication [@Bik2018] either due to error or, more rarely, misconduct.
-It would be reckless to assume that the palaeoecological literature does not have equivalent problems: the low rate of retractions in tha ecological and geological literature [@Grieneisen2012] may partially reflect our ability to detect errors and misconduct rather than their prevalence.
+It would be reckless to assume that the palaeoecological literature does not have equivalent problems: the low rate of retractions in the ecological and geological literature [@Grieneisen2012] may partially reflect our limited ability to detect errors and misconduct rather than their prevalence.
 
@@
 Brown et al.
-[-@Brown2017] develop a simple test, GRIM, of whether means of integer data are consistent with the reported sample size.
+[-@Brown2017] developed a simple test, GRIM, of whether means of integer data are consistent with the reported sample size.
 However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data.
@@
 A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often 50 for chironomids and several hundred for pollen and diatoms.
-This is important because larger count sums are associated with smaller uncertainties, both is in relative abundance of taxa and derived statistics such are transfer function reconstructions [@REFS].
-However, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there is an insentive to misreport the minimum count sum.
+This is important because larger count sums are associated with smaller uncertainties, both is in relative abundance of taxa and derived statistics such are transfer function reconstructions [@Heiri2001].
+However, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there is an incentive to misreport the minimum count sum.
 
 If the genuine assemblage counts are archived, it is trivial to identify undercounts.
-Unfortunately, count data may be falsified so that it appears to meet the reported count sum.
-A more common problem is that, regretably, many palaeoecologists archive percent data without an indication of the count sum.
+Unfortunately, count data might be falsified so that it appears to meet the reported count sum.
+A more common problem is that, regrettably, many palaeoecologists archive percent data without an indication of the count sum.
 This paper develops simple tests that can flag if the count data might have been misrepresented, or if the count sum of percent data is perhaps lower than was been reported.
 
-The key observation that there are many rare taxa in most communities [@Darwin18XX].
-A direct consequence of these rare taxa is that most community or assemblage samples will include taxa represented by a single individual [@Coddlington2009], hereafter singletons, unless the sampling effort is high relative to the species richness.
-Given this, we can estimate the count sum _N_ as $1/p * 100$ where _p_ is the percent abundance of the rarest taxon.
+The key insight that allows inference about the count sum is that assemblage data are expected to follow the typical features of a rank abundance curve.
+In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual [@Coddington2009], hereafter singletons, unless the sampling effort is high relative to the species richness.
+In the occasional assemblages without singletons, the greatest common divider should usually be one, especially if species richness is high.
 
-<!-- Since it is not normally possible to estimate the count sum from percent data, we need to use some extra information.   -->
+In general, it is not possible to estimate the sum from which percent were calculated, but given that we expect the rarest taxon to be a singleton, we can estimate the count sum _N_ as $1/p * 100$ where _p_ is the percent abundance of the rarest taxon.
+In cases where the rarest taxon is not a singleton, it is unlikely that all percent are integer multiples of _p_ (which they should be otherwise).
 
@@
 
-This paper first estimates the prevalance of microfossil assembages without singletons, and develops additional tests for samples without singletons.
+This paper aims to test the methods presented above, and discusses some complications and caveats.
+
+<!-- estimates the prevalence of microfossil assemblages without singletons, and develops additional tests for samples without singletons.
 Next, these tests are applied to percent data where the count sums are known, and then to percent data where the count sums are not known.
-Finally, some caveats and problems are discussed.
+-->
 
@@
 
-Data sets were downloaded from the Palaeodata center, Neotoma and Pangaea and other sources.
+Data sets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources.
 A range of microfossil data were sought to allow for difference in typical count sum and species richness.
@@
 Dataset 1 consists of `r #nrow(d1)` chironomid samples from REF [-@REF].
-As the count sums are given for each sample, the data have been backtransformed to count data.
+As the count sums are given for each sample, the data have been back-transformed to count data.
 Each sample `r #stopifnot(FALSE)` has at least one singleton (whole or half chironomid head capsule).
@@
 
-Dataset 2 is the `r #nrow(d2)` sample North American testate amoeba training set collated by REF [-@REF].
-`r #prop` of the assemblages have at least one singleton (mean `r #mean`).
-In the samples that lack singletons, the taxa are not multiples of the rarest taxon.
+Dataset 2 is the `r n_distinct(testate_counts$sampleID)` sample North American testate amoeba training set with available counts collated by Amesbury et al.
+[-@AMESBURY2018].
+The count sums range between `r testate_summary$count_summary$sum %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$sum %>% median()`) tests, and species richness ranges between `r testate_summary$count_summary$n %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$n %>% median()`) species.
+`r testate_summary$prop_singletons * 100`% of the assemblages have at least one singleton (median `r testate_summary$median_singletons`).
+In the samples that lack singletons, `r testate_summary$gcd1 * 100`% have a greatest common divider of one.
+
+#predictors of singletons
+#description of samples with gcd > 1
 
@@
 
-Dataset 5 diatoms sibera
+Dataset 5 diatoms Siberia
 
@@
 Substantial or pervasive undercounts will potentially seriously effect the precision of the results.
-Such undercounts might either constitute a data handling error (for example, if samples with low counts were supposed to be merged but that step was forgotton, or using boilerplate method text), or falsification.
+Such undercounts might either constitute a data handling error (for example, if samples with low counts were supposed to be merged but that step was forgotten, or using boilerplate method text), or falsification.
 Action to correct the literature is probably required.
@@
 
-non-negligable fraction of the literature
+non-negligible fraction of the literature
 
@@
 
+```{r}
+#set digits to default
+options(digits = 7)
+```
+
 ## References{-}

### 2019-04-01 R/download_birds.R

@@ -1,0 +1,108 @@
+#breeding bird survey
+
+bird_plan <- drake_plan(
+  #download data
+  bird_download = {
+    URL <- "ftp://ftpext.usgs.gov/pub/er/md/laurel/BBS/DataFiles/50-StopData/1997ToPresent_SurveyWide/Fifty"
+    1:10 %>% 
+      map(
+        ~download.file(
+          url = paste0(URL, ., ".zip"), 
+          destfile = paste0("./data/birds/Fifty", ., ".zip")
+          )
+        )
+  },
+  ##species list
+  bird_species = read_lines("ftp://ftpext.usgs.gov/pub/er/md/laurel/BBS/DataFiles/SpeciesList.txt", skip = 7)[-2] %>% 
+    paste(collapse = "\n") %>% 
+    read_fwf(col_positions = fwf_widths(widths = c(6, 6, 51, 51, 51, 51, 51, 51, 51))) %>% 
+    set_names(slice(., 1)) %>% 
+    slice(-1) %>%
+    select(-French_Common_Name, -Species) %>%
+    rename(Latin_name = Spanish_Common_Name),
+  
+  #extract bird data
+  bird_data = list.files(path = "data/birds/", pattern = "\\.zip$", full.names = TRUE) %>% 
+    map(read_csv) %>% 
+    map(mutate, statenum  = as.numeric(statenum)) %>% 
+    map(~mutate(., count = rowSums(select(., starts_with("Stop"))))) %>% 
+    map_df(select, -starts_with("Stop")),
+  
+  #aggregate to different taxonomic levels
+  bird_singletons = bird_process(bird_data, bird_species)
+
+)
+
+bird_singletons %>% 
+  group_by(taxonomic_level) %>%   
+  summarise(mean_singletons = mean(singletons == 0), n_singletons = sum(singletons == 0), mean_richness = mean(n_species))
+
+  
+
+cf <- drake_config(bird_plan)
+make(bird_plan)
+
+loadd(bird_data)
+
+
+p1 <- nsf %>%
+  mutate(no_si = singletons == 0) %>% 
+  ggplot(aes(x = n_species, fill = no_si)) +
+  geom_bar()+
+  facet_wrap(~no_si, ncol = 1, scales = "free_y")
+p1
+p1 + aes(x = count_sum) + scale_x_log10()
+
+bird_recount_singletons = bird_data %>% 
+  group_by(RouteDataID) %>% 
+  filter(sum(count) >= 400) %>% 
[…]

### 2019-04-01 R/download_testates.R

@@ -1,0 +1,44 @@
+library("neotoma")
+
+testate_plan <- drake_plan(
+  #get dataset list
+  testate_datasets = get_dataset(datasettype = "testate amoebae surface sample"),
+
+  #download_data
+  testate_data = testate_datasets %>% get_download(), #SLOW
+  
+  #
+  #extract site information
+  testate_meta = testate_datasets %>% 
+    map("site.data") %>% 
+    map_df(as_data_frame) %>% 
+    bind_cols(testate_datasets %>% 
+                map_df("dataset.meta")) %>% 
+    mutate(authors = testate_datasets %>% 
+             map("pi.data") %>% 
+             map("ContactName") %>% 
+             map_chr(paste, collapse = " ")),
+
+  #some sites have only percent data
+  testate_has_percent = testate_data %>%
+    map("taxon.list") %>%
+    map_lgl(~any(.$variable.units == "percent", na.rm = TRUE)),  
+  
+  #get counts
+  testate_counts = testate_data[!testate_has_percent] %>%
+    map(~({
+      cnt <- counts(.)
+      tl <- .$taxon.list %>% filter(variable.units == "NISP")
+      cnt[, names(cnt) %in% tl$taxon.name]
+      })) %>%
+    map_df(gather, key = taxon, value = count, .id = "sampleID") %>% 
+    mutate(sampleID = as.numeric(sampleID))
+
+  
+)
+
+# testate_config <- drake_config(testate_plan)
+# 
+# make(testate_plan, jobs = 2)
+# vis_drake_graph(testate_config, targets_only = TRUE)
+

### 2019-04-01 R/summarise_counts.R

@@ -1,0 +1,51 @@
+loadd(testate_counts)
+
+summarise_counts <- function(counts){#columns sampleID/taxon/count
+  results <- list()
+  
+  results$count_summary <- counts %>% 
+    group_by(sampleID) %>% 
+    summarise(min = min(count), n = n(), sum = sum(count), singletons = sum(count == 1))
+  
+  results$prop_singletons <- results$count_summary %>% 
+    summarise(m = mean(min == 1)) %>% 
+    pull(m)
+
+  results$median_singletons <- results$count_summary %>% 
+    summarise(m = median(singletons)) %>% 
+    pull(m)
+
+    #samples without singletons  
+  
+
+  
+  results$no_singletons <- counts %>% 
+    group_by(sampleID) %>%
+    mutate(m = min(count)) %>%
+    filter(m > 1) %>% #remove samples with singletons
+    summarise(gcd = numbers::mGCD(count)) #find greatest common divisor
+    
+  
+  results$gcd1 <- results$no_singletons %>% 
+    summarise(prop1 = mean(gcd == 1)) %>% 
+    pull(prop1)
+
+  return(results)
+  
+}
+
+# 
+# 
+# loadd(testate_meta)
+# 
+# testate_meta %>% 
+#   filter(!readd(testate_has_percent)) %>% 
+#   left_join(
+#     testate_counts %>% 
+#       mutate(sampleID = as.numeric(sampleID)), 
+#     by =c("dataset.id" = "sampleID")) %>% 
+#   group_by(dataset.id) %>% 
+#   filter(min(count) > 1) %>% 
+#   slice(1)
+#                                                                                                                                                                                                     
+# 

### 2019-04-01 make_count_check_ms_with_drake.R

@@ -8,7 +8,8 @@ library("broom")
 library("rjt.misc")
 
 #import scripts
-
+source("R/download_testates.R")
+source("R/summarise_counts.R")
 
 #drake configuration
 pkgconfig::set_config("drake::strings_in_dots" = "literals")
@@ -16,9 +17,10 @@ pkgconfig::set_config("drake::strings_in_dots" = "literals")
 #construct drake plan
 analyses <- drake_plan(
   
-  #import & clean data
-  
 
+  #run analyses
+  testate_summary = summarise_counts(testate_counts),
+  
   #make plots
   
   #add extra packages to bibliography
@@ -29,16 +31,23 @@ analyses <- drake_plan(
   
   #knit manuscript
   #knit manuscript
-  manuscript = target(
-    command = rmarkdown::render(input = knitr_in("Rmd/count_check_MS.Rmd"), output_dir = "./output", output_file = file_out("output/count_check_MS.pdf")), 
-    trigger = trigger(change =list(biblio2))
-  )
+  manuscript = {
+    file_in("Rmd/extra/countMS2.bib")
+    rmarkdown::render(input = knitr_in("Rmd/count_check_MS.Rmd"), output_dir = "./output", output_file = file_out("output/count_check_MS.pdf"))
+  }
 )
 
+#put plans together
+plans <- bind_rows(testate_plan, analyses)
+
 #configure and make drake plan
-config <- drake_config(analyses)
-outdated(config)        # Which targets need to be (re)built?
-make(analyses)          # Build the right things.
+config <- drake_config(plans)
+
+#set up parallel processing for drake
+future::plan(future::multiprocess) 
+
+#Build the right things
+make(plans, jobs = 2, parallelism = "future")
 
 system("evince output/count_check_MS.pdf", wait = FALSE)#display pdf - only linux
 

### 2019-04-03 R/bird_functions.R
moved bird functions to new file

@@ -1,0 +1,36 @@
+#resample to smaller count sum with multivariate hypergeometric & count singletons
+recount_singletons <- function(counts, k, nrep = 10) {
+  k %>% set_names() %>%
+    map_df(~{
+      y <- extraDistr::rmvhyper(nn = nrep, n = counts, k = .)
+      tibble(no_singletons = rowSums(y == 1), rep = 1:nrep)
+    }, .id = "new_count_sum")
+}
+
+#functions
+#join and process at different taxonomic levels
+
+bird_process <- function(bird_data, bird_species) {
+  joined <- bird_data %>%
+    left_join(bird_species, by = "AOU")
+  
+  out <- bind_rows(
+    species = joined %>% rename(taxon = Latin_name),
+    genus = joined %>% rename(taxon = Genus),
+    family = joined %>% rename(taxon = Family),
+    order = joined %>% rename(taxon = ORDER),
+    .id = "taxonomic_level"
+  ) %>%
+    group_by(RouteDataID, year, taxonomic_level, taxon) %>%
+    summarise(count = sum(count)) %>%
+    group_by(RouteDataID, year, taxonomic_level) %>%
+    summarise(
+      n_taxa = n(),
+      singletons = sum(count == 1),
+      count_sum = sum(count), 
+      GCD = ifelse(singletons == 0 & n_taxa > 1, numbers::mGCD(count), NA)#only find GCD where no singletons for speed ( and > 1 taxa). Using ifelse as it doesn't run TRUE code when it would crash
+    )
+  return(out)
+}
+
+

### 2019-04-03 R/download_birds.R

@@ -1,4 +1,5 @@
 #breeding bird survey
+source("R/bird_functions.R")
 
 bird_plan <- drake_plan(
   #download data
@@ -29,80 +30,46 @@ bird_plan <- drake_plan(
     map_df(select, -starts_with("Stop")),
   
   #aggregate to different taxonomic levels
-  bird_singletons = bird_process(bird_data, bird_species)
+  bird_singletons = bird_process(bird_data, bird_species),
 
-)
-
-bird_singletons %>% 
-  group_by(taxonomic_level) %>%   
-  summarise(mean_singletons = mean(singletons == 0), n_singletons = sum(singletons == 0), mean_richness = mean(n_species))
-
+  bird_singleton_summary = bird_singletons %>% 
+    group_by(taxonomic_level) %>%   
+    summarise(pc_singletons = mean(singletons == 0) * 100, n_singletons = sum(singletons == 0), mean_richness = mean(n_taxa), n_routes = n(), GCD = mean(GCD > 1, na.rm = TRUE) * 100) %>% 
+    mutate(taxonomic_level = factor(taxonomic_level, levels = c("species", "genus", "family", "order"))) %>% 
+    arrange(taxonomic_level),
   
-
-cf <- drake_config(bird_plan)
-make(bird_plan)
-
-loadd(bird_data)
-
-
-p1 <- nsf %>%
-  mutate(no_si = singletons == 0) %>% 
-  ggplot(aes(x = n_species, fill = no_si)) +
-  geom_bar()+
-  facet_wrap(~no_si, ncol = 1, scales = "free_y")
-p1
-p1 + aes(x = count_sum) + scale_x_log10()
-
-bird_recount_singletons = bird_data %>% 
-  group_by(RouteDataID) %>% 
-  filter(sum(count) >= 400) %>% 
-  left_join(bird_species, by = "AOU") %>% 
-  group_by(RouteDataID, ORDER) %>% 
-  summarise(count = sum(count)) %>% 
-  nest() %>% 
-  mutate(recount = map(data, ~recount_singletons(counts = .$count, k = c(10, 20, 30, 50, 100, 200, 300, 400)))) %>% 
-  unnest(recount)
-
-bird_recount_singletons %>% 
-  mutate(new_count_sum = as.numeric(new_count_sum)) %>% 
-  group_by(new_count_sum, rep) %>% 
-  summarise(m = mean(no_singletons == 0)) %>% 
[…]

### 2019-04-03 R/download_chironomids.R
initial commit

@@ -1,0 +1,20 @@
+#chironomids
+
+chironomid_plan <- drake_plan(
+  #data for Last Chance Lake
+  last_chance0 = read.table("https://www1.ncdc.noaa.gov/pub/data/paleo/insecta/chironomidae/northamerica/greenland/last-chance2017chironomid.txt", header = TRUE),
+  
+  last_chance_meta = last_chance0 %>% select(age_calBP, totcaps),
+  last_chance = last_chance0 %>% select(-age_calBP, -totcaps),
+  
+  last_chance_n = estimate_n(spp = last_chance, digits = 2) %>% 
+    bind_cols(last_chance_meta),
+  
+  last_chance_plot = last_chance_n %>%  
+    ggplot(aes(x = totcaps, y = est_n, ymax = est_max, ymin = est_min)) + 
+    geom_point() +
+    geom_errorbar() +
+    geom_abline(aes(slope = sl, intercept = 0, colour = sf, linetype = sf), data = tibble(sl = 1:2, sf = factor(sl)), show.legend = FALSE) +
+    coord_equal() +
+    labs(x = "Reported count sum", y = "Estimated count sum")
+)

### 2019-04-03 Rmd/count_check_MS.Rmd
added chironomids

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = FALSE, message = FALSE)
+knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = TRUE, message = TRUE)
+
+#set ggplot theme
+theme_set(theme_bw(base_size = 12))
 options(digits = 1)
@@
 
-loadd(testate_counts)
-loadd(testate_summary)
+loadd(bird_singleton_summary)
+loadd(testate_counts, testate_summary)
+loadd(last_chance_meta)
 ```
@@
 However, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there is an incentive to misreport the minimum count sum.
+There is a documented case of mis-reported count sums: Larocque-Tobler et al [@Larocque-Tobler2015] reported that their chironomid count sums were at least 50 head capsules, a subsequent corrigendum  [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
+Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.
 
@@
 In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual [@Coddington2009], hereafter singletons, unless the sampling effort is high relative to the species richness.
-In the occasional assemblages without singletons, the greatest common divider should usually be one, especially if species richness is high.
+In the occasional assemblages without singletons, the greatest common divisor should usually be one, especially if species richness is high.
 
@@
 Data sets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources.
-A range of microfossil data were sought to allow for difference in typical count sum and species richness.
+A range of ecological and palaeoecological data were sought to allow for difference in typical count sum and species richness.
 This paper does not attempt to be an exhaustive survey of all the data available.
@@
 
-Dataset 1 consists of `r #nrow(d1)` chironomid samples from REF [-@REF].
-As the count sums are given for each sample, the data have been back-transformed to count data.
-Each sample `r #stopifnot(FALSE)` has at least one singleton (whole or half chironomid head capsule).
-The mean number of singletons is `r "##MEAN"`.
-Assuming that the number of singletons comes from an approximately Poisson distribution (Fig \@ref(fig:singleton-hist)), the expected proportion of samples without singletons is `r #estnos`.
+The first dataset consists of over `r format(floor(bird_singleton_summary$n_routes[1]/1000) * 1000, big.mark = ",")` bird counts from the North American breeding bird survey [@NAmerican_Breeding_Bird2018].
+The vast majority of counts have singletons species (Table \@ref(tab:bird-table))).
+To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the birds counts to progressively lower taxonomic resolution.
+As richness declines, the proportion of counts lacking singletons increases (Table \@ref(tab:bird-table))), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
 
-Dataset 2 is the `r n_distinct(testate_counts$sampleID)` sample North American testate amoeba training set with available counts collated by Amesbury et al.
+The vast majority of counts without singletons have a greatest common divisor of one, except when taxonomic richness is low.
+
+```{r bird-table}
+bird_singleton_summary %>%
+  select(`Taxonomic level` = taxonomic_level, `Percent without singletons` = pc_singletons, `Percent GCD > 1` = GCD, `Mean richness` = mean_richness) %>% 
+  knitr::kable(digits = c(0, 4, 4, 3), caption = "Percent of bird counts without singletons and of greatest common divisor (GCD) greater than one of counts without singletons, at different taxonomic levels.")
+```
+
+To test the sensitivity of the prevalence of counts without singletons, I re-sample the counts with 400 or more observations to smaller count sums using the multivariate hypergeometric distribution in the extraDistr package version `r packageVersion("extraDistr")` [@extraDistr].
+At the order level, the proportion of counts without singletons declines steeply until it reaches a minimum at about 200, thereafter is rises slowly (Fig.
+\@ref(fig:bird-recount)).
+The prevalence of counts without singleton is high for small counts because it is possible that only the common taxa are counted, with larger counts, the chance of counting a single individual of a rare taxon increases.
+With even larger counts the prevalence of counts without singletons increases due to saturation as singletons become doubletons and few new taxa are found.
+
+```{r bird-recount, fig.cap="Effect of count sum on the proportion of samples without singletons.
+Results are the mean of ten trials, error bars are two standard errors.", fig.width = 90/25.4, fig.height = 90/25.4}
+readd(bird_recount_singleton_plot)
+```
+
+The second dataset is the `r n_distinct(testate_counts$sampleID)` sample North American testate amoeba training set with available counts collated by Amesbury et al.
 [-@AMESBURY2018].
@@
 `r testate_summary$prop_singletons * 100`% of the assemblages have at least one singleton (median `r testate_summary$median_singletons`).
-In the samples that lack singletons, `r testate_summary$gcd1 * 100`% have a greatest common divider of one.
+In the assemblages that lack singletons, `r testate_summary$gcd1 * 100`% have a greatest common divider of one.
 
-#predictors of singletons
-#description of samples with gcd > 1
+<!-- description of samples with gcd > 1 -->
+
+## Estimating the percent sum
+
+The third dataset consists of `r nrow(last_chance_meta)` chironomid head capsule counts from Last Chance Lake [@Axford2017].
+It includes both the count sums and the percentage of each taxa, allowing the count sums estimated from the rarest taxon to be checked.
+The percent are given to two decimal places: to account for rounding errors, I use the smallest and largest values consistent with the reported percentage and show this range as a error bar.
+All the estimated count sums are, within error, either identical to the reported count sum or twice as much (Fig.
+\@ref(fig:last-chance-plot)).
+The factor of two difference is because in some counts the rarest taxon is represented by a half head capsule.
+Reporting half microfossils is common for several microfossil groups, including chironomids and pollen (especially for bisaccate conifers), occasionally other fractions are reported.
+Fractional counts make the estimated count too high, so do no risk incorretly flagging counts as being too small.
+
+```{r last-chance-plot, fig.cap = "Estimated and reported chironomid count sums from Axford et al.
+[-@Axford2017].
+Lines show the 1:1 and 2:1 relationships.", fig.width = 90/25.4, fig.height = 90/25.4}
+readd(last_chance_plot)
+```
+
+## Unexpected count data
 
@@
 
-Dataset 1. reconstructing reported percent. Problem with halves.
-
-Dataset 4. Larocque-Tobler et al (2015) originally archived counts
-
+## Unexpected percent data
 Dataset 5 diatoms Siberia
@@
 
+Chao index
 relationship to GRIM
@@
 
+different count sums (eg trees & herbs vs trees herbs and pteridiophytes)
+
 Low diversity 
+Truncated diversity
 
@@
 
-Some taxonomic groups have microfossils that often come in groups of attached individuals. For example, for diatoms valves are counted, but these are commonly paired or in long chains.
+Some taxonomic groups have microfossils that often come in groups of attached individuals. For example, for diatoms valves are counted, but these are commonly paired or, for some taxa, in long chains.
 

### 2019-04-03 make_count_check_ms_with_drake.R

@@ -6,9 +6,12 @@ library("gridExtra")
 library("broom")
 #devtools::install_github("richardjtelford/rjt.misc")
 library("rjt.misc")
+library("countChecker")
 
 #import scripts
+source("R/download_birds.R")
 source("R/download_testates.R")
+source("R/download_chironomids.R")
 source("R/summarise_counts.R")
 
 #drake configuration
@@ -25,7 +28,7 @@ analyses <- drake_plan(
   
   #add extra packages to bibliography
   biblio2 = package_citations(
-    packages = c("rioja"), 
+    packages = c("extraDistr"), 
     old_bib = file_in("Rmd/extra/countMS.bib"), 
     new_bib = file_out("Rmd/extra/countMS2.bib")),
   
@@ -38,7 +41,7 @@ analyses <- drake_plan(
 )
 
 #put plans together
-plans <- bind_rows(testate_plan, analyses)
+plans <- bind_rows(bird_plan, testate_plan, chironomid_plan, analyses)
 
 #configure and make drake plan
 config <- drake_config(plans)

### 2019-04-04 Rmd/count_check_MS.Rmd
pollen in ms

@@
 loadd(last_chance_meta)
+loadd(pollen1_summary)
+
 ```
@@
 
-Dataset 3 is a `r #nrow(d3)` pollen stratigraphy extracted from Neotoma.
-The associated paper (anonymsed here) reports that count sums are at least ###.
-All counts in dataset 3 sum to ###-###...
-No singletons.
+Dataset 3 is a `r n_groups(readd(pollen1_counts))` count pollen stratigraphy extracted from Neotoma.
+The associated paper (anonymised here) reports that count sums are 200 (excluding fern spores).
+Unexpectedly, only `r pollen1_summary$count$has_singletons %>% insist(.
+== 1) %>% as.english()` of the counts has a singleton.
+The other counts have minimum abundances of two or four pollen grains.
+With the exception of the single singleton, all `r pollen1_summary$individual$n %>% invisible()` species counts are divisible by two`r pollen1_summary$individual$not_two %>% insist(.
+== 1) %>% invisible()`.
+The probablility that these counts reflects the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled in some way.
 

### 2019-04-04 R/download_pollen.R

@@ -1,0 +1,31 @@
+#pollen
+
+pollen_plan <- drake_plan(
+ #download
+ pollen1_data = neotoma::get_download(3132),
+ #reformat
+ pollen1_counts = neotoma::counts(pollen1_data)[[1]] %>%
+   as_tibble(rownames = ".id") %>% 
+   gather(key = "taxa", value = "count", -.id) %>%
+   filter(count > 0) %>% 
+   group_by(.id),
+ 
+ #summarise
+ pollen1_summary = list(
+   count = pollen1_counts %>% 
+     summarise(GCD = numbers::mGCD(count), min = min(count), n = n()) %>% 
+     summarise(has_singletons = sum(min == 1), GCD2 = sum(GCD > 1)),
+   individual = pollen1_counts %>% 
+     ungroup() %>%
+     summarise(one = sum(count == 1), two = sum(count %% 2 == 0), n = n(), not_two = n - two)
+ )
+)
+
+
+cf <- drake_config(pollen_plan)
+make(pollen_plan, keep_going = TRUE)
+vis_drake_graph(cf)
+
+
+
+

### 2019-04-04 make_count_check_ms_with_drake.R

@@ -7,11 +7,13 @@ library("broom")
 #devtools::install_github("richardjtelford/rjt.misc")
 library("rjt.misc")
 library("countChecker")
+library("english")
 
 #import scripts
 source("R/download_birds.R")
 source("R/download_testates.R")
 source("R/download_chironomids.R")
+source("R/download_pollen.R")
 source("R/summarise_counts.R")
 
 #drake configuration
@@ -41,7 +43,7 @@ analyses <- drake_plan(
 )
 
 #put plans together
-plans <- bind_rows(bird_plan, testate_plan, chironomid_plan, analyses)
+plans <- bind_rows(bird_plan, testate_plan, chironomid_plan, analyses, pollen_plan)
 
 #configure and make drake plan
 config <- drake_config(plans)
@@ -50,7 +52,7 @@ config <- drake_config(plans)
 future::plan(future::multiprocess) 
 
 #Build the right things
-make(plans, jobs = 2, parallelism = "future")
+make(plans, jobs = 2, parallelism = "future", session = callr::r_vanilla())
 
 system("evince output/count_check_MS.pdf", wait = FALSE)#display pdf - only linux
 

### 2019-04-06 Rmd/count_check_MS.Rmd
yaml

@@
 title: "Tools for identifying implausible counts"
-author: |
-  | Richard J. Telford 
-  | Department of Biological Sciences, University of Bergen, Postboks 7803, N-5020 Bergen, Norway
-  | email: richard.telford@uib.no
+author:
+  - name: Richard J. Telford
+    email: richard.telford@uib.no
+    affiliation: University of Bergen
+    footnote: Corresponding Author
+address:
+  - code: University of Bergen
+    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Postboks 7803, N-5020 Bergen, Norway
+abstract: |
+  Astract goes here
+journal: "Quaternary Science Reviews"
 date: "`r format(Sys.Date(), '%d %B %Y')`"
@@
 output:
-  bookdown::pdf_document2:
-      latex_engine: xelatex
+  bookdown::pdf_book:
+    base_format: 
+      rticles::elsevier_article
     number_sections: true
@@
 csl: extra/journal-of-paleolimnology.csl
-toc: false
 ---
@@
 ```
-
-# Abstract{-}
 
@@
 A range of ecological and palaeoecological data were sought to allow for difference in typical count sum and species richness.
+Datasets with possible data mishandling are anonymised.
 This paper does not attempt to be an exhaustive survey of all the data available.
@@
 
-Dataset 3 is a `r n_groups(readd(pollen1_counts))` count pollen stratigraphy extracted from Neotoma.
-The associated paper (anonymised here) reports that count sums are 200 (excluding fern spores).
+Dataset 3 is a `r n_groups(readd(pollen1_counts))` count pollen stratigraphy.
+The associated paper reports that count sums are 200 (excluding fern spores).
 Unexpectedly, only `r pollen1_summary$count$has_singletons %>% insist(.

### 2019-04-06 R/download_pollen.R
secrets

@@ -2,7 +2,13 @@
 
 pollen_plan <- drake_plan(
  #download
- pollen1_data = neotoma::get_download(3132),
+ pollen1_data = {
+   target <- secrets %>% 
+     filter(dataset == "pollen1") %>% 
+     pull(secret) %>% 
+     as.numeric()
+   neotoma::get_download(target)
+   },
  #reformat
  pollen1_counts = neotoma::counts(pollen1_data)[[1]] %>%
    as_tibble(rownames = ".id") %>% 

### 2019-04-06 make_count_check_ms_with_drake.R

@@ -9,6 +9,9 @@ library("rjt.misc")
 library("countChecker")
 library("english")
 
+#drake configuration
+pkgconfig::set_config("drake::strings_in_dots" = "literals")
+
 #import scripts
 source("R/download_birds.R")
 source("R/download_testates.R")
@@ -16,12 +19,13 @@ source("R/download_chironomids.R")
 source("R/download_pollen.R")
 source("R/summarise_counts.R")
 
-#drake configuration
-pkgconfig::set_config("drake::strings_in_dots" = "literals")
 
 #construct drake plan
 analyses <- drake_plan(
   
+  #secrets
+  secrets = readRDS("data/secrets.RDS") %>%
+      encryptr::decrypt(secret),
 
   #run analyses
   testate_summary = summarise_counts(testate_counts),

### 2019-04-07 make_count_check_ms_with_drake.R
askpass

@@ -12,6 +12,9 @@ library("english")
 #drake configuration
 pkgconfig::set_config("drake::strings_in_dots" = "literals")
 
+## askpass
+options(askpass = function(x)readLines("pw.txt"))
+
 #import scripts
 source("R/download_birds.R")
 source("R/download_testates.R")
@@ -42,7 +45,10 @@ analyses <- drake_plan(
   #knit manuscript
   manuscript = {
     file_in("Rmd/extra/countMS2.bib")
-    rmarkdown::render(input = knitr_in("Rmd/count_check_MS.Rmd"), output_dir = "./output", output_file = file_out("output/count_check_MS.pdf"))
+    rmarkdown::render(
+      input = knitr_in("Rmd/count_check_MS.Rmd"), 
+      knit_root_dir = "../", 
+      clean = FALSE)
   }
 )
 
@@ -56,9 +62,9 @@ config <- drake_config(plans)
 future::plan(future::multiprocess) 
 
 #Build the right things
-make(plans, jobs = 2, parallelism = "future", session = callr::r_vanilla())
+make(plans, jobs = 2, parallelism = "future")
 
-system("evince output/count_check_MS.pdf", wait = FALSE)#display pdf - only linux
+system("evince Rmd/count_check_MS.pdf", wait = FALSE)#display pdf - only linux
 
 #view dependency graph
 vis_drake_graph(config, targets_only = TRUE)

### 2019-04-07 R/download_pollen.R
more summary options

@@ -19,18 +19,25 @@ pollen_plan <- drake_plan(
  #summarise
  pollen1_summary = list(
    count = pollen1_counts %>% 
-     summarise(GCD = numbers::mGCD(count), min = min(count), n = n()) %>% 
-     summarise(has_singletons = sum(min == 1), GCD2 = sum(GCD > 1)),
+     summarise(GCD = numbers::mGCD(count), min = min(count), n = n(), sum = sum(count)) %>% 
+     summarise(has_singletons = sum(min == 1), GCD2 = sum(GCD > 1), GCD4 = sum(GCD ==4), min_count = min(sum), max_count = max(sum), max2_count = max(sum[-which.max(sum)])),
+   
    individual = pollen1_counts %>% 
      ungroup() %>%
-     summarise(one = sum(count == 1), two = sum(count %% 2 == 0), n = n(), not_two = n - two)
+     summarise(one = sum(count == 1), two = sum(count %% 2 == 0), n = n(), not_two = n - two),
+   
+   max_trilete =  pollen1_counts %>% 
+     filter(taxa == "Unknown (trilete)") %>% 
+     ungroup() %>% 
+     summarise(m = max(count)) %>% 
+     pull(m)
  )
 )
 
 
-cf <- drake_config(pollen_plan)
-make(pollen_plan, keep_going = TRUE)
-vis_drake_graph(cf)
+# cf <- drake_config(pollen_plan)
+# make(pollen_plan, keep_going = TRUE)
+# vis_drake_graph(cf)
 
 
 

### 2019-04-07 Rmd/count_check_MS.Rmd
line returns after every sentence

@@
 Carlisle [-@Carlisle2017] identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data. 
-Brown et al.
-[-@Brown2017] developed a simple test, GRIM, of whether means of integer data are consistent with the reported sample size.
+Brown et al. [-@Brown2017] developed a simple test, GRIM, of whether means of integer data are consistent with the reported sample size. 
 However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data.
@@
 However, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there is an incentive to misreport the minimum count sum. 
-There is a documented case of mis-reported count sums: Larocque-Tobler et al [@Larocque-Tobler2015] reported that their chironomid count sums were at least 50 head capsules, a subsequent corrigendum  [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
+The risk that count sums will be mis-reported is not just theoretical: Larocque-Tobler et al [@Larocque-Tobler2015] reported that their chironomid count sums were at least 50 head capsules, a subsequent corrigendum  [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
 Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.
@@
 
-<!-- This paper develops a method, related to GRIM [@Brown2017], that can test if the count sums are probably as large as reported. -->
-
-This paper aims to test the methods presented above, and discusses some complications and caveats.
-
-<!-- estimates the prevalence of microfossil assemblages without singletons, and develops additional tests for samples without singletons.
-Next, these tests are applied to percent data where the count sums are known, and then to percent data where the count sums are not known.
--->
+This paper aims to test the methods presented above and present some cases with unexpected results. 
+Some complications and caveats are discussed.
 
@@
 
-Data sets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources.
+Datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources. 
 A range of ecological and palaeoecological data were sought to allow for difference in typical count sum and species richness. 
-Datasets with possible data mishandling are anonymised.
+Datasets with possible data mis-reporting are anonymised, but no attempt is made to diagnose whether error or misconduct are responsible.
 This paper does not attempt to be an exhaustive survey of all the data available.
@@
 To test the sensitivity of the prevalence of counts without singletons, I re-sample the counts with 400 or more observations to smaller count sums using the multivariate hypergeometric distribution in the extraDistr package version `r packageVersion("extraDistr")` [@extraDistr].
-At the order level, the proportion of counts without singletons declines steeply until it reaches a minimum at about 200, thereafter is rises slowly (Fig.
-\@ref(fig:bird-recount)).
+At the order level, the proportion of counts without singletons declines steeply until it reaches a minimum at about 200, thereafter is rises slowly (Fig. \@ref(fig:bird-recount)). 
 The prevalence of counts without singleton is high for small counts because it is possible that only the common taxa are counted, with larger counts, the chance of counting a single individual of a rare taxon increases.
@@
 
-The second dataset is the `r n_distinct(testate_counts$sampleID)` sample North American testate amoeba training set with available counts collated by Amesbury et al.
-[-@AMESBURY2018].
+The second dataset is the `r n_distinct(testate_counts$sampleID)` sample North American testate amoeba training set with available counts collated by Amesbury et al. [-@AMESBURY2018]. 
 The count sums range between `r testate_summary$count_summary$sum %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$sum %>% median()`) tests, and species richness ranges between `r testate_summary$count_summary$n %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$n %>% median()`) species.
@@
 The percent are given to two decimal places: to account for rounding errors, I use the smallest and largest values consistent with the reported percentage and show this range as a error bar. 
-All the estimated count sums are, within error, either identical to the reported count sum or twice as much (Fig.
-\@ref(fig:last-chance-plot)).
+All the estimated count sums are, within error, either identical to the reported count sum or twice as much (Fig. \@ref(fig:last-chance-plot)). 
 The factor of two difference is because in some counts the rarest taxon is represented by a half head capsule. 
 Reporting half microfossils is common for several microfossil groups, including chironomids and pollen (especially for bisaccate conifers), occasionally other fractions are reported.
-Fractional counts make the estimated count too high, so do no risk incorretly flagging counts as being too small.
+Fractional counts make the estimated count too high, so do no risk incorrectly flagging counts as being too small.
 
@@
 
-Dataset 3 is a `r n_groups(readd(pollen1_counts))` count pollen stratigraphy.
-The associated paper reports that count sums are 200 (excluding fern spores).
-Unexpectedly, only `r pollen1_summary$count$has_singletons %>% insist(.
-== 1) %>% as.english()` of the counts has a singleton.
+Dataset 4 is a `r n_groups(readd(pollen1_counts))` count pollen stratigraphy.
+The associated paper reports that count sums are 200 (excluding fern and fungal spores).
+The archived data have counts sums between `r pollen1_summary$count$min_count` and `r pollen1_summary$count$max2_count`, with the exception of one count of `r pollen1_summary$count$max_count` (which may be so high because of a possible data entry error).
+However, these counts sums include trilete spores, probably from a fern or fern ally, with an abundance as high as `r pollen1_summary$max_trilete`.
+Unexpectedly, only `r pollen1_summary$count$has_singletons %>% insist(. == 1) %>% as.english() %>% as.character()` of the counts has a singleton.
 The other counts have minimum abundances of two or four pollen grains.
@@
 == 1) %>% invisible()`.
-The probablility that these counts reflects the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled in some way.
+The probability that these counts reflects the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled in some way.
 
@@
 
-Some taxonomic groups have microfossils that often come in groups of attached individuals. For example, for diatoms valves are counted, but these are commonly paired or, for some taxa, in long chains.
+Some taxonomic groups have microfossils that often come in groups of attached individuals. 
+For example, for diatoms valves are counted, but these are commonly paired or, for some taxa, in long chains.
 

### 2019-04-27 Rmd/count_check_MS.Rmd
after stockholm

@@
 ---
-title: "Tools for identifying implausible counts"
+title: "Tools for identifying unexpectely low microfossil counts"
 author:
@@
   - code: University of Bergen
-    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Postboks 7803, N-5020 Bergen, Norway
+    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
 abstract: |
-  Astract goes here
+  Microfossil counts are a key data type in palaeoecology.
+Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so.
+This paper introduces methods that can flag assemblages with potentially misreported count sums and finds that some assemblage datasets appear to fail these tests.
 journal: "Quaternary Science Reviews"
@@
 
-loadd(chiron_clean)
-loadd(chiron_air)
-loadd(plot_base)
-
 loadd(bird_singleton_summary)
-loadd(testate_counts, testate_summary)
-loadd(last_chance_meta)
-loadd(pollen1_summary)
+loadd(testate_summ, testate_summary)
+loadd(last_chance_summ, pollen1_summary)
+loadd(owens_summ, owens_summary2)
+loadd(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ)
 
@@
 
-**Keywords:** Quantitative reconstructions, transfer functions, performance statistics, chironomids
+**Keywords:** Microfossil counts; quantitative methods
 
@@
 Several numerical tools have been developed to identify questionable data.
-Deviations from the expected distribution of digits from the Newcomb-Benford law has been used to detect issues with scientific and financial data [@Barabesi2017]. 
+Deviations from the distribution of digits expected from the Newcomb-Benford law has been used to detect issues with scientific and financial data [@Barabesi2017].
 Carlisle [-@Carlisle2017] identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data.
-Brown et al. [-@Brown2017] developed a simple test, GRIM, of whether means of integer data are consistent with the reported sample size. 
+Brown et al. [-@Brown2017] developed the granularity-related inconsistency means test (GRIM) of whether means of integer data are consistent with the reported sample size.
 However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data.
 
-A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often 50 for chironomids and several hundred for pollen and diatoms.
-This is important because larger count sums are associated with smaller uncertainties, both is in relative abundance of taxa and derived statistics such are transfer function reconstructions [@Heiri2001].
-However, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there is an incentive to misreport the minimum count sum. 
-The risk that count sums will be mis-reported is not just theoretical: Larocque-Tobler et al [@Larocque-Tobler2015] reported that their chironomid count sums were at least 50 head capsules, a subsequent corrigendum  [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
+A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often fifty for chironomids and several hundred for pollen and diatoms.
+This is important because larger count sums are associated with smaller uncertainties, both in relative abundance of taxa and derived statistics such are transfer function reconstructions [@Heiri2001].
+However, mistakes happen and metadata such as count sums can be forgotten once percentages are calculated.
+In addition, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there may be an incentive to misreport the minimum count sum.
+The risk that count sums will be mis-reported is not just theoretical: Larocque-Tobler et al [-@Larocque-Tobler2015] reported that their chironomid count sums were at least fifty head capsules, a subsequent corrigendum  [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
 Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.
 
-If the genuine assemblage counts are archived, it is trivial to identify undercounts. 
+If the genuine assemblage counts are archived for all taxa, it is trivial to identify undercounts.
 Unfortunately, count data might be falsified so that it appears to meet the reported count sum.
 A more common problem is that, regrettably, many palaeoecologists archive percent data without an indication of the count sum.
-This paper develops simple tests that can flag if the count data might have been misrepresented, or if the count sum of percent data is perhaps lower than was been reported.
+This paper develops simple tests that can flag if the count data might have been misrepresented, or if the count sum of percent data is perhaps lower than reported.
 
 The key insight that allows inference about the count sum is that assemblage data are expected to follow the typical features of a rank abundance curve.
-In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual [@Coddington2009], hereafter singletons, unless the sampling effort is high relative to the species richness.
+In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual [@Coddington2009], hereafter singletons, unless the sampling effort is high relative to the taxonomic richness.
 In the occasional assemblages without singletons, the greatest common divisor should usually be one, especially if species richness is high.
 
-In general, it is not possible to estimate the sum from which percent were calculated, but given that we expect the rarest taxon to be a singleton, we can estimate the count sum _N_ as $1/p * 100$ where _p_ is the percent abundance of the rarest taxon.
-In cases where the rarest taxon is not a singleton, it is unlikely that all percent are integer multiples of _p_ (which they should be otherwise).
+In general, it is not possible to estimate the sum from which percent were calculated, but the properties of community and assemblage counts make it possible.
+Given that we expect the rarest taxon to be a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
+This method will fail for assemblages without singletons.
+We also expect the percent _p_ to be calculated from integer counts, therefore, there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
+Possible values for _N_ can be found by a direct search algorithm over the range of plausible values of _N_.
+An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
+These tests can be viewed as special cases of GRIM [@Brown2017].
 
@@
 
-Datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources. 
+Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources.
 A range of ecological and palaeoecological data were sought to allow for difference in typical count sum and species richness.
-Datasets with possible data mis-reporting are anonymised, but no attempt is made to diagnose whether error or misconduct are responsible.
+Datasets where the minimum count sum was not reported in the associated paper were excluded.
+Datasets from Lakes Zabinskie, Silvaplana and Seebergsee discussed by Telford [-@Telford2019] as possibly having under-reported count sums were also excluded to avoid double reporting.
+Datasets with possible mis-reporting are anonymised, but no attempt is made to diagnose whether errors or misconduct are responsible.
 This paper does not attempt to be an exhaustive survey of all the data available.
 
-All analyses were done in R version `r getRversion()` [@R].
+All analyses were done in R version `r getRversion()` [@R] and used the packages extraDistr version `r packageVersion("extraDistr")` [@extraDistr], numbers version `r packageVersion("numbers")` [@numbers], and countChecker `r packageVersion("countChecker")` [@countChecker].
 Code to replicate all the analyses shown above is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms).
@@
 
-The first dataset consists of over `r format(floor(bird_singleton_summary$n_routes[1]/1000) * 1000, big.mark = ",")` bird counts from the North American breeding bird survey [@NAmerican_Breeding_Bird2018].
-The vast majority of counts have singletons species (Table \@ref(tab:bird-table))). 
-To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the birds counts to progressively lower taxonomic resolution. 
-As richness declines, the proportion of counts lacking singletons increases (Table \@ref(tab:bird-table))), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
-
+The a tiny minority of the over `r format(floor(bird_singleton_summary$n_routes[1]/1000) * 1000, big.mark = ",")` bird counts from the North American breeding bird survey [@NAmerican_Breeding_Bird2018]
+lack singletons at the species level (Table \@ref(tab:bird-table))).
+To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the birds counts to progressively lower taxonomic resolutions.
+As richness declines, the proportion of counts lacking singletons increases (Table \@ref(tab:bird-table)), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
 The vast majority of counts without singletons have a greatest common divisor of one, except when taxonomic richness is low.
@@
 bird_singleton_summary %>%
-  select(`Taxonomic level` = taxonomic_level, `Percent without singletons` = pc_singletons, `Percent GCD > 1` = GCD, `Mean richness` = mean_richness) %>%
+  select(`Taxonomic level` = taxonomic_level, `Percent without singletons` = pc_singletons, `Percent GCD > 1` = GCD2plus, `Mean richness` = mean_richness) %>%
   knitr::kable(digits = c(0, 4, 4, 3), caption = "Percent of bird counts without singletons and of greatest common divisor (GCD) greater than one of counts without singletons, at different taxonomic levels.")
@@
 
-To test the sensitivity of the prevalence of counts without singletons, I re-sample the counts with 400 or more observations to smaller count sums using the multivariate hypergeometric distribution in the extraDistr package version `r packageVersion("extraDistr")` [@extraDistr].
+To test the sensitivity of the prevalence of counts without singletons to the count sum, I re-sample the counts with 400 or more observations to smaller count sums using a multivariate hypergeometric distribution.
 At the order level, the proportion of counts without singletons declines steeply until it reaches a minimum at about 200, thereafter is rises slowly (Fig. \@ref(fig:bird-recount)).
-The prevalence of counts without singleton is high for small counts because it is possible that only the common taxa are counted, with larger counts, the chance of counting a single individual of a rare taxon increases.
+The prevalence of counts without singletons is high for small counts because it is possible that only the common taxa are counted; with larger counts, the chance of counting a single individual of a rare taxon increases.
 With even larger counts the prevalence of counts without singletons increases due to saturation as singletons become doubletons and few new taxa are found.
 
-```{r bird-recount, fig.cap="Effect of count sum on the proportion of samples without singletons.
+```{r bird-recount, fig.cap="Effect of count sum on the proportion of bird counts without singleton orders.
 Results are the mean of ten trials, error bars are two standard errors.", fig.width = 90/25.4, fig.height = 90/25.4}
@@
 
-The second dataset is the `r n_distinct(testate_counts$sampleID)` sample North American testate amoeba training set with available counts collated by Amesbury et al. [-@AMESBURY2018]. 
-The count sums range between `r testate_summary$count_summary$sum %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$sum %>% median()`) tests, and species richness ranges between `r testate_summary$count_summary$n %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$n %>% median()`) species.
+The `r n_distinct(readd(owens)$SampleId)` diatoms counts from Owen's Lake span a wide range of count sums (`r range(owens_summ$sum) %>% paste(collapse = "--")`) as, if concentrations were low, the requirement that at least 300 valves were counted, was replaced by a minimum area of slide to count [@Platt1997].
+Discounting the `r sum(owens_summ$n_taxa == 1)` mono-specific assemblages (all with very low count sums`r max(owens_summ$sum[owens_summ$n_taxa == 1]) %>% insist(.
+<= 10) %>% invisible()`), the median number of singletons per count is `r owens_summary2$median_sing %>% as.english()` and only `r owens_summary2$no_singletons`% of counts lack singletons.
+Assemblages with counts below 50 have a higher chance of lacking singletons (`r owens_summary2$no_single_lt_50`% vs `r owens_summary2$no_single50_`%).
+Of the assemblages without singletons, `r owens_summary2$gcd2`% have a greatest common divisor of one.
+The maximum greatest common divisor is `r  owens_summary2$gcd_max %>% as.english()`.
+Assemblages with a greatest common divisor above one all have low taxonomic richness`r owens_summ %>% filter(gcd > 1) %>% pull(n_taxa) %>% max() %>%  insist(. <= 6) %>% invisible()`.
+
+The `r testate_summ$nsamples` assemblages in the North American testate amoeba training set with available count data [@AMESBURY2018] have count sums between `r testate_summary$count_summary$sum %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$sum %>% median()`) tests, and taxonomic richness ranges between `r testate_summary$count_summary$n_taxa %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$n_taxa %>% median()`) taxa.
 `r testate_summary$prop_singletons * 100`% of the assemblages have at least one singleton (median `r testate_summary$median_singletons`).
-In the assemblages that lack singletons, `r testate_summary$gcd1 * 100`% have a greatest common divider of one.
-
-<!-- description of samples with gcd > 1 -->
+In the assemblages that lack singletons, `r testate_summary$gcd1 * 100`% have a greatest common divisor of one.
+Of these assemblages, `r testate_summ$n_gcd2_low_div`% have fewer than five taxa; `r testate_summ$n_gcd2 %>% as.english()` have a greatest common divisor of two, and the other `r testate_summ$n_gcd3 %>% as.english()` has a greatest common divisor of three.
 
@@
 
-The third dataset consists of `r nrow(last_chance_meta)` chironomid head capsule counts from Last Chance Lake [@Axford2017]. 
-It includes both the count sums and the percentage of each taxa, allowing the count sums estimated from the rarest taxon to be checked. 
-The percent are given to two decimal places: to account for rounding errors, I use the smallest and largest values consistent with the reported percentage and show this range as a error bar. 
-All the estimated count sums are, within error, either identical to the reported count sum or twice as much (Fig. \@ref(fig:last-chance-plot)). 
-The factor of two difference is because in some counts the rarest taxon is represented by a half head capsule. 
-Reporting half microfossils is common for several microfossil groups, including chironomids and pollen (especially for bisaccate conifers), occasionally other fractions are reported.
-Fractional counts make the estimated count too high, so do no risk incorrectly flagging counts as being too small.
+The `r last_chance_summ$nsamples` chironomid head capsule assemblage dataset from Last Chance Lake [@Axford2017] includes both the count sums and the percentage of each taxa, allowing the count sums estimated from the rarest taxon to be checked.
+The percentages are given to two decimal places: to account for rounding errors, the countChecker package uses the smallest and largest values consistent with the reported percentage.
+With the minimum percent method the estimated count sums are, within error, either identical to the reported count sum or twice as much (Fig. \@ref(fig:last-chance-plot)).
+With the direct search method, the estimated count sums are all exactly twice the reported count sum.
+The factor of two difference is because all the counts include half head capsules (but only sometimes is the rarest taxon represented by a half head capsule).
+Reporting half microfossils is common for several microfossil groups, including chironomids and pollen (especially for bisaccate conifers); occasionally other fractions are reported.
+Fractional counts make the estimated count too high by a factor of two, so do not risk incorrectly flagging counts as being too small.
+The direct search method is more precise than the minimum percent method as the rounding error is relatively smaller on the larger percent values the method uses.
 
-```{r last-chance-plot, fig.cap = "Estimated and reported chironomid count sums from Axford et al.
-[-@Axford2017].
-Lines show the 1:1 and 2:1 relationships.", fig.width = 90/25.4, fig.height = 90/25.4}
+```{r last-chance-plot, fig.cap = "Estimated and reported chironomid count sums by the minumum percent method (solid symbols with error bars) and the direct search method (red crosses) from Last Chance Lake [@Axford2017].
+Lines show the 1:1 (solid) and 2:1 (dashed) relationships.", fig.width = 90/25.4, fig.height = 90/25.4}
 readd(last_chance_plot)
@@
 
-Dataset 4 is a `r n_groups(readd(pollen1_counts))` count pollen stratigraphy.
+Dataset pollen1 is a `r n_groups(readd(pollen1_counts))` count pollen stratigraphy.
+The published stratigraphy and the archived dataset generally resemble each other, but there are several discrepancies.
 The associated paper reports that count sums are 200 (excluding fern and fungal spores).
 The archived data have counts sums between `r pollen1_summary$count$min_count` and `r pollen1_summary$count$max2_count`, with the exception of one count of `r pollen1_summary$count$max_count` (which may be so high because of a possible data entry error).
-However, these counts sums include trilete spores, probably from a fern or fern ally, with an abundance as high as `r pollen1_summary$max_trilete`.
-Unexpectedly, only `r pollen1_summary$count$has_singletons %>% insist(. == 1) %>% as.english() %>% as.character()` of the counts has a singleton.
+However, these count sums include trilete spores, probably from a fern or fern ally, with an abundance as high as `r pollen1_summary$max_trilete`.
+Unexpectedly, only `r pollen1_summary$count$has_singletons %>% insist(. == 1) %>% as.english()` of the counts has a singleton.
 The other counts have minimum abundances of two or four pollen grains.
-With the exception of the single singleton, all `r pollen1_summary$individual$n %>% invisible()` species counts are divisible by two`r pollen1_summary$individual$not_two %>% insist(.
-== 1) %>% invisible()`.
-The probability that these counts reflects the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled in some way.
+With the exception of the single singleton, all `r pollen1_summary$individual$n` species counts are divisible by two`r pollen1_summary$individual$not_two %>% insist(. == 1) %>% invisible()`.
+The probability that these counts reflect the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled in some way.
 
 ## Unexpected percent data
-Dataset 5 diatoms Siberia
+```{r line169}
 
-Dataset 6 diatoms Bianca
+```
 
-Dataset 7??? Percent checker
+Dataset diatom1 includes `r diatom1_summ$nsamples` diatom assemblages with a median richness of `r diatom1_summ$median_taxa` taxa.
+The associated paper reports that at least 400 valves were counted from each assemblage, which means that singletons should have a relative abundance of 0.25% or less.
+However, the rarest taxon in two assemblages have a relative abundance of `r diatom1_summ$maxpc`%, and the relative abundance of all other taxa in these assemblages are, within rounding error, integer multiples of this`r diatom1_summ$prop_integer %>% insist(.
+== 1) %>% invisible()`.
+If the count sums actually are at least 400, this would imply that each species count in these assemblages is an integer multiple of at least `r diatom1_summ$integer_mult`.
+This seems unlikely.
+Alternatively, the actual count sum for these assemblages could be as low as `r diatom1_summ$low_n` valves.
+In total, `r diatom1_summ$n_low_n %>% as.english()` assemblages appear to have a count sum below 400 valves.
+
+```{r line178}
+
+```
+
+The archived data include `r diatom1_summ$ntaxa` taxa that have a maximum abundance of at least 1.25%; this is about half of the 109 taxa reported in the paper.
+This pruning of rare taxa will increase the risk of counts without singletons, but does not appear to have caused any problem here (which would be reflected by non-integer multiples).
+
+Dataset diatom2 has `r diatom2_summ$nsamples` assemblages with a median richness of `r diatom2_summ$median_taxa` taxa.
+The associated paper reports that the diatom counts included 400–500 valves per assemblage except for four diatom-poor assemblages with at least 100 valves.
+The direct search method estimates that `r diatom2_summ$n400 %>% as.english()` of the assemblages have a count sum of less than 400, and `r diatom2_summ$n100 %>% as.english()` have less than 100.
+The direct search and minimum percent methods agree, within rounding error, `r diatom2_summ$meth_agree %>% insist(isTRUE(.)) %>% invisible()`on the estimated count sums, implying that either all assemblages have singletons, or that the greatest common divisor of the raw counts is greater than one.
+This means that for the assemblage with the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc %>% as.english()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.
+
+Dataset chironomid1 has `r chironomid1_summ$nsamples` assemblages.
+Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 %>% as.english` assemblages had count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
+These assemblages are diverse (richness between `r chironomid1_summ$richness %>% paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.
+
+Palaeoceanographic dataset marine1 has `r marine1_summ$nsamples` assemblages with a median taxonomic richness of `r marine1_summ$median_taxa` taxa.
+The associated paper reports that assemblages with count sums below 100 were omitted, and that there were 20 assemblages with count sums below 200.
+The direct search method estimates that `r marine1_summ$n100 %>% as.english()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa %>% as.english()` mono-specific assemblage for which no estimate of the count sum is possible), with count sums as low as `r marine1_summ$est_n_min`.
+With this dataset, the direct search and the minimum percent method give identical results for some assemblages but highly divergent estimates for others, with the direct search method sometimes giving estimates of several thousand, in one case over two orders of magnitude higher than the minimum percent method.
+Some of the divergence appears to be because some, mainly low diversity, assemblages lack singletons.
+The most extreme divergences appear to be data entry errors with incorrect rounding, perhaps during taxonomic revisions of the percent data rather than the raw counts.
 
@@
 
-Chao index
-relationship to GRIM
+My search through archived microfossil count data found several datasets where count sums are, or appear to be, smaller than that reported.
+The datasets examined are a convenience sample of available count data that met the inclusion criteria.
+As such, this analysis cannot be used to accurately determine the prevalence of datasets with possible undercounts, but it appears that a non-negligible fraction of the literature is affected.
+It is possible in some cases that the assemblages with low counts were omitted or merged prior to analysis, but the number of assemblages reported was not updated.
+Care was taken to identify any such cases by, for example, examining stratigraphic diagrams to see if the low count assemblages were included.
+In some cases, the assemblages with low apparent counts can be seen in the published stratigraphic diagrams.
 
-Caveats
+It is also possible that the apparently low count sums are a false positive. 
+This can only happen in assemblages that have a greatest common divisor greater than one.
+The analysis of the breeding bird, testate amoeba and Owen's Lake datasets shows that assemblages without singletons are rare, and assemblages with a greatest common divisor greater than one are very rare, especially for taxonomically diverse assemblages.
+If a dataset contains several taxonomically diverse assemblages that have a greatest common divisor of two (or expecially if greater than two) doubts should be raised about whether the counts sums are accurately described.
 
-decimal places -- precision
+Count sums may be smaller than expected if rare taxa have been omitted from the archived data.
+This should be a minor problem.
 
-Unused/unmerged samples - care taken here
+Microfossil percent data are often given to two decimal places. 
+This is sufficient precision for the direct search method to have utility with counts sums of several thousand. 
+Some data are only given to the nearest percent, which means that neither method has utility with count sums above 100.
 
-different count sums (eg trees & herbs vs trees herbs and pteridiophytes)
-
-Low diversity 
-Truncated diversity
-
-Large counts
+The direct search method will fail if the dataset includes taxa calculated with different count sums, for example pollen sums of trees, shrubs and  upland herbs and a pollen and spore sum that also includes pteridiophytes.
+It should be possible to identify such cases from the meta-data and knowledge of usual practice with different proxies. 
+It will also fail, as show by dataset marine1, if percentages are incorrectly calculated or rounded.
 
 Some taxonomic groups have microfossils that often come in groups of attached individuals.
-For example, for diatoms valves are counted, but these are commonly paired or, for some taxa, in long chains.
+For example, for each diatom cell has two valves, the counting unit.
+These valves usually separate during processing, but paired valves are often found, and some taxa such as _Aulacoseira_ produce long chains of strongly bound valves.
+This might make singletons slighly less likely. 
+The results from Owen's Lake suggest that this is not an important problem, as most assemblages have many singletons.
 
-Assemblages vs communities
+Small undercounts in a few assemblages will have minimal impact on the precision of any palaeoenvironmental reconstruction or other statistics derived from the assemblage.
+Substantial undercounts will potentially seriously effect the precision of the results.
+Such undercounts might constitute a data handling error, for example, if samples with low counts were supposed to be merged but that step was forgotten.
+Substantial and pervasive undercounting could be construed as scientific misconduct due to either negligence or falsification, and action to correct the literature is probably required.
 
-Consequences
-
-Small undercounts in a few samples will have minimal impact on the precision of any palaeoenvironmental reconstruction or other statistics derived from the assemblage. 
-Consequently, such undercounts can be regarded as bad practice that do not necessarily require correcting.
-
-Substantial or pervasive undercounts will potentially seriously effect the precision of the results. 
-Such undercounts might either constitute a data handling error (for example, if samples with low counts were supposed to be merged but that step was forgotten, or using boilerplate method text), or falsification.
-Action to correct the literature is probably required.
-
-Some papers don't report count sums. 
+Some papers do not report count sums.
 When this important quality metric is omitted, the reader should be able to assume that the standard minimum count sum for the taxonomic group has been used (i.e.
 50 for chironomids, several hundred for pollen and diatoms).
-If the actual count sum is materially below this, then this potentially constitutes falsification by omission [@Fanelli2013].
+If the actual count sums are materially below this, then this potentially constitutes falsification by omission [@Fanelli2013].
+
+These results highlight the importance of archiving both the raw data and the code required to process the data.
 
@@
 
-non-negligible fraction of the literature
+A non-negligible fraction of the archived assemblage data includes counts that appear to have a count sum below that reported in the paper. 
+Some of the apparent counts are small enough that the uncertainty of the count and derived statistics will be substantially larger than expected.
 
-## Acknowledgements{-}
+# Acknowledgements{-}
+
+I thank ... Neotoma?? Pangaea??
 
@@
 
-## References{-}
+# References{-}

### 2019-04-27 R/download_birds.R

@@ -23,19 +23,32 @@ bird_plan <- drake_plan(
     rename(Latin_name = Spanish_Common_Name),
   
   #extract bird data
-  bird_data = list.files(path = "data/birds/", pattern = "\\.zip$", full.names = TRUE) %>% 
+  bird_data = {
+    bird_download # force dependency
+    list.files(path = "data/birds/", pattern = "\\.zip$", full.names = TRUE) %>% 
     map(read_csv) %>% 
     map(mutate, statenum  = as.numeric(statenum)) %>% 
     map(~mutate(., count = rowSums(select(., starts_with("Stop"))))) %>% 
-    map_df(select, -starts_with("Stop")),
+    map_df(select, -starts_with("Stop"))
+    },
   
   #aggregate to different taxonomic levels
   bird_singletons = bird_process(bird_data, bird_species),
 
-  bird_singleton_summary = bird_singletons %>% 
-    group_by(taxonomic_level) %>%   
-    summarise(pc_singletons = mean(singletons == 0) * 100, n_singletons = sum(singletons == 0), mean_richness = mean(n_taxa), n_routes = n(), GCD = mean(GCD > 1, na.rm = TRUE) * 100) %>% 
-    mutate(taxonomic_level = factor(taxonomic_level, levels = c("species", "genus", "family", "order"))) %>% 
+  bird_singleton_summary = bird_singletons %>%
+    group_by(taxonomic_level) %>%
+    summarise(
+      pc_singletons = mean(singletons == 0) * 100,
+      n_singletons = sum(singletons == 0),
+      mean_richness = mean(n_taxa),
+      n_routes = n(),
+      GCD2plus = mean(GCD > 1, na.rm = TRUE) * 100,
+      GCD_max = max(GCD, na.rm = TRUE)
+    ) %>%
+    mutate(taxonomic_level = factor(
+      taxonomic_level,
+      levels = c("species", "genus", "family", "order")
+    )) %>%
     arrange(taxonomic_level),
   
   #bird order GCD summary

### 2019-04-27 R/download_chironomids.R

@@ -2,19 +2,70 @@
 
 chironomid_plan <- drake_plan(
   #data for Last Chance Lake
-  last_chance0 = read.table("https://www1.ncdc.noaa.gov/pub/data/paleo/insecta/chironomidae/northamerica/greenland/last-chance2017chironomid.txt", header = TRUE),
+  last_chance = read.table("https://www1.ncdc.noaa.gov/pub/data/paleo/insecta/chironomidae/northamerica/greenland/last-chance2017chironomid.txt", header = TRUE),
   
-  last_chance_meta = last_chance0 %>% select(age_calBP, totcaps),
-  last_chance = last_chance0 %>% select(-age_calBP, -totcaps),
+  last_chance_n = last_chance %>% 
+    gather(key = taxon, value = percent, -age_calBP, -totcaps) %>% 
+    filter(percent > 0) %>% 
+    estimate_n(ID_cols = c("age_calBP", "totcaps"), digits = 2),
   
-  last_chance_n = estimate_n(spp = last_chance, digits = 2) %>% 
-    bind_cols(last_chance_meta),
-  
-  last_chance_plot = last_chance_n %>%  
-    ggplot(aes(x = totcaps, y = est_n, ymax = est_max, ymin = est_min)) + 
+  last_chance_direct_search_plot = last_chance_n %>% 
+    filter(age_calBP == 1099) %>% 
+    unnest(direct_search) %>% 
+    ggplot(aes(x  = n, y = score)) + 
     geom_point() +
-    geom_errorbar() +
-    geom_abline(aes(slope = sl, intercept = 0, colour = sf, linetype = sf), data = tibble(sl = 1:2, sf = factor(sl)), show.legend = FALSE) +
+    labs(x = "Putative count sum", y = "Score") + 
+    xlim(NA, 300),
+  
+  last_chance_plot = last_chance_n %>% 
+    unnest(direct_search_est) %>%   
+    ggplot(aes(x = totcaps, y = est_n_minpc)) + 
+    geom_abline(aes(slope = sl, intercept = 0, linetype = sf), data = tibble(sl = 1:2, sf = factor(sl)), show.legend = FALSE, colour = "grey50") +
+    geom_point() +
+    geom_errorbar(aes(ymax = est_max_minpc, ymin = est_min_minpc)) +
+    geom_point(mapping = aes(x = totcaps, y = est_n_direct), shape = "x", colour = "red") +
     coord_equal() +
[…]

### 2019-04-27 R/download_diatoms.R

@@ -1,0 +1,149 @@
+#diatoms
+diatom_plan <- drake_plan(
+  ####Owens Lake####
+  owens = read_file(file = "http://diatom.ansp.org/dpdc/download.asp?type=count&DatasubsetID=51") %>% 
+    gsub(pattern = "<.*?>", replacement = "", x = .) %>% 
+    read_delim(delim = "\t"),
+  
+  owens_summ = owens %>% 
+  group_by(SampleId) %>% 
+  summarise(mn = min(Count), n_sing = sum(Count == 1), sum = sum(Count), n_taxa = n(), gcd = ifelse(n_taxa > 1, numbers::mGCD(Count), NA)),
+  
+  owens_summary2 = owens_summ %>% 
+    filter(n_taxa > 1) %>% 
+    summarise(
+      n = n(),
+      median_sing = median(n_sing),
+      no_singletons = mean(mn != 1) * 100,
+      no_single_lt_50 = mean(mn[sum < 50] != 1) * 100, 
+      no_single50_ = mean(mn[sum >= 50] != 1) * 100, 
+      gcd2 = mean(gcd[mn > 1] == 1) * 100,
+      gcd_max = max(gcd, na.rm = TRUE)),
+  
+  
+  ####diatom1####
+  diatom1_data = {target <- secrets %>% 
+    filter(dataset == "diatom1") %>% 
+    pull(secret)
+    read_delim(target, delim = "\t", comment = "#")
+  },
+  
+  diatom1_estn = diatom1_data %>% 
+    gather(key = taxon, value = percent, -(depth_cm:age_calBP)) %>%
+    filter(percent > 0) %>% 
+    estimate_n(ID_cols = c("depth_cm", "age_calBP"), digits = 2) %>% 
+    assertr::verify(map_int(direct_search_est, nrow) == 1) %>% 
+    unnest(direct_search_est) %>% #check only one row in each direct_search_est
+    assertr::verify(score == 1),
+  
+  diatom1_pcheck = diatom1_data %>% 
+    select(-age_calBP) %>%
+    percent_checker( site_column = "depth_cm", digits = 2) %>% 
+    select(-one_max,-one_min, -est_min, -est_max) %>% 
+    count(contains) %>% 
+    mutate(prop = n/sum(n)),
+  
+  diatom1_summ = list(
+    nsamples = nrow(diatom1_data),
[…]

### 2019-04-27 R/download_marine.R

@@ -1,0 +1,61 @@
+marine_plan <- drake_plan(
+
+  marine1 = {
+    target <- secrets %>%
+      filter(dataset == "marine1") %>%
+      pull(secret)
+    read_delim(target, delim = "\t", skip = 324)
+  },
+
+  marine1_est_n = marine1 %>% 
+    select(-(Event:`Si [µmol/l] (summer 10 m)`)) %>% 
+    rowid_to_column("sampleID") %>%  
+    gather(key = taxon, value = percent, -sampleID) %>% 
+    filter(percent > 0) %>% 
+    estimate_n(ID_cols = "sampleID", digits = 2, nmax = 10000) %>% 
+    mutate(direct_search_est = map(direct_search_est, slice, 1)) %>% #take only first estimate if multiple  - only affects v large estimates
+    unnest(direct_search_est) %>%
+    assertr::verify(score == 1),
+  
+  marine1_summ = list(
+    nsamples = nrow(marine1),
+    median_taxa = median(marine1_est_n$n_taxa),
+    n_one_taxa = sum(marine1_est_n$n_taxa == 1),
+    n100 = sum(between(marine1_est_n$est_n_direct, 2, 100)),
+    est_n_min = marine1_est_n %>% 
+      filter(n_taxa > 1) %>% 
+      pull(est_n_direct) %>% 
+      min(na.rm = TRUE)
+    )
+  
+)
+
+
+# x %>% unnest(direct_search_est) %>% group_by(sampleID) %>% slice(1) %>% ungroup() %>% select(c(1:3, 6), score, est_n_direct, n_taxa) %>% filter(est_n_minpc < 100)
+# 
+# marine1 %>% 
+#   select(-(Event:`Si [µmol/l] (summer 10 m)`)) %>% 
+#   rowid_to_column("sampleID") %>%  
+#   gather(key = taxon, value = percent, -sampleID) %>% 
+#   filter(percent > 0) %>%
+#   group_by(sampleID) %>% 
+#   mutate(minpc = min(percent)) %>% 
+#   filter(percent >= 5) %>% 
+#   group_by(sampleID, percent) %>% 
+#   mutate(n = n() - 1) %>% 
+#   group_by(sampleID) %>% 
+#   summarise(mn = first(minpc), s = sum(n))
+#   
+# 
+# marine1 %>% 
+#   select(-(Event:`Si [µmol/l] (summer 10 m)`)) %>% 
[…]

### 2019-04-27 R/download_testates.R

@@ -32,9 +32,19 @@ testate_plan <- drake_plan(
       cnt[, names(cnt) %in% tl$taxon.name]
       })) %>%
     map_df(gather, key = taxon, value = count, .id = "sampleID") %>% 
-    mutate(sampleID = as.numeric(sampleID))
-
+    mutate(sampleID = as.numeric(sampleID)),
   
+  #run analyses
+  testate_summary = summarise_counts(testate_counts),
+
+  testate_summ = list(
+    nsamples = n_distinct(testate_counts$sampleID),
+    n_gcd2 = sum(testate_summary$gcd2plus$gcd == 2),
+    n_gcd3 = sum(testate_summary$gcd2plus$gcd == 3),
+    n_gcd4 = sum(testate_summary$gcd2plus$gcd > 3) %>% insist(. == 0),
+    n_gcd2_low_div = mean(testate_summary$gcd2plus$gcd < 5) * 100,    
+    n_gcd2_low_div = mean(testate_summary$gcd2plus$gcd < 5) * 100
+  )
 )
 
 # testate_config <- drake_config(testate_plan)

### 2019-04-27 R/summarise_counts.R

@@ -5,7 +5,7 @@ summarise_counts <- function(counts){#columns sampleID/taxon/count
   
   results$count_summary <- counts %>% 
     group_by(sampleID) %>% 
-    summarise(min = min(count), n = n(), sum = sum(count), singletons = sum(count == 1))
+    summarise(min = min(count), n_taxa = n(), sum = sum(count), singletons = sum(count == 1))
   
   results$prop_singletons <- results$count_summary %>% 
     summarise(m = mean(min == 1)) %>% 
@@ -23,8 +23,10 @@ summarise_counts <- function(counts){#columns sampleID/taxon/count
     group_by(sampleID) %>%
     mutate(m = min(count)) %>%
     filter(m > 1) %>% #remove samples with singletons
-    summarise(gcd = numbers::mGCD(count)) #find greatest common divisor
+    summarise(gcd = numbers::mGCD(count), n_taxa = n()) #find greatest common divisor
     
+  results$gcd2plus <- results$no_singletons %>%
+    filter(gcd > 1)
   
   results$gcd1 <- results$no_singletons %>% 
     summarise(prop1 = mean(gcd == 1)) %>% 
@@ -33,19 +35,3 @@ summarise_counts <- function(counts){#columns sampleID/taxon/count
   return(results)
   
 }
-
-# 
-# 
-# loadd(testate_meta)
-# 
-# testate_meta %>% 
-#   filter(!readd(testate_has_percent)) %>% 
-#   left_join(
-#     testate_counts %>% 
-#       mutate(sampleID = as.numeric(sampleID)), 
-#     by =c("dataset.id" = "sampleID")) %>% 
-#   group_by(dataset.id) %>% 
-#   filter(min(count) > 1) %>% 
-#   slice(1)
-#                                                                                                                                                                                                     
-# 

### 2019-04-27 make_count_check_ms_with_drake.R

@@ -2,15 +2,17 @@
 library("drake")
 library("tidyverse")
 library("readr")
-library("gridExtra")
-library("broom")
+#library("gridExtra")
+#library("broom")
 #devtools::install_github("richardjtelford/rjt.misc")
 library("rjt.misc")
+#devtools::install_github("richardjtelford/count_checker")#need to make package and repo names identical (_ not permitted)
 library("countChecker")
-library("english")
 
 #drake configuration
 pkgconfig::set_config("drake::strings_in_dots" = "literals")
+#set up parallel processing for drake
+future::plan(future::multiprocess) 
 
 ## askpass
 options(askpass = function(x)readLines("pw.txt"))
@@ -20,6 +22,8 @@ source("R/download_birds.R")
 source("R/download_testates.R")
 source("R/download_chironomids.R")
 source("R/download_pollen.R")
+source("R/download_diatoms.R")
+source("R/download_marine.R")
 source("R/summarise_counts.R")
 
 
@@ -27,17 +31,14 @@ source("R/summarise_counts.R")
 analyses <- drake_plan(
   
   #secrets
-  secrets = readRDS("data/secrets.RDS") %>%
+  secrets = readRDS(file_in("data/secrets.RDS")) %>%
       encryptr::decrypt(secret),
-
-  #run analyses
-  testate_summary = summarise_counts(testate_counts),
   
   #make plots
   
   #add extra packages to bibliography
   biblio2 = package_citations(
-    packages = c("extraDistr"), 
+    packages = c("extraDistr", "countChecker", "numbers"), 
     old_bib = file_in("Rmd/extra/countMS.bib"), 
     new_bib = file_out("Rmd/extra/countMS2.bib")),
   
@@ -53,18 +54,25 @@ analyses <- drake_plan(
 )
 
 #put plans together
-plans <- bind_rows(bird_plan, testate_plan, chironomid_plan, analyses, pollen_plan)
+plans <- bind_rows(
+  bird_plan, 
+  testate_plan, 
+  chironomid_plan, 
+  pollen_plan,
+  diatom_plan,
+  marine_plan,
+  analyses)
 
 #configure and make drake plan
 config <- drake_config(plans)
 
-#set up parallel processing for drake
-future::plan(future::multiprocess) 
-
 #Build the right things
-make(plans, jobs = 2, parallelism = "future")
+make(plans, jobs = 2, parallelism = "future", keep_going = TRUE)
[…]

### 2019-04-28 Rmd/count_check_MS.Rmd
Sunday edits Owens %

@@
 ---
-title: "Tools for identifying unexpectely low microfossil counts"
+title: "Tools for identifying unexpectely low microfossil count sums"
 author:
@@
 loadd(last_chance_summ, pollen1_summary)
-loadd(owens_summ, owens_summary2)
+loadd(owens_summ, owens_summary2, owens_est_summ)
 loadd(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ)
@@
 In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual [@Coddington2009], hereafter singletons, unless the sampling effort is high relative to the taxonomic richness.
-In the occasional assemblages without singletons, the greatest common divisor should usually be one, especially if species richness is high.
+In the occasional assemblages without singletons, the greatest common divisor should usually be one, i.e.
+few assemblages should have all counts divisable by an integer larger than one, especially if species richness is high.
 
-In general, it is not possible to estimate the sum from which percent were calculated, but the properties of community and assemblage counts make it possible.
+In general, it is not possible to determine the sum from which percent were calculated, but the properties of community and assemblage counts make it possible to estimate it.
 Given that we expect the rarest taxon to be a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
@@
 An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
-These tests can be viewed as special cases of GRIM [@Brown2017].
+These tests are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.
 
@@
 Datasets where the minimum count sum was not reported in the associated paper were excluded.
-Datasets from Lakes Zabinskie, Silvaplana and Seebergsee discussed by Telford [-@Telford2019] as possibly having under-reported count sums were also excluded to avoid double reporting.
+Datasets discussed by Telford [-@Telford2019] as possibly having under-reported count sums were also excluded to avoid double reporting.
 Datasets with possible mis-reporting are anonymised, but no attempt is made to diagnose whether errors or misconduct are responsible.
@@
 
-The a tiny minority of the over `r format(floor(bird_singleton_summary$n_routes[1]/1000) * 1000, big.mark = ",")` bird counts from the North American breeding bird survey [@NAmerican_Breeding_Bird2018]
+A tiny minority of the over `r format(floor(bird_singleton_summary$n_routes[1]/1000) * 1000, big.mark = ",")` bird counts from the North American breeding bird survey [@NAmerican_Breeding_Bird2018]
 lack singletons at the species level (Table \@ref(tab:bird-table))).
@@
 ```{r bird-recount, fig.cap="Effect of count sum on the proportion of bird counts without singleton orders.
-Results are the mean of ten trials, error bars are two standard errors.", fig.width = 90/25.4, fig.height = 90/25.4}
+Results are the mean of ten trials, error bars are two standard errors.", fig.width = 70/25.4, fig.height = 70/25.4}
 readd(bird_recount_singleton_plot)
@@
 Discounting the `r sum(owens_summ$n_taxa == 1)` mono-specific assemblages (all with very low count sums`r max(owens_summ$sum[owens_summ$n_taxa == 1]) %>% insist(.
-<= 10) %>% invisible()`), the median number of singletons per count is `r owens_summary2$median_sing %>% as.english()` and only `r owens_summary2$no_singletons`% of counts lack singletons.
-Assemblages with counts below 50 have a higher chance of lacking singletons (`r owens_summary2$no_single_lt_50`% vs `r owens_summary2$no_single50_`%).
+<= 10) %>% invisible()`), the median number of singletons per count is `r owens_summary2$median_sing %>% as.english()` and only `r owens_summary2$no_singletons`% of assemblages lack singletons.
+Assemblages with count sums below 50 have a higher chance of lacking singletons (`r owens_summary2$no_single_lt_50`% vs `r owens_summary2$no_single50_`%).
 Of the assemblages without singletons, `r owens_summary2$gcd2`% have a greatest common divisor of one.
@@
 
-The `r testate_summ$nsamples` assemblages in the North American testate amoeba training set with available count data [@AMESBURY2018] have count sums between `r testate_summary$count_summary$sum %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$sum %>% median()`) tests, and taxonomic richness ranges between `r testate_summary$count_summary$n_taxa %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$n_taxa %>% median()`) taxa.
+The `r testate_summ$nsamples` assemblages in the North American testate amoeba training set with available count data [@AMESBURY2018] have count sums between `r testate_summary$count_summary$sum %>% range() %>% paste(collapse = " and ")` tests (median = `r testate_summary$count_summary$sum %>% median()`), and taxonomic richness ranges between `r testate_summary$count_summary$n_taxa %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$n_taxa %>% median()`) taxa.
 `r testate_summary$prop_singletons * 100`% of the assemblages have at least one singleton (median `r testate_summary$median_singletons`).
 In the assemblages that lack singletons, `r testate_summary$gcd1 * 100`% have a greatest common divisor of one.
-Of these assemblages, `r testate_summ$n_gcd2_low_div`% have fewer than five taxa; `r testate_summ$n_gcd2 %>% as.english()` have a greatest common divisor of two, and the other `r testate_summ$n_gcd3 %>% as.english()` has a greatest common divisor of three.
+Of the `r testate_summ$n_gcd2plus %>% as.english()` assemblages with a greatest common divisor above one, `r testate_summ$n_gcd2_low_div %>% as.english()` have fewer than five taxa; `r testate_summ$n_gcd2 %>% as.english()` have a greatest common divisor of two, and the other `r testate_summ$n_gcd3 %>% as.english()` has a greatest common divisor of three.
 
@@
 
-The `r last_chance_summ$nsamples` chironomid head capsule assemblage dataset from Last Chance Lake [@Axford2017] includes both the count sums and the percentage of each taxa, allowing the count sums estimated from the rarest taxon to be checked.
+The Owen's Lake dataset [@Platt1997] also includes percentages for each taxa, allowing the count sums estimated from the percent data to be verified.
+Excluding monospecific assemblages, for which no meaningful estimate of the count sum is possible, the minimum percent method and the direct search method correctly estimate the count sum for, respectively, `r owens_est_summ$minpc`% and `r owens_est_summ$direct`% of the assemblages.
+The few assemblages that fail the direct seach method are species poor and have low count sums.
+
+The `r last_chance_summ$nsamples` chironomid head capsule assemblage dataset from Last Chance Lake [@Axford2017] also includes both the count sums and the percentage of each taxa.
 The percentages are given to two decimal places: to account for rounding errors, the countChecker package uses the smallest and largest values consistent with the reported percentage.
@@
 Reporting half microfossils is common for several microfossil groups, including chironomids and pollen (especially for bisaccate conifers); occasionally other fractions are reported.
-Fractional counts make the estimated count too high by a factor of two, so do not risk incorrectly flagging counts as being too small.
+Half counts make the estimated count too high by a factor of two, so do not risk incorrectly flagging count sums as being too small.
 The direct search method is more precise than the minimum percent method as the rounding error is relatively smaller on the larger percent values the method uses.
@@
 The archived data have counts sums between `r pollen1_summary$count$min_count` and `r pollen1_summary$count$max2_count`, with the exception of one count of `r pollen1_summary$count$max_count` (which may be so high because of a possible data entry error).
-However, these count sums include trilete spores, probably from a fern or fern ally, with an abundance as high as `r pollen1_summary$max_trilete`.
-Unexpectedly, only `r pollen1_summary$count$has_singletons %>% insist(. == 1) %>% as.english()` of the counts has a singleton.
+Count sums that are smaller than 200 may be because some rare taxa have been omitted from the archived data.
+This should normally only cause small discrepancies.
+More importantly, these count sums include trilete spores, probably from a fern or fern ally, with an abundance as high as `r pollen1_summary$max_trilete` spores.
+Unexpectedly, only `r pollen1_summary$count$has_singletons %>% insist(. == 1) %>% as.english()` of the counts has a singleton, a _`r pollen1_summary$individual$singleton_spp`_ grain.
 The other counts have minimum abundances of two or four pollen grains.
@@
 
-The archived data include `r diatom1_summ$ntaxa` taxa that have a maximum abundance of at least 1.25%; this is about half of the 109 taxa reported in the paper.
-This pruning of rare taxa will increase the risk of counts without singletons, but does not appear to have caused any problem here (which would be reflected by non-integer multiples).
+The archived data include `r diatom1_summ$n_taxa` taxa that have a maximum abundance of at least 1.25%; this is about half of the 109 taxa reported in the paper.
+This pruning of rare taxa will increase the risk of counts without singletons, which would cause the minimum percent method to fail, but will have little impact on the direct search method.
 
@@
 The associated paper reports that the diatom counts included 400–500 valves per assemblage except for four diatom-poor assemblages with at least 100 valves.
-The direct search method estimates that `r diatom2_summ$n400 %>% as.english()` of the assemblages have a count sum of less than 400, and `r diatom2_summ$n100 %>% as.english()` have less than 100.
-The direct search and minimum percent methods agree, within rounding error, `r diatom2_summ$meth_agree %>% insist(isTRUE(.)) %>% invisible()`on the estimated count sums, implying that either all assemblages have singletons, or that the greatest common divisor of the raw counts is greater than one.
+The direct search method estimates that `r diatom2_summ$n400 %>% as.english()` assemblages have a count sum of less than 400, and `r diatom2_summ$n100 %>% as.english()` have less than 100.
+The direct search and minimum percent methods agree, within rounding error, `r diatom2_summ$meth_agree %>% insist(isTRUE(.)) %>% invisible()`on the estimated count sums, implying that either all assemblages have singletons and count sums are low in some assemblages, or that the greatest common divisor of the raw counts is greater than one.
 This means that for the assemblage with the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc %>% as.english()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.
@@
 The associated paper reports that assemblages with count sums below 100 were omitted, and that there were 20 assemblages with count sums below 200.
-The direct search method estimates that `r marine1_summ$n100 %>% as.english()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa %>% as.english()` mono-specific assemblage for which no estimate of the count sum is possible), with count sums as low as `r marine1_summ$est_n_min`.
+The direct search method estimates that `r marine1_summ$n100 %>% as.english()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa %>% as.english()` mono-specific assemblage), with count sums as low as `r marine1_summ$est_n_min`.
 With this dataset, the direct search and the minimum percent method give identical results for some assemblages but highly divergent estimates for others, with the direct search method sometimes giving estimates of several thousand, in one case over two orders of magnitude higher than the minimum percent method.
@@
 My search through archived microfossil count data found several datasets where count sums are, or appear to be, smaller than that reported.
+Some of the count sums appear to be an order of magnitude below what was reported.
 The datasets examined are a convenience sample of available count data that met the inclusion criteria.
@@
 
-Count sums may be smaller than expected if rare taxa have been omitted from the archived data.
-This should be a minor problem.
-
 Microfossil percent data are often given to two decimal places. 
@@
 Some taxonomic groups have microfossils that often come in groups of attached individuals.
-For example, for each diatom cell has two valves, the counting unit.
+For example, for each diatom cell has two valves which are the counting unit.
 These valves usually separate during processing, but paired valves are often found, and some taxa such as _Aulacoseira_ produce long chains of strongly bound valves.
@@
 
-These results highlight the importance of archiving both the raw data and the code required to process the data.
-
 # Conclusions
@@
 Some of the apparent counts are small enough that the uncertainty of the count and derived statistics will be substantially larger than expected.
+These results highlight the importance of archiving both the raw data and the code required to process the data.
 
@@
 
-I thank ... Neotoma?? Pangaea??
+Some data were obtained from the Neotoma Paleoecology Database (http://www.neotomadb.org), and the work of the data contributors and the Neotoma community is gratefully acknowledged.
 

### 2019-04-28 R/download_diatoms.R

@@ -20,6 +20,36 @@ diatom_plan <- drake_plan(
       gcd2 = mean(gcd[mn > 1] == 1) * 100,
       gcd_max = max(gcd, na.rm = TRUE)),
   
+  ####owens diatoms####
+  owens_est = owens %>% 
+    estimate_n(ID_cols = c("SampleId", "SampleCode", "CountSum"), percent_col = "Percent", taxon_col = "TaxonShortName", digits = 4) %>%
+    assertr::verify(map_int(direct_search_est, nrow) == 1) %>% #check only one row in each direct_search_est
+    unnest(direct_search_est) %>% 
+    assertr::verify(score == 1),#check score ==1
+  
+  owens_est_summ = owens_est %>% 
+    select(est_n_minpc, est_n_direct, CountSum, n_taxa) %>% 
+    filter(n_taxa > 1) %>% 
+    summarise(
+      minpc = mean(round(est_n_minpc, 3) == CountSum) * 100, 
+      direct = mean(est_n_direct == CountSum) * 100
+    ),
+  
+  owens_fail = owens_est %>% 
+    select(est_n_minpc, est_n_direct, CountSum, n_taxa) %>%
+    filter(n_taxa >1, est_n_direct!=CountSum),
+  
+  owens_plot = owens_est %>% 
+    filter(between(n_taxa, 8, 15), between(CountSum, 80, 150)) %>% 
+    slice(4) %>%  
+    select(c(SampleId:n_taxa), score, est_n_direct) %>% 
+    unnest(direct_search) %>% 
+    gather(key = taxon, value = presence, -c(SampleId:est_n_direct, n, score1)) %>% 
+    ggplot(aes(x = taxon, y = n, fill = presence)) + 
+    geom_raster(show.legend = FALSE) +
+    scale_y_continuous(limits = c(0, 300), expand = c(0, 0)) +
+    labs(y = "Putative count sum") +
+    scale_fill_manual(values = c("white", "black")),
   
   ####diatom1####
   diatom1_data = {target <- secrets %>% 
@@ -80,12 +110,6 @@ diatom_plan <- drake_plan(
     assertr::verify(map_int(direct_search_est, nrow) == 1) %>% #check only one row in each direct_search_est, 
[…]

### 2019-04-28 R/download_pollen.R

@@ -24,7 +24,7 @@ pollen_plan <- drake_plan(
    
    individual = pollen1_counts %>% 
      ungroup() %>%
-     summarise(one = sum(count == 1), two = sum(count %% 2 == 0), n = n(), not_two = n - two),
+     summarise(one = sum(count == 1), two = sum(count %% 2 == 0), n = n(), not_two = n - two, singleton_spp = taxa[count == 1]),
    
    max_trilete =  pollen1_counts %>% 
      filter(taxa == "Unknown (trilete)") %>% 

### 2019-04-28 R/download_testates.R

@@ -39,11 +39,11 @@ testate_plan <- drake_plan(
 
   testate_summ = list(
     nsamples = n_distinct(testate_counts$sampleID),
+    n_gcd2plus = nrow(testate_summary$gcd2plus),
     n_gcd2 = sum(testate_summary$gcd2plus$gcd == 2),
     n_gcd3 = sum(testate_summary$gcd2plus$gcd == 3),
     n_gcd4 = sum(testate_summary$gcd2plus$gcd > 3) %>% insist(. == 0),
-    n_gcd2_low_div = mean(testate_summary$gcd2plus$gcd < 5) * 100,    
-    n_gcd2_low_div = mean(testate_summary$gcd2plus$gcd < 5) * 100
+    n_gcd2_low_div = sum(testate_summary$gcd2plus$n_taxa < 5) 
   )
 )
 

### 2019-04-28 make_count_check_ms_with_drake.R

@@ -66,6 +66,9 @@ plans <- bind_rows(
 #configure and make drake plan
 config <- drake_config(plans)
 
+#view dependency graph
+vis_drake_graph(config, targets_only = TRUE)
+
 #Build the right things
 make(plans, jobs = 2, parallelism = "future", keep_going = TRUE)
 failed()

### 2019-05-02 Rmd/count_check_MS.Rmd
countChecker -> countSum package

@@
 
-All analyses were done in R version `r getRversion()` [@R] and used the packages extraDistr version `r packageVersion("extraDistr")` [@extraDistr], numbers version `r packageVersion("numbers")` [@numbers], and countChecker `r packageVersion("countChecker")` [@countChecker].
+All analyses were done in R version `r getRversion()` [@R] and used the packages extraDistr version `r packageVersion("extraDistr")` [@extraDistr], numbers version `r packageVersion("numbers")` [@numbers], and countSum `r packageVersion("countSum")` [@countSum].
 Code to replicate all the analyses shown above is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms).
@@
 The `r last_chance_summ$nsamples` chironomid head capsule assemblage dataset from Last Chance Lake [@Axford2017] also includes both the count sums and the percentage of each taxa.
-The percentages are given to two decimal places: to account for rounding errors, the countChecker package uses the smallest and largest values consistent with the reported percentage.
+The percentages are given to two decimal places: to account for rounding errors, the countSum package uses the smallest and largest values consistent with the reported percentage.
 With the minimum percent method the estimated count sums are, within error, either identical to the reported count sum or twice as much (Fig. \@ref(fig:last-chance-plot)).

### 2019-05-02 make_count_check_ms_with_drake.R

@@ -7,7 +7,7 @@ library("readr")
 #devtools::install_github("richardjtelford/rjt.misc")
 library("rjt.misc")
 #devtools::install_github("richardjtelford/count_checker")#need to make package and repo names identical (_ not permitted)
-library("countChecker")
+library("countSum")
 
 #drake configuration
 pkgconfig::set_config("drake::strings_in_dots" = "literals")
@@ -38,7 +38,7 @@ analyses <- drake_plan(
   
   #add extra packages to bibliography
   biblio2 = package_citations(
-    packages = c("extraDistr", "countChecker", "numbers"), 
+    packages = c("extraDistr", "countSum", "numbers"), 
     old_bib = file_in("Rmd/extra/countMS.bib"), 
     new_bib = file_out("Rmd/extra/countMS2.bib")),
   

### 2019-05-07 Rmd/count_check_MS.Rmd
on the train to oslo

@@
 Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so.
-This paper introduces methods that can flag assemblages with potentially misreported count sums and finds that some assemblage datasets appear to fail these tests.
+This paper introduces methods that can flag assemblages with potentially misreported count sums and finds that some assemblage datasets that fail these tests.
 journal: "Quaternary Science Reviews"
@@
 loadd(bird_singleton_summary)
-loadd(testate_summ, testate_summary)
-loadd(last_chance_summ, pollen1_summary)
-loadd(owens_summ, owens_summary2, owens_est_summ)
+loadd(testate_summ)
+loadd(last_chance_summ)
+loadd(pollen_summ, pollenA, pollen1234)
+loadd(owens_summ1, owens_summ, owens_est_summ)
 loadd(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ)
@@
 Carlisle [-@Carlisle2017] identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data.
-Brown et al. [-@Brown2017] developed the granularity-related inconsistency means test (GRIM) of whether means of integer data are consistent with the reported sample size.
+Brown et al. [-@Brown2017] developed the granularity-related inconsistency means (GRIM) test of whether means of integer data are consistent with the reported sample size.
 However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data.
@@
 In addition, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there may be an incentive to misreport the minimum count sum.
-The risk that count sums will be mis-reported is not just theoretical: Larocque-Tobler et al [-@Larocque-Tobler2015] reported that their chironomid count sums were at least fifty head capsules, a subsequent corrigendum  [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
+The risk that count sums will be mis-reported is not just theoretical: Larocque-Tobler et al [-@Larocque-Tobler2015] reported that their chironomid count sums were at least fifty head capsules; a subsequent corrigendum [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
 Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.
@@
 The key insight that allows inference about the count sum is that assemblage data are expected to follow the typical features of a rank abundance curve.
-In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual [@Coddington2009], hereafter singletons, unless the sampling effort is high relative to the taxonomic richness.
+In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual, hereafter singletons, unless the sampling effort is high relative to the taxonomic richness [@Coddington2009].
 In the occasional assemblages without singletons, the greatest common divisor should usually be one, i.e.
@@
 A range of ecological and palaeoecological data were sought to allow for difference in typical count sum and species richness.
-Datasets where the minimum count sum was not reported in the associated paper were excluded.
+Fossil pollen assemblage in the Neotoma database [@neotoma_2018] (downloaded `r readd(pollen_download_date)`) with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded.
+Datasets with percent data where the minimum count sum was not reported in the associated paper were excluded.
 Datasets discussed by Telford [-@Telford2019] as possibly having under-reported count sums were also excluded to avoid double reporting.
@@
 
-The `r n_distinct(readd(owens)$SampleId)` diatoms counts from Owen's Lake span a wide range of count sums (`r range(owens_summ$sum) %>% paste(collapse = "--")`) as, if concentrations were low, the requirement that at least 300 valves were counted, was replaced by a minimum area of slide to count [@Platt1997].
-Discounting the `r sum(owens_summ$n_taxa == 1)` mono-specific assemblages (all with very low count sums`r max(owens_summ$sum[owens_summ$n_taxa == 1]) %>% insist(.
-<= 10) %>% invisible()`), the median number of singletons per count is `r owens_summary2$median_sing %>% as.english()` and only `r owens_summary2$no_singletons`% of assemblages lack singletons.
-Assemblages with count sums below 50 have a higher chance of lacking singletons (`r owens_summary2$no_single_lt_50`% vs `r owens_summary2$no_single50_`%).
-Of the assemblages without singletons, `r owens_summary2$gcd2`% have a greatest common divisor of one.
-The maximum greatest common divisor is `r  owens_summary2$gcd_max %>% as.english()`.
-Assemblages with a greatest common divisor above one all have low taxonomic richness`r owens_summ %>% filter(gcd > 1) %>% pull(n_taxa) %>% max() %>%  insist(. <= 6) %>% invisible()`.
+The `r nrow(owens_summ1)` diatoms counts from Owen's Lake span a wide range of count sums (`r min(owens_summ1$count_sum)` -- `r max(owens_summ1$count_sum)`) as, if concentrations were low, the requirement that at least 300 valves were counted, was replaced by a minimum area of slide to count [@Platt1997].
+Discounting the `r sum(owens_summ1$n_taxa == 1)` mono-specific assemblages (all with very low count sums`r max(owens_summ1$count_sum[owens_summ1$n_taxa == 1]) %>% insist(.
+<= 10) %>% invisible()`), `r owens_summ$p_singletons`% of assemblages have singletons with a median number of singletons per assemblage of `r owens_summ$median_sing %>% as.english()`.
+Assemblages with count sums above 50 have a higher chance of having singletons (`r owens_summ$no_single_gt_50`% vs `r owens_summ$no_single50_`%).
+`r owens_summ$gcd2`% of assemblages have a greatest common divisor of one, the remainder have a greatest common divisor of `r owens_summ$gcd_max %>% as.english()`.
+Assemblages with a greatest common divisor above one all have low taxonomic richness`r owens_summ1 %>% filter(gcd > 1) %>% pull(n_taxa) %>% max() %>% insist(. <= 6) %>% invisible()`.
 
-The `r testate_summ$nsamples` assemblages in the North American testate amoeba training set with available count data [@AMESBURY2018] have count sums between `r testate_summary$count_summary$sum %>% range() %>% paste(collapse = " and ")` tests (median = `r testate_summary$count_summary$sum %>% median()`), and taxonomic richness ranges between `r testate_summary$count_summary$n_taxa %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$n_taxa %>% median()`) taxa.
-`r testate_summary$prop_singletons * 100`% of the assemblages have at least one singleton (median `r testate_summary$median_singletons`).
-In the assemblages that lack singletons, `r testate_summary$gcd1 * 100`% have a greatest common divisor of one.
-Of the `r testate_summ$n_gcd2plus %>% as.english()` assemblages with a greatest common divisor above one, `r testate_summ$n_gcd2_low_div %>% as.english()` have fewer than five taxa; `r testate_summ$n_gcd2 %>% as.english()` have a greatest common divisor of two, and the other `r testate_summ$n_gcd3 %>% as.english()` has a greatest common divisor of three.
+The `r testate_summ$n_assemblages` assemblages in the North American testate amoeba training set with available count data [@AMESBURY2018] have count sums between `r testate_summ$count_min` and `r testate_summ$count_max` tests (median = `r testate_summ$count_median`), and taxonomic richness ranges between `r testate_summ$tax_min` and `r testate_summ$tax_max` (median = `r testate_summ$tax_median`) taxa.
+`r testate_summ$p_singletons`% of the assemblages have at least one singleton (median `r testate_summ$median_singletons`) and `r testate_summ$p_gcd1`% of assemblages have a greatest common divisor of one.
+Of the `r testate_summ$n_gcd1 %>% as.english()` assemblages with a greatest common divisor above one, `r testate_summ$n_gcd2_low_div %>% as.english()` have fewer than five taxa; `r testate_summ$n_gcd2 %>% as.english()` have a greatest common divisor of two, and the other `r testate_summ$n_gcd3 %>% as.english()` has a greatest common divisor of three.
+
+The pollen data from the Neotoma database included over `r pollen_summ$n_assemblages` thousand assemblages in `r pollen_summ$n_datasets %>% format()` datasets. 
+Nearly all assemblages have singletons (`r pollen_summ$p_singletons`%) and a greatest common divisor of one (`r pollen_summ$p_gcd1`%). 
+The assemblages with a greatest common dividor above one are not randomly distributed among the datasets: `r pollen_summ$dataset_gcd1`% of the datasets have no such assemblages.
+Some data sets with a high proportion of assemblages with a greatest common divisor above are older datasets which may, despite the available metadata, have been digitised with the loss of rare taxa and precision.
+For example, one dataset includes `r pollenA$n_samp` assemblages with a median richness of `r pollenA$taxa_median` taxa.
+Count sums vary between `r pollenA$count_min` and `r pollenA$count_max`.
+The greatest common divisor is three for all assemblages, that is all `r pollenA$n_count` counts in the dataset are divisible by three.
+Excluding such datasets would further increase the proportion of assemblages with singletons. 
+Some other Neotoma datasets are discussed is a subsequent section.
 
@@
 
-Dataset pollen1 is a `r n_groups(readd(pollen1_counts))` count pollen stratigraphy.
-The published stratigraphy and the archived dataset generally resemble each other, but there are several discrepancies.
-The associated paper reports that count sums are 200 (excluding fern and fungal spores).
-The archived data have counts sums between `r pollen1_summary$count$min_count` and `r pollen1_summary$count$max2_count`, with the exception of one count of `r pollen1_summary$count$max_count` (which may be so high because of a possible data entry error).
-Count sums that are smaller than 200 may be because some rare taxa have been omitted from the archived data.
-This should normally only cause small discrepancies.
-More importantly, these count sums include trilete spores, probably from a fern or fern ally, with an abundance as high as `r pollen1_summary$max_trilete` spores.
-Unexpectedly, only `r pollen1_summary$count$has_singletons %>% insist(. == 1) %>% as.english()` of the counts has a singleton, a _`r pollen1_summary$individual$singleton_spp`_ grain.
-The other counts have minimum abundances of two or four pollen grains.
-With the exception of the single singleton, all `r pollen1_summary$individual$n` species counts are divisible by two`r pollen1_summary$individual$not_two %>% insist(. == 1) %>% invisible()`.
-The probability that these counts reflect the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled in some way.
+Some of the pollen data in the Neotoma database have an inexplicably high proportions of assemblages with a greatest common divisor greater than one. 
+Here I focus on four datasets produced by the same research group. 
+All the datasets, which are relatively recent, include information on the spike used to calculate pollen concentration, so cannot have been digitised.
+The four datasets have some assemblages with a greatest common divisor of two interspersed, sometimes regularly, amongst assembages with the expected greatest common divisor of one.
+The assemblages with a greatest common divisor of one typically have several singletons (Table \@ref(tab:neotoma-table))).
+
+```{r neotoma-table}
+pollen1234$table %>% knitr::kable(caption = "Caption")
+```
+
+Assemblages with a greatest common divisor of one typically have a higher taxonomic richness than assemblages with a greatest common divisor of two (Fig. \@ref(fig:neotoma-plot)).
+The latter have a richness that is typically of the former with count sums half as large.
+
+```{r neotoma-plot, fig.height = 100/24.5, fig.width=100/24.5, fig.cap="Number of taxa against count sum for assemblages with a greatest common divisor of one (red circles) or two (blue triangles).
+All four datasets show the same pattern, but it is clearest in datasets 1 and 2 which have the highest proportion of assemblages with a greatest common divisor of two."}
+pollen1234$figure %>% print()
+```
+
+One assemblage with a greatest common divisor of two has `r pollen1234$summ$max_richness` taxa. 
+Assuming the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $1/k^n$: for a single assemblage with this many taxa, the probability is 1 in 10^`r floor(log10(2^pollen1234$summ$max_richness))`^.
+The probability that these counts reflect the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled at some stage in some way.
 
@@
 
-My search through archived microfossil count data found several datasets where count sums are, or appear to be, smaller than that reported.
+Analysis of the breeding bird, Owen's Lake, testate amoeba and pollen data sets has shown that, as expected, the vast majority of community and assemblages counts should have singletons, and even more should have a greatest common divisor of one.
+This means that these characteristics are potentially useful for identifying data where the count sum is misreported.
+
+My search through archived microfossil data found several datasets where count sums are, or appear to be, smaller than that reported.
 Some of the count sums appear to be an order of magnitude below what was reported.
-The datasets examined are a convenience sample of available count data that met the inclusion criteria.
+The percent datasets examined are a convenience sample of available data that met the inclusion criteria.
 As such, this analysis cannot be used to accurately determine the prevalence of datasets with possible undercounts, but it appears that a non-negligible fraction of the literature is affected.
@@
 It is also possible that the apparently low count sums are a false positive. 
-This can only happen in assemblages that have a greatest common divisor greater than one.
-The analysis of the breeding bird, testate amoeba and Owen's Lake datasets shows that assemblages without singletons are rare, and assemblages with a greatest common divisor greater than one are very rare, especially for taxonomically diverse assemblages.
+This can only happen in assemblages that have a greatest common divisor greater than one, which is very rare, especially for taxonomically diverse assemblages.
 If a dataset contains several taxonomically diverse assemblages that have a greatest common divisor of two (or expecially if greater than two) doubts should be raised about whether the counts sums are accurately described.

### 2019-05-07 R/download_diatoms.R

@@ -5,22 +5,25 @@ diatom_plan <- drake_plan(
     gsub(pattern = "<.*?>", replacement = "", x = .) %>% 
     read_delim(delim = "\t"),
   
-  owens_summ = owens %>% 
+  owens_summ1 = owens %>% 
   group_by(SampleId) %>% 
-  summarise(mn = min(Count), n_sing = sum(Count == 1), sum = sum(Count), n_taxa = n(), gcd = ifelse(n_taxa > 1, numbers::mGCD(Count), NA)),
+  summarise(mn = min(Count), n_sing = sum(Count == 1), count_sum = sum(Count), n_taxa = n(), gcd = ifelse(n_taxa > 1, numbers::mGCD(Count), NA)),
   
-  owens_summary2 = owens_summ %>% 
+  owens_summ = owens_summ1 %>% 
     filter(n_taxa > 1) %>% 
     summarise(
       n = n(),
+      count_min = min(count_sum),
+      count_max = max(count_sum),
       median_sing = median(n_sing),
+      p_singletons  = mean(n_sing  > 0) * 100,
       no_singletons = mean(mn != 1) * 100,
-      no_single_lt_50 = mean(mn[sum < 50] != 1) * 100, 
-      no_single50_ = mean(mn[sum >= 50] != 1) * 100, 
-      gcd2 = mean(gcd[mn > 1] == 1) * 100,
+      no_single_gt_50 = mean(mn[count_sum > 50] == 1) * 100, 
+      no_single50_ = mean(mn[count_sum <= 50] == 1) * 100, 
+      gcd2 = mean(gcd == 1) * 100,
       gcd_max = max(gcd, na.rm = TRUE)),
   
-  ####owens diatoms####
+  ####owens diatoms percent####
   owens_est = owens %>% 
     estimate_n(ID_cols = c("SampleId", "SampleCode", "CountSum"), percent_col = "Percent", taxon_col = "TaxonShortName", digits = 4) %>%
     assertr::verify(map_int(direct_search_est, nrow) == 1) %>% #check only one row in each direct_search_est

### 2019-05-07 R/download_neotoma.R

@@ -1,0 +1,150 @@
+
+#source functions
+source("R/get_sites_meta.R")
+
+#### drake plan ####
+neotoma_plan <- drake_plan(
+  #get dataset list
+  pollen_sites = get_dataset(datasettype = "pollen"),
+
+  #download_data
+  pollen_data = pollen_sites %>% get_download(), #SLOW
+  pollen_download_date = {
+    pollen_data
+    lubridate::today()#datestamp
+  },
+  
+  #extract pollen counts
+  pollen = pollen_data %>% 
+    #remove elements known to be backtransformed from diagrams
+    #1776 mud lake - cohmap documentation
+    list_modify(`1776` = NULL) %>% 
+    map_df(~mk_thin_pollen(counts(.), .$taxon.list), .id = "datasetID") %>% 
+    filter(ecological.group %in% pollen_wanted, variable.units == "NISP") %>% 
+    assert(in_set("pollen"), variable.element) %>% 
+    select(-variable.units, -variable.element) %>% 
+    mutate(
+      datasetID = as.integer(datasetID), 
+      sampleID = as.integer(sampleID)
+          ) %>% 
+    group_by(datasetID, sampleID) %>% 
+    #remove very low count sums and monospecific assemblages
+    filter(sum(count) >= 50, n() > 1) %>%
+    #round slightly to correct near-integers backtransformed? from percent
+    mutate(count = round(count, 3)) %>% 
+    #filter out counts of 0.01 (in datasetID 17391 - presumably slide scanning for rareties)
+    filter(count != 0.01) %>% 
+    #filter out two datasets with probable percent data
+    filter(!datasetID %in% c(15059, 25609)) %>% 
+    mutate(count = ceiling(count)), 
+
+
+  #process
+  pollen_summ = tibble(
+    n_datasets = n_distinct(pollen$datasetID),
+    n_assemblages = floor(n_distinct(pollen$sampleID)/1000),
+    p_singletons = mean(pollen_summ1$n_singletons > 0) * 100,
+    p_gcd1 = mean(pollen_summ1$gcd == 1) * 100,
+    n_gcd1 = sum(pollen_summ1$gcd > 1),
+    threshold = 0.5, 
+    n_gcd1_high = pollen_summ1 %>% 
+      summarise(dgcd = mean(gcd == 1),  n = n()) %>% 
[…]

### 2019-05-07 R/download_testates.R

@@ -32,23 +32,30 @@ testate_plan <- drake_plan(
       cnt[, names(cnt) %in% tl$taxon.name]
       })) %>%
     map_df(gather, key = taxon, value = count, .id = "sampleID") %>% 
-    mutate(sampleID = as.numeric(sampleID)),
+    mutate(sampleID = as.numeric(sampleID)) %>% 
+    group_by(sampleID),
   
   #run analyses
-  testate_summary = summarise_counts(testate_counts),
+  testate_summ1 = testate_counts %>% 
+    summarise(count_sum = sum(count), n_taxa = n(), gcd = numbers::mGCD(count), n_singletons = sum(count == 1), min = min(count)) %>% 
+    assert(within_bounds(1, 3), gcd),
 
-  testate_summ = list(
-    nsamples = n_distinct(testate_counts$sampleID),
-    n_gcd2plus = nrow(testate_summary$gcd2plus),
-    n_gcd2 = sum(testate_summary$gcd2plus$gcd == 2),
-    n_gcd3 = sum(testate_summary$gcd2plus$gcd == 3),
-    n_gcd4 = sum(testate_summary$gcd2plus$gcd > 3) %>% insist(. == 0),
-    n_gcd2_low_div = sum(testate_summary$gcd2plus$n_taxa < 5) 
-  )
+ 
+  testate_summ = testate_summ1 %>% 
+     summarise(
+       n_assemblages = n(),
+       count_max = max(count_sum),
+       count_min = min(count_sum),
+       count_median = median(count_sum),
+       tax_min = min(n_taxa),
+       tax_max = max(n_taxa),
+       tax_median = median(n_taxa),
+       p_singletons  = mean(n_singletons > 0) * 100,
+       median_singletons = median(n_singletons),
+       p_gcd1 = mean(gcd == 1) * 100,
+       n_gcd1 = sum(gcd > 1),
+       n_gcd2 = sum(gcd == 2),
+       n_gcd3 = sum(gcd == 3),
+       n_gcd2_low_div = sum(gcd[n_taxa < 5] > 1)
+     )
 )
-
-# testate_config <- drake_config(testate_plan)
-# 
-# make(testate_plan, jobs = 2)
-# vis_drake_graph(testate_config, targets_only = TRUE)
-

### 2019-05-07 make_count_check_ms_with_drake.R

@@ -2,12 +2,15 @@
 library("drake")
 library("tidyverse")
 library("readr")
+library("readxl")
 #library("gridExtra")
 #library("broom")
-#devtools::install_github("richardjtelford/rjt.misc")
-library("rjt.misc")
 #devtools::install_github("richardjtelford/count_checker")#need to make package and repo names identical (_ not permitted)
 library("countSum")
+library("assertr")
+#devtools::install_github("richardjtelford/rjt.misc")
+library("rjt.misc")
+library("neotoma")
 
 #drake configuration
 pkgconfig::set_config("drake::strings_in_dots" = "literals")
@@ -21,10 +24,9 @@ options(askpass = function(x)readLines("pw.txt"))
 source("R/download_birds.R")
 source("R/download_testates.R")
 source("R/download_chironomids.R")
-source("R/download_pollen.R")
+source("R/download_neotoma.R")
 source("R/download_diatoms.R")
 source("R/download_marine.R")
-source("R/summarise_counts.R")
 
 
 #construct drake plan
@@ -33,6 +35,9 @@ analyses <- drake_plan(
   #secrets
   secrets = readRDS(file_in("data/secrets.RDS")) %>%
       encryptr::decrypt(secret),
+  neotoma_secrets = readRDS(file_in("data/neotoma_secrets.RDS")) %>%
+    encryptr::decrypt(datasetID) %>% 
+    mutate(datasetID = as.integer(datasetID)),
   
   #make plots
   
@@ -58,7 +63,7 @@ plans <- bind_rows(
   bird_plan, 
   testate_plan, 
   chironomid_plan, 
-  pollen_plan,
+  neotoma_plan,
   diatom_plan,
   marine_plan,
   analyses)

### 2019-05-07 Rmd/count_check_MS.Rmd
changed sign of bird analyses

@@
 theme_set(theme_bw(base_size = 12))
-options(digits = 1)
+options(digits = 2)
 
@@
 
-A tiny minority of the over `r format(floor(bird_singleton_summary$n_routes[1]/1000) * 1000, big.mark = ",")` bird counts from the North American breeding bird survey [@NAmerican_Breeding_Bird2018]
-lack singletons at the species level (Table \@ref(tab:bird-table))).
+The vast majority of the over `r format(floor(bird_singleton_summary$n_routes[1]/1000) * 1000, big.mark = ",")` bird counts from the North American breeding bird survey [@NAmerican_Breeding_Bird2018]
+have singletons at the species level (Table \@ref(tab:bird-table))).
 To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the birds counts to progressively lower taxonomic resolutions.
-As richness declines, the proportion of counts lacking singletons increases (Table \@ref(tab:bird-table)), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
-The vast majority of counts without singletons have a greatest common divisor of one, except when taxonomic richness is low.
+As richness declines, the proportion of counts having singletons declines (Table \@ref(tab:bird-table)), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
+The vast majority of counts have a greatest common divisor of one, except when taxonomic richness is low.
 
 ```{r bird-table}
+options(digits = 4)
 bird_singleton_summary %>%
-  select(`Taxonomic level` = taxonomic_level, `Percent without singletons` = pc_singletons, `Percent GCD > 1` = GCD2plus, `Mean richness` = mean_richness) %>%
-  knitr::kable(digits = c(0, 4, 4, 3), caption = "Percent of bird counts without singletons and of greatest common divisor (GCD) greater than one of counts without singletons, at different taxonomic levels.")
+  select(`Taxonomic level` = taxonomic_level, `Percent with singletons` = p_singletons, `Percent GCD = 1` = GCD1, `Median richness` = median_richness) %>%
+  knitr::kable(digits = c(0, 6, 6, 3), caption = "Percent of bird counts with singletons and of greatest common divisor (GCD) of one, at different taxonomic levels.")
+options(digits = 2)
 ```
@@
 To test the sensitivity of the prevalence of counts without singletons to the count sum, I re-sample the counts with 400 or more observations to smaller count sums using a multivariate hypergeometric distribution.
-At the order level, the proportion of counts without singletons declines steeply until it reaches a minimum at about 200, thereafter is rises slowly (Fig. \@ref(fig:bird-recount)).
-The prevalence of counts without singletons is high for small counts because it is possible that only the common taxa are counted; with larger counts, the chance of counting a single individual of a rare taxon increases.
-With even larger counts the prevalence of counts without singletons increases due to saturation as singletons become doubletons and few new taxa are found.
+At the order level, the proportion of counts with singletons increases steeply until it reaches a maximum at about 200, thereafter is declines slowly (Fig. \@ref(fig:bird-recount)).
+The prevalence of counts with singletons is low for small counts because it is possible that only the common taxa are counted; with larger counts, the chance of counting a single individual of a rare taxon increases.
+With even larger counts the prevalence of counts with singletons decreases due to saturation as singletons become doubletons and few new taxa are found.
 

### 2019-05-07 R/bird_functions.R

@@ -28,7 +28,7 @@ bird_process <- function(bird_data, bird_species) {
       n_taxa = n(),
       singletons = sum(count == 1),
       count_sum = sum(count), 
-      GCD = ifelse(singletons == 0 & n_taxa > 1, numbers::mGCD(count), NA)#only find GCD where no singletons for speed ( and > 1 taxa). Using ifelse as it doesn't run TRUE code when it would crash
+      GCD = ifelse(n_taxa > 1, numbers::mGCD(count), NA)#only find GCD where no singletons for speed ( and > 1 taxa). Using ifelse as it doesn't run TRUE code when it would crash
     )
   return(out)
 }

### 2019-05-07 R/download_birds.R

@@ -38,10 +38,11 @@ bird_plan <- drake_plan(
   bird_singleton_summary = bird_singletons %>%
     group_by(taxonomic_level) %>%
     summarise(
-      pc_singletons = mean(singletons == 0) * 100,
+      p_singletons = mean(singletons > 0) * 100,
       n_singletons = sum(singletons == 0),
-      mean_richness = mean(n_taxa),
+      median_richness = median(n_taxa),
       n_routes = n(),
+      GCD1 = mean(GCD == 1, na.rm = TRUE) * 100,
       GCD2plus = mean(GCD > 1, na.rm = TRUE) * 100,
       GCD_max = max(GCD, na.rm = TRUE)
     ) %>%
@@ -51,12 +52,12 @@ bird_plan <- drake_plan(
     )) %>%
     arrange(taxonomic_level),
   
-  #bird order GCD summary
-  bird_order_GCD = bird_singletons %>% 
-    ungroup() %>% 
-    filter(singletons == 0, taxonomic_level == "order", !is.na(GCD)) %>%
-    group_by(n_taxa) %>% 
-    summarise(t = sum(GCD > 1, na.rm = TRUE), n = n()),
+  # #bird order GCD summary
+  # bird_order_GCD = bird_singletons %>% 
+  #   ungroup() %>% 
+  #   filter(singletons == 0, taxonomic_level == "order", !is.na(GCD)) %>%
+  #   group_by(n_taxa) %>% 
+  #   summarise(t = sum(GCD > 1, na.rm = TRUE), n = n()),
   
   #simulate counts with mv hypergeometric
   bird_recount_singletons = bird_data %>% 
@@ -70,14 +71,14 @@ bird_plan <- drake_plan(
     unnest(recount) %>% 
     mutate(new_count_sum = as.numeric(new_count_sum)) %>% 
     group_by(new_count_sum, rep) %>% 
-    summarise(m = mean(no_singletons == 0)) %>% 
+    summarise(m = mean(no_singletons > 0) * 100) %>% 
     summarise(m_no_singletons = mean(m), sd = sd(m), se = sd/sqrt(n())),
   
   bird_recount_singleton_plot = bird_recount_singletons %>% 
     ggplot(aes(x = new_count_sum, y = m_no_singletons, ymin = m_no_singletons - 1.96 * se, ymax = m_no_singletons + 2 * se)) + 
     geom_errorbar() + 
     geom_point(size = 0.5) +
[…]

### 2019-05-07 R/download_neotoma.R

@@ -97,7 +97,7 @@ neotoma_plan <- drake_plan(
     table <- pollen_summ1 %>% 
       inner_join(datasets, by  = "datasetID") %>% 
       group_by(dataset) %>% 
-      summarise(`Number of assemblages` = n(), `Median number of taxa` = median(n_taxa), GCD = mean(gcd == 1), `Median number of singletons (GCD = 1)` = median(n_singletons[gcd == 1])) 
+      summarise(`No. assemblages` = n(), `Median no. taxa` = median(n_taxa), `GCD = 1` = mean(gcd == 1), `Median no. singletons (GCD = 1)` = median(n_singletons[gcd == 1])) 
 
     #figure    
     figure <- pollen_summ1 %>% 

### 2019-05-07 R/summarise_counts.R

@@ -1,4 +1,4 @@
-loadd(testate_counts)
+#loadd(testate_counts)
 
 summarise_counts <- function(counts){#columns sampleID/taxon/count
   results <- list()

### 2019-05-08 R/summarise_counts.R
tidying files

@@ -1,37 +1,0 @@
-#loadd(testate_counts)
-
-summarise_counts <- function(counts){#columns sampleID/taxon/count
-  results <- list()
-  
-  results$count_summary <- counts %>% 
-    group_by(sampleID) %>% 
-    summarise(min = min(count), n_taxa = n(), sum = sum(count), singletons = sum(count == 1))
-  
-  results$prop_singletons <- results$count_summary %>% 
-    summarise(m = mean(min == 1)) %>% 
-    pull(m)
-
-  results$median_singletons <- results$count_summary %>% 
-    summarise(m = median(singletons)) %>% 
-    pull(m)
-
-    #samples without singletons  
-  
-
-  
-  results$no_singletons <- counts %>% 
-    group_by(sampleID) %>%
-    mutate(m = min(count)) %>%
-    filter(m > 1) %>% #remove samples with singletons
-    summarise(gcd = numbers::mGCD(count), n_taxa = n()) #find greatest common divisor
-    
-  results$gcd2plus <- results$no_singletons %>%
-    filter(gcd > 1)
-  
-  results$gcd1 <- results$no_singletons %>% 
-    summarise(prop1 = mean(gcd == 1)) %>% 
-    pull(prop1)
-
-  return(results)
-  
-}

### 2019-05-08 make_count_check_ms_with_drake.R

@@ -39,15 +39,12 @@ analyses <- drake_plan(
     encryptr::decrypt(datasetID) %>% 
     mutate(datasetID = as.integer(datasetID)),
   
-  #make plots
-  
   #add extra packages to bibliography
   biblio2 = package_citations(
     packages = c("extraDistr", "countSum", "numbers"), 
     old_bib = file_in("Rmd/extra/countMS.bib"), 
     new_bib = file_out("Rmd/extra/countMS2.bib")),
   
-  #knit manuscript
   #knit manuscript
   manuscript = {
     file_in("Rmd/extra/countMS2.bib")

### 2019-05-08 Rmd/count_check_MS.Rmd
typos

@@
 Nearly all assemblages have singletons (`r pollen_summ$p_singletons`%) and a greatest common divisor of one (`r pollen_summ$p_gcd1`%). 
-The assemblages with a greatest common dividor above one are not randomly distributed among the datasets: `r pollen_summ$dataset_gcd1`% of the datasets have no such assemblages.
+The assemblages with a greatest common divisor above one are not randomly distributed among the datasets: `r pollen_summ$dataset_gcd1`% of the datasets have no such assemblages.
 Some data sets with a high proportion of assemblages with a greatest common divisor above are older datasets which may, despite the available metadata, have been digitised with the loss of rare taxa and precision.
@@
 The Owen's Lake dataset [@Platt1997] also includes percentages for each taxa, allowing the count sums estimated from the percent data to be verified.
-Excluding monospecific assemblages, for which no meaningful estimate of the count sum is possible, the minimum percent method and the direct search method correctly estimate the count sum for, respectively, `r owens_est_summ$minpc`% and `r owens_est_summ$direct`% of the assemblages.
-The few assemblages that fail the direct seach method are species poor and have low count sums.
+Excluding mono-specific assemblages, for which no meaningful estimate of the count sum is possible, the minimum percent method and the direct search method correctly estimate the count sum for, respectively, `r owens_est_summ$minpc`% and `r owens_est_summ$direct`% of the assemblages.
+The few assemblages that fail the direct search method are species poor and have low count sums.
 
@@
 All the datasets, which are relatively recent, include information on the spike used to calculate pollen concentration, so cannot have been digitised.
-The four datasets have some assemblages with a greatest common divisor of two interspersed, sometimes regularly, amongst assembages with the expected greatest common divisor of one.
+The four datasets have some assemblages with a greatest common divisor of two interspersed, sometimes regularly, amongst assemblages with the expected greatest common divisor of one.
 The assemblages with a greatest common divisor of one typically have several singletons (Table \@ref(tab:neotoma-table))).
@@
 This can only happen in assemblages that have a greatest common divisor greater than one, which is very rare, especially for taxonomically diverse assemblages.
-If a dataset contains several taxonomically diverse assemblages that have a greatest common divisor of two (or expecially if greater than two) doubts should be raised about whether the counts sums are accurately described.
+If a dataset contains several taxonomically diverse assemblages that have a greatest common divisor of two (or especially if greater than two) doubts should be raised about whether the counts sums are accurately described.
 
@@
 These valves usually separate during processing, but paired valves are often found, and some taxa such as _Aulacoseira_ produce long chains of strongly bound valves.
-This might make singletons slighly less likely. 
+This might make singletons slightly less likely. 
 The results from Owen's Lake suggest that this is not an important problem, as most assemblages have many singletons.
@@
 Such undercounts might constitute a data handling error, for example, if samples with low counts were supposed to be merged but that step was forgotten.
-Substantial and pervasive undercounting could be construed as scientific misconduct due to either negligence or falsification, and action to correct the literature is probably required.
+Substantial and pervasive under-counting could be construed as scientific misconduct due to either negligence or falsification, and action to correct the literature is probably required.
 

### 2019-05-18 Rmd/count_check_MS.Rmd
chironomid2 added

@@
 ---
-title: "Tools for identifying unexpectely low microfossil count sums"
+title: "Tools for identifying unexpectedly low microfossil count sums"
 author:
@@
 Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so.
-This paper introduces methods that can flag assemblages with potentially misreported count sums and finds that some assemblage datasets that fail these tests.
+This paper introduces methods that can flag assemblages with potentially misreported count sums and finds that some assemblage datasets fail these tests.
 journal: "Quaternary Science Reviews"
@@
 loadd(owens_summ1, owens_summ, owens_est_summ)
-loadd(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ)
+loadd(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ, chironomid2_summ)
 
@@
 In the occasional assemblages without singletons, the greatest common divisor should usually be one, i.e.
-few assemblages should have all counts divisable by an integer larger than one, especially if species richness is high.
+few assemblages should have all counts divisible by an integer larger than one, especially if species richness is high.
 
@@
 
-Analysis of the breeding bird, Owen's Lake, testate amoeba and pollen data sets has shown that, as expected, the vast majority of community and assemblages counts should have singletons, and even more should have a greatest common divisor of one.
-This means that these characteristics are potentially useful for identifying data where the count sum is misreported.
+Analysis of the breeding bird, Owen's Lake diatom, testate amoeba and pollen data sets has shown that, as expected, the vast majority of community and assemblages counts have singletons, and an even larger proportion have a greatest common divisor of one.
+Therefore, these characteristics are potentially useful for identifying data where the count sum is misreported.
 
@@
 It is also possible that the apparently low count sums are a false positive. 
-This can only happen in assemblages that have a greatest common divisor greater than one, which is very rare, especially for taxonomically diverse assemblages.
-If a dataset contains several taxonomically diverse assemblages that have a greatest common divisor of two (or especially if greater than two) doubts should be raised about whether the counts sums are accurately described.
+However, as taxonomically diverse assemblages with a greatest common divisor greater than one are rare and all the datasets with apparently low count sums contained multiple taxonomically diverse assemblages with apparently low count sums, this is unlikely.
+
+These tests require the assemblage data which are often not archived. 
+However, if the count sums are very low it is possible to recognise this from a stratigraphic diagram. 
+For example, a chironomid stratigraphy had an assemblage where all five taxa had a relative abundance of about 20%, suggesting a count sum of five, far below the reported fifty chironomids.
+Another assemblage had six taxa with a relative abundance of just over 10% and a seventh with twice as much, suggesting a count sum of eight chironomids.
+On request, the authors sent the data. 
+None of the `r chironomid2_summ$n_samp` assemblages in the dataset had more fifty chironomids: the count sums varied between `r chironomid2_summ$count_min` and  `r chironomid2_summ$count_max`.
+
+## Challenges
 
@@
 The results from Owen's Lake suggest that this is not an important problem, as most assemblages have many singletons.
+
+## Consequences
 

### 2019-05-18 R/download_chironomids.R

@@ -67,5 +67,13 @@ chironomid_plan <- drake_plan(
       pull(n_taxa) %>% 
       range()
   
-  )
+  ),
+  
+  #chironomid2
+  chironomid2_summ = read_csv(file_in("data/chironomid2.csv")) %>% 
+    select(ID, starts_with("Chironomid")) %>% 
+    gather(key = taxon, value = count, -ID) %>% 
+    group_by(ID) %>% 
+    summarise(count_sum = sum(count), no_singletons = sum(count == 1)) %>% 
+    summarise(n_samp = n(), count_min = min(count_sum), count_max = max(count_sum), min_singletons = min(no_singletons))
 )

### 2019-05-18 Rmd/count_check_MS.Rmd
Cathy's edits

@@
 Carlisle [-@Carlisle2017] identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data.
-Brown et al. [-@Brown2017] developed the granularity-related inconsistency means (GRIM) test of whether means of integer data are consistent with the reported sample size.
+Brown et al. [-@Brown2017] developed the granularity-related inconsistency of means (GRIM) test which assesses whether means of integer data are consistent with the reported sample size.
 However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data.
@@
 A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often fifty for chironomids and several hundred for pollen and diatoms.
-This is important because larger count sums are associated with smaller uncertainties, both in relative abundance of taxa and derived statistics such are transfer function reconstructions [@Heiri2001].
+This is important because larger count sums are associated with smaller uncertainties, both in relative abundance of taxa and derived statistics such as transfer function reconstructions [@Heiri2001; @Payne2009].
 However, mistakes happen and metadata such as count sums can be forgotten once percentages are calculated.
 In addition, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there may be an incentive to misreport the minimum count sum.
-The risk that count sums will be mis-reported is not just theoretical: Larocque-Tobler et al [-@Larocque-Tobler2015] reported that their chironomid count sums were at least fifty head capsules; a subsequent corrigendum [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
+The risk that count sums will be misreported is not just theoretical: Larocque-Tobler et al [-@Larocque-Tobler2015] reported that their chironomid count sums were at least fifty head capsules; a subsequent corrigendum [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
 Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.
@@
 This method will fail for assemblages without singletons.
-We also expect the percent _p_ to be calculated from integer counts, therefore, there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
+We also expect the percent _p_ to be calculated from integer counts, therefore there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
 Possible values for _N_ can be found by a direct search algorithm over the range of plausible values of _N_.
@@
 Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources.
-A range of ecological and palaeoecological data were sought to allow for difference in typical count sum and species richness.
+A range of ecological and palaeoecological data were sought to allow for differences in typical count sums and species richness.
 Fossil pollen assemblage in the Neotoma database [@neotoma_2018] (downloaded `r readd(pollen_download_date)`) with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded.
@@
 Datasets discussed by Telford [-@Telford2019] as possibly having under-reported count sums were also excluded to avoid double reporting.
-Datasets with possible mis-reporting are anonymised, but no attempt is made to diagnose whether errors or misconduct are responsible.
+Datasets with possible misreporting are anonymised, but no attempt is made to diagnose whether errors or misconduct are responsible.
 This paper does not attempt to be an exhaustive survey of all the data available.
@@
 
-To test the sensitivity of the prevalence of counts without singletons to the count sum, I re-sample the counts with 400 or more observations to smaller count sums using a multivariate hypergeometric distribution.
+To test the sensitivity of the prevalence of counts without singletons to the count sum, I resample the counts with 400 or more observations to smaller count sums using a multivariate hypergeometric distribution.
 At the order level, the proportion of counts with singletons increases steeply until it reaches a maximum at about 200, thereafter is declines slowly (Fig. \@ref(fig:bird-recount)).
@@
 
-```{r bird-recount, fig.cap="Effect of count sum on the proportion of bird counts without singleton orders.
-Results are the mean of ten trials, error bars are two standard errors.", fig.width = 70/25.4, fig.height = 70/25.4}
+```{r bird-recount, fig.cap="Effect of count sum on the proportion of bird counts without singleton at the order level.
+Results are the mean of ten trials, error bars are two standard errors.", fig.width = 80/25.4, fig.height = 70/25.4}
 readd(bird_recount_singleton_plot)
@@
 
-The `r nrow(owens_summ1)` diatoms counts from Owen's Lake span a wide range of count sums (`r min(owens_summ1$count_sum)` -- `r max(owens_summ1$count_sum)`) as, if concentrations were low, the requirement that at least 300 valves were counted, was replaced by a minimum area of slide to count [@Platt1997].
+The `r nrow(owens_summ1)` diatom counts from Owen's Lake span a wide range of count sums (`r min(owens_summ1$count_sum)` -- `r max(owens_summ1$count_sum)`) because, if concentrations were low, the requirement that at least 300 valves were counted was replaced by a minimum area of slide to count [@Platt1997].
 Discounting the `r sum(owens_summ1$n_taxa == 1)` mono-specific assemblages (all with very low count sums`r max(owens_summ1$count_sum[owens_summ1$n_taxa == 1]) %>% insist(.
@@
 
-The pollen data from the Neotoma database included over `r pollen_summ$n_assemblages` thousand assemblages in `r pollen_summ$n_datasets %>% format()` datasets. 
+The pollen data from the Neotoma database included over `r pollen_summ$n_assemblages` assemblages in `r pollen_summ$n_datasets %>% format()` datasets. 
 Nearly all assemblages have singletons (`r pollen_summ$p_singletons`%) and a greatest common divisor of one (`r pollen_summ$p_gcd1`%). 
-The assemblages with a greatest common divisor above one are not randomly distributed among the datasets: `r pollen_summ$dataset_gcd1`% of the datasets have no such assemblages.
-Some data sets with a high proportion of assemblages with a greatest common divisor above are older datasets which may, despite the available metadata, have been digitised with the loss of rare taxa and precision.
+The assemblages with a greatest common divisor above one are not randomly distributed among the datasets: only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
+Some datasets with a high proportion of assemblages with a greatest common divisor above one are older datasets which may, despite the available metadata, have been digitised with the loss of rare taxa and precision.
 For example, one dataset includes `r pollenA$n_samp` assemblages with a median richness of `r pollenA$taxa_median` taxa.
@@
 
-The Owen's Lake dataset [@Platt1997] also includes percentages for each taxa, allowing the count sums estimated from the percent data to be verified.
+The Owen's Lake dataset [@Platt1997] includes percentages for each taxa, allowing the count sums estimated from the percent data to be verified.
 Excluding mono-specific assemblages, for which no meaningful estimate of the count sum is possible, the minimum percent method and the direct search method correctly estimate the count sum for, respectively, `r owens_est_summ$minpc`% and `r owens_est_summ$direct`% of the assemblages.
@@
 ```{r last-chance-plot, fig.cap = "Estimated and reported chironomid count sums by the minumum percent method (solid symbols with error bars) and the direct search method (red crosses) from Last Chance Lake [@Axford2017].
-Lines show the 1:1 (solid) and 2:1 (dashed) relationships.", fig.width = 90/25.4, fig.height = 90/25.4}
+Lines show the 1:1 (solid) and 2:1 (dashed) relationships.", fig.width = 80/25.4, fig.height = 80/25.4}
 readd(last_chance_plot)
@@
 
-Some of the pollen data in the Neotoma database have an inexplicably high proportions of assemblages with a greatest common divisor greater than one. 
+Some of the pollen datasets in the Neotoma database have an inexplicably high proportion of assemblages with a greatest common divisor greater than one. 
 Here I focus on four datasets produced by the same research group. 
@@
 The four datasets have some assemblages with a greatest common divisor of two interspersed, sometimes regularly, amongst assemblages with the expected greatest common divisor of one.
-The assemblages with a greatest common divisor of one typically have several singletons (Table \@ref(tab:neotoma-table))).
+The assemblages with a greatest common divisor of one typically have several singletons (Table \@ref(tab:neotoma-table)).
 
 ```{r neotoma-table}
-pollen1234$table %>% knitr::kable(caption = "Caption")
+pollen1234$table %>% knitr::kable(caption = "The number of assemblages, median number of taxa, percent of assemblages with a greatest common divisor (GCD) of one, and the median number of singletons in assemblages with a greatest common divisor of one, in four pollen datasets")
 ```
@@
 Assemblages with a greatest common divisor of one typically have a higher taxonomic richness than assemblages with a greatest common divisor of two (Fig. \@ref(fig:neotoma-plot)).
-The latter have a richness that is typically of the former with count sums half as large.
+In the datasets with a wide range of count sums, the richness of the assemblages with a greatest common divisor of two is comparable with the richness of the assemblages with a greatest common divisor of one that have a count sum that is half as large.
 
-```{r neotoma-plot, fig.height = 100/24.5, fig.width=100/24.5, fig.cap="Number of taxa against count sum for assemblages with a greatest common divisor of one (red circles) or two (blue triangles).
-All four datasets show the same pattern, but it is clearest in datasets 1 and 2 which have the highest proportion of assemblages with a greatest common divisor of two."}
+```{r neotoma-plot, fig.height = 70/24.5, fig.width=80/24.5, fig.cap="Number of taxa in against count sum for assemblages in two pollen datasets with a greatest common divisor of one (red circles) or two (blue triangles).
+Lines are loess with a span of 1.
+All four pollen datasets show the same pattern, but it is clearest in datasets 1 and 2 which have the highest proportion of assemblages with a greatest common divisor of two.", message = FALSE}
 pollen1234$figure %>% print()
@@
 The direct search and minimum percent methods agree, within rounding error, `r diatom2_summ$meth_agree %>% insist(isTRUE(.)) %>% invisible()`on the estimated count sums, implying that either all assemblages have singletons and count sums are low in some assemblages, or that the greatest common divisor of the raw counts is greater than one.
-This means that for the assemblage with the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc %>% as.english()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.
+This means that for the assemblage where the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc %>% as.english()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.
 
 Dataset chironomid1 has `r chironomid1_summ$nsamples` assemblages.
-Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 %>% as.english` assemblages had count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
+Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 %>% as.english` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
 These assemblages are diverse (richness between `r chironomid1_summ$richness %>% paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.
@@
 Palaeoceanographic dataset marine1 has `r marine1_summ$nsamples` assemblages with a median taxonomic richness of `r marine1_summ$median_taxa` taxa.
-The associated paper reports that assemblages with count sums below 100 were omitted, and that there were 20 assemblages with count sums below 200.
+The associated paper reports that assemblages with count sums below 100 were omitted.
 The direct search method estimates that `r marine1_summ$n100 %>% as.english()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa %>% as.english()` mono-specific assemblage), with count sums as low as `r marine1_summ$est_n_min`.
@@
 
-Analysis of the breeding bird, Owen's Lake diatom, testate amoeba and pollen data sets has shown that, as expected, the vast majority of community and assemblages counts have singletons, and an even larger proportion have a greatest common divisor of one.
+Analysis of the breeding bird, Owen's Lake diatom, testate amoeba and pollen datasets has shown that, as expected, the vast majority of community and assemblage counts have singletons, and an even larger proportion have a greatest common divisor of one.
 Therefore, these characteristics are potentially useful for identifying data where the count sum is misreported.
@@
 Some of the count sums appear to be an order of magnitude below what was reported.
-The percent datasets examined are a convenience sample of available data that met the inclusion criteria.
+The non-pollen datasets examined here are a convenience sample of available data that met the inclusion criteria.
 As such, this analysis cannot be used to accurately determine the prevalence of datasets with possible undercounts, but it appears that a non-negligible fraction of the literature is affected.
@@
 It is also possible that the apparently low count sums are a false positive. 
-However, as taxonomically diverse assemblages with a greatest common divisor greater than one are rare and all the datasets with apparently low count sums contained multiple taxonomically diverse assemblages with apparently low count sums, this is unlikely.
+However, as taxonomically diverse assemblages with a greatest common divisor greater than one are rare and all the datasets with apparently low count sums contained several such assemblages, this is unlikely.
 
@@
 On request, the authors sent the data. 
-None of the `r chironomid2_summ$n_samp` assemblages in the dataset had more fifty chironomids: the count sums varied between `r chironomid2_summ$count_min` and  `r chironomid2_summ$count_max`.
+None of the `r chironomid2_summ$n_samp` assemblages in the dataset had more that fifty chironomids: the count sums varied between `r chironomid2_summ$count_min` and  `r chironomid2_summ$count_max`.
 
@@
 It should be possible to identify such cases from the meta-data and knowledge of usual practice with different proxies. 
-It will also fail, as show by dataset marine1, if percentages are incorrectly calculated or rounded.
+It will also fail, as shown by dataset marine1, if percentages are incorrectly calculated or rounded.
 
 Some taxonomic groups have microfossils that often come in groups of attached individuals.
-For example, for each diatom cell has two valves which are the counting unit.
+For example, each diatom cell has two valves which are the counting unit.
 These valves usually separate during processing, but paired valves are often found, and some taxa such as _Aulacoseira_ produce long chains of strongly bound valves.

### 2019-05-18 R/download_neotoma.R

@@ -42,7 +42,7 @@ neotoma_plan <- drake_plan(
   #process
   pollen_summ = tibble(
     n_datasets = n_distinct(pollen$datasetID),
-    n_assemblages = floor(n_distinct(pollen$sampleID)/1000),
+    n_assemblages = (floor(n_distinct(pollen$sampleID)/1000) * 1000) %>% format(big.mark = ","),
     p_singletons = mean(pollen_summ1$n_singletons > 0) * 100,
     p_gcd1 = mean(pollen_summ1$gcd == 1) * 100,
     n_gcd1 = sum(pollen_summ1$gcd > 1),
@@ -53,7 +53,7 @@ neotoma_plan <- drake_plan(
       summarise(n = sum(n)),
     p_gcd1_high = n_gcd1_high/n_gcd1 * 100,
     
-    dataset_gcd1 = mean(pollen_summ2$gcd1 == 1) * 100,
+    dataset_gcd1 = mean(pollen_summ2$gcd1 != 1) * 100,
     dataset_gcd.5 = sum(pollen_summ2$gcd1 < threshold)
     
   ),
@@ -96,15 +96,17 @@ neotoma_plan <- drake_plan(
     #table of results
     table <- pollen_summ1 %>% 
       inner_join(datasets, by  = "datasetID") %>% 
-      group_by(dataset) %>% 
-      summarise(`No. assemblages` = n(), `Median no. taxa` = median(n_taxa), `GCD = 1` = mean(gcd == 1), `Median no. singletons (GCD = 1)` = median(n_singletons[gcd == 1])) 
+      rename(Dataset = dataset) %>% 
+      group_by(Dataset) %>% 
+      summarise(`No. assemblages` = n(), `Median no. taxa` = median(n_taxa), `% GCD = 1` = mean(gcd == 1) * 100, `Median no. singletons (GCD = 1)` = median(n_singletons[gcd == 1])) 
 
     #figure    
     figure <- pollen_summ1 %>% 
       inner_join(datasets, by = "datasetID") %>%
       filter(dataset %in% 1:2) %>% 
       mutate(gcd = factor(gcd)) %>% 
-      ggplot(aes(x = count_sum, y = n_taxa, colour = gcd, shape = gcd)) + 
+      ggplot(aes(x = count_sum, y = n_taxa, colour = gcd, shape = gcd)) +
[…]

### 2019-05-19 make_count_check_ms_with_drake.R
remove unused packages

@@ -3,9 +3,7 @@ library("drake")
 library("tidyverse")
 library("readr")
 library("readxl")
-#library("gridExtra")
-#library("broom")
-#devtools::install_github("richardjtelford/count_checker")#need to make package and repo names identical (_ not permitted)
+#devtools::install_github("richardjtelford/countSum")
 library("countSum")
 library("assertr")
 #devtools::install_github("richardjtelford/rjt.misc")

### 2019-05-19 Rmd/count_check_MS.Rmd
tweaks

@@
 This paper introduces methods that can flag assemblages with potentially misreported count sums and finds that some assemblage datasets fail these tests.
-journal: "Quaternary Science Reviews"
+journal: "Journal of Quaternary Science"
 date: "`r format(Sys.Date(), '%d %B %Y')`"
@@
 bibliography: extra/countMS2.bib
-csl: extra/journal-of-paleolimnology.csl
+csl: extra/elsevier-harvard_rjt.csl
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = TRUE, message = TRUE)
+knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = TRUE, message = TRUE, error = TRUE)
 
@@
 In total, `r diatom1_summ$n_low_n %>% as.english()` assemblages appear to have a count sum below 400 valves. 
+The results section of the paper acknowledges that counts were below the standard 400 valves for part of the Late Glacial.
+All the assemblages identified with possible low count sums are in this part of the stratigraphy, demonstrating the utility of the method.
 

### 2020-04-07 Rmd/count_check_MS.Rmd
more keywords and funding
reviewers comments

@@
 
-**Keywords:** Microfossil counts; quantitative methods
+**Keywords:** Microfossils assemblages, Counts; Quantitative methods, Reproducible research, Meta-research
 
@@
 
-In general, it is not possible to determine the sum from which percent were calculated, but the properties of community and assemblage counts make it possible to estimate it.
+In general, it is not possible to determine the sum from which percentages were calculated, but the properties of community and assemblage counts make it possible to estimate it.
 Given that we expect the rarest taxon to be a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
@@
 
-Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources.
+Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea, and other sources.
 A range of ecological and palaeoecological data were sought to allow for differences in typical count sums and species richness.
@@
 Datasets with percent data where the minimum count sum was not reported in the associated paper were excluded.
-Datasets discussed by Telford [-@Telford2019] as possibly having under-reported count sums were also excluded to avoid double reporting.
+Datasets discussed by Telford [-@Telford2019] as possibly having misreported the minimum count sum were also excluded to avoid double reporting.
 Datasets with possible misreporting are anonymised, but no attempt is made to diagnose whether errors or misconduct are responsible.
@@
 Some of the pollen datasets in the Neotoma database have an inexplicably high proportion of assemblages with a greatest common divisor greater than one. 
-Here I focus on four datasets produced by the same research group. 
+Here I focus on four datasets produced by one research group. 
 All the datasets, which are relatively recent, include information on the spike used to calculate pollen concentration, so cannot have been digitised.
@@
 The non-pollen datasets examined here are a convenience sample of available data that met the inclusion criteria.
-As such, this analysis cannot be used to accurately determine the prevalence of datasets with possible undercounts, but it appears that a non-negligible fraction of the literature is affected.
+As such, this analysis cannot be used to accurately determine the prevalence of datasets with possible undercounts, but it appears that a small but non-negligible fraction of the literature is affected.
 It is possible in some cases that the assemblages with low counts were omitted or merged prior to analysis, but the number of assemblages reported was not updated.
@@
 Some data were obtained from the Neotoma Paleoecology Database (http://www.neotomadb.org), and the work of the data contributors and the Neotoma community is gratefully acknowledged.
+This work was partly supported by Norwegian Research Council project PalaeoDrivers (213607).
 

### 2020-04-07 Rmd/count_check_MS.Rmd
as.english -> words()

@@
 Discounting the `r sum(owens_summ1$n_taxa == 1)` mono-specific assemblages (all with very low count sums`r max(owens_summ1$count_sum[owens_summ1$n_taxa == 1]) %>% insist(.
-<= 10) %>% invisible()`), `r owens_summ$p_singletons`% of assemblages have singletons with a median number of singletons per assemblage of `r owens_summ$median_sing %>% as.english()`.
+<= 10) %>% invisible()`), `r owens_summ$p_singletons`% of assemblages have singletons with a median number of singletons per assemblage of `r owens_summ$median_sing %>% words()`.
 Assemblages with count sums above 50 have a higher chance of having singletons (`r owens_summ$no_single_gt_50`% vs `r owens_summ$no_single50_`%).
-`r owens_summ$gcd2`% of assemblages have a greatest common divisor of one, the remainder have a greatest common divisor of `r owens_summ$gcd_max %>% as.english()`.
+`r owens_summ$gcd2`% of assemblages have a greatest common divisor of one, the remainder have a greatest common divisor of `r owens_summ$gcd_max %>% words()`.
 Assemblages with a greatest common divisor above one all have low taxonomic richness`r owens_summ1 %>% filter(gcd > 1) %>% pull(n_taxa) %>% max() %>% insist(. <= 6) %>% invisible()`.
@@
 `r testate_summ$p_singletons`% of the assemblages have at least one singleton (median `r testate_summ$median_singletons`) and `r testate_summ$p_gcd1`% of assemblages have a greatest common divisor of one.
-Of the `r testate_summ$n_gcd1 %>% as.english()` assemblages with a greatest common divisor above one, `r testate_summ$n_gcd2_low_div %>% as.english()` have fewer than five taxa; `r testate_summ$n_gcd2 %>% as.english()` have a greatest common divisor of two, and the other `r testate_summ$n_gcd3 %>% as.english()` has a greatest common divisor of three.
+Of the `r testate_summ$n_gcd1 %>% words()` assemblages with a greatest common divisor above one, `r testate_summ$n_gcd2_low_div %>% words()` have fewer than five taxa; `r testate_summ$n_gcd2 %>% words()` have a greatest common divisor of two, and the other `r testate_summ$n_gcd3 %>% words()` has a greatest common divisor of three.
 
@@
 Alternatively, the actual count sum for these assemblages could be as low as `r diatom1_summ$low_n` valves.
-In total, `r diatom1_summ$n_low_n %>% as.english()` assemblages appear to have a count sum below 400 valves. 
+In total, `r diatom1_summ$n_low_n %>% words()` assemblages appear to have a count sum below 400 valves. 
 The results section of the paper acknowledges that counts were below the standard 400 valves for part of the Late Glacial.
@@
 The associated paper reports that the diatom counts included 400–500 valves per assemblage except for four diatom-poor assemblages with at least 100 valves.
-The direct search method estimates that `r diatom2_summ$n400 %>% as.english()` assemblages have a count sum of less than 400, and `r diatom2_summ$n100 %>% as.english()` have less than 100.
+The direct search method estimates that `r diatom2_summ$n400 %>% words()` assemblages have a count sum of less than 400, and `r diatom2_summ$n100 %>% words()` have less than 100.
 The direct search and minimum percent methods agree, within rounding error, `r diatom2_summ$meth_agree %>% insist(isTRUE(.)) %>% invisible()`on the estimated count sums, implying that either all assemblages have singletons and count sums are low in some assemblages, or that the greatest common divisor of the raw counts is greater than one.
-This means that for the assemblage where the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc %>% as.english()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.
+This means that for the assemblage where the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc %>% words()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.
 
 Dataset chironomid1 has `r chironomid1_summ$nsamples` assemblages.
-Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 %>% as.english` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
+Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 %>% words` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
 These assemblages are diverse (richness between `r chironomid1_summ$richness %>% paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.
@@
 The associated paper reports that assemblages with count sums below 100 were omitted.
-The direct search method estimates that `r marine1_summ$n100 %>% as.english()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa %>% as.english()` mono-specific assemblage), with count sums as low as `r marine1_summ$est_n_min`.
+The direct search method estimates that `r marine1_summ$n100 %>% words()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa %>% words()` mono-specific assemblage), with count sums as low as `r marine1_summ$est_n_min`.
 With this dataset, the direct search and the minimum percent method give identical results for some assemblages but highly divergent estimates for others, with the direct search method sometimes giving estimates of several thousand, in one case over two orders of magnitude higher than the minimum percent method.

### 2020-04-07 make_manuscript.R
failed -> drake_failed

@@ -1,11 +1,10 @@
 library("drake")
-library("rjt.misc")
 
 plan <- "plan_count_check_ms_with_drake.R"
 
 #make ms
 r_make(source = plan) # Build the right things.
-failed()
+drake_failed()
 
 system("evince Rmd/count_check_MS.pdf", wait = FALSE)#display pdf - only linux
 

### 2020-04-07 plan_count_check_ms_with_drake.R
need english package

@@ -7,6 +7,7 @@ library("countSum")
 library("assertr")
 #remotes::install_github("richardjtelford/rjt.misc")
 library("rjt.misc")
+library("english")
 library("neotoma")
 #remotes::install_github("richardjtelford/neotoma2tibble")
 library("neotoma2tibble")

### 2020-05-13 R/download_diatoms.R
ungroup fix

@@ -33,6 +33,7 @@ diatom_plan <- drake_plan(
   owens_est_summ = owens_est %>% 
     select(est_n_minpc, est_n_direct, CountSum, n_taxa) %>% 
     filter(n_taxa > 1) %>% 
+    ungroup() %>% 
     summarise(
       minpc = mean(round(est_n_minpc, 3) == CountSum) * 100, 
       direct = mean(est_n_direct == CountSum) * 100
@@ -40,10 +41,15 @@ diatom_plan <- drake_plan(
   
   owens_fail = owens_est %>% 
     select(est_n_minpc, est_n_direct, CountSum, n_taxa) %>%
-    filter(n_taxa >1, est_n_direct!=CountSum),
+    filter(n_taxa > 1, est_n_direct != CountSum),
+  
+  owens_fail_minpc = owens_est %>% 
+    select(est_n_minpc, est_max_minpc, est_min_minpc, CountSum, n_taxa) %>%
+    filter(n_taxa > 1, !between(CountSum, est_min_minpc, est_max_minpc)),
   
   owens_plot = owens_est %>% 
     filter(between(n_taxa, 8, 15), between(CountSum, 80, 150)) %>% 
+    ungroup() %>% 
     slice(4) %>%  
     select(c(SampleId:n_taxa), score_best = score, est_n_direct) %>% 
     unnest(direct_search) %>% 

### 2020-05-13 R/download_neotoma.R
comment

@@ -94,6 +94,7 @@ neotoma_plan <- drake_plan(
       taxa_median = median(n_taxa)
     ),
   
+  ## unexpected gcd == 2
   pollen1234 = {
     datasets = neotoma_secrets %>% filter(dataset %in% 1:4) 
     #check has spike

### 2020-05-13 Rmd/count_check_MS.Rmd
edits to ms from reviewers

@@
 Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so.
-This paper introduces methods that can flag assemblages with potentially misreported count sums and finds that some assemblage datasets fail these tests.
+This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions.
+Several assemblage datasets that fail these tests, and potentially have misreported count sums, are identified.
 journal: "Journal of Quaternary Science"
@@
 This is important because larger count sums are associated with smaller uncertainties, both in relative abundance of taxa and derived statistics such as transfer function reconstructions [@Heiri2001; @Payne2009].
-However, mistakes happen and metadata such as count sums can be forgotten once percentages are calculated.
-In addition, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there may be an incentive to misreport the minimum count sum.
-The risk that count sums will be misreported is not just theoretical: Larocque-Tobler et al [-@Larocque-Tobler2015] reported that their chironomid count sums were at least fifty head capsules; a subsequent corrigendum [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen.
+There is a risk that count sums will be misreported.
+This could happen by mistake, especially with the use of secondary data.
+It could also happen knowingly, especially when microfossil preservation is poor or concentrations are low.
+This risk that count sums will be misreported is demonstrated by Larocque-Tobler et al [@Larocque-Tobler2016] who acknowledged that the chironomid count sums in [-@Larocque-Tobler2015] were not at least fifty head capsules as reported but as low as nineteen.
 Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.
@@
 All analyses were done in R version `r getRversion()` [@R] and used the packages extraDistr version `r packageVersion("extraDistr")` [@extraDistr], numbers version `r packageVersion("numbers")` [@numbers], and countSum `r packageVersion("countSum")` [@countSum].
-Code to replicate all the analyses shown above is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms).
+Code to replicate all the analyses is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms).
 
@@
 
-The vast majority of the over `r format(floor(bird_singleton_summary$n_routes[1]/1000) * 1000, big.mark = ",")` bird counts from the North American breeding bird survey [@NAmerican_Breeding_Bird2018]
-have singletons at the species level (Table \@ref(tab:bird-table))).
-To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the birds counts to progressively lower taxonomic resolutions.
-As richness declines, the proportion of counts having singletons declines (Table \@ref(tab:bird-table)), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
+The North American breeding bird survey is a very large, annual monitoring program to track the status of North American bird populations [@NAmerican_Breeding_Bird2018].
+All birds seen in point counts along ~40 km survey routes at the height of the breeding season are counted.
+The very large number of routes, over `r format(floor(bird_singleton_summary$n_routes[1]/1000) * 1000, big.mark = ",")`, typically with large numbers of individuals, allows the ideas developed above to be tested: do most routes have singletons and is the greatest common divisor one, and identify circumstances for which they fail.
+This can be tested for the observed data, and by simulating smaller counts and lower richness.
+
+The vast majority of the bird routes have singletons at the species level (Table \@ref(tab:bird-table))).
+To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the birds routes to progressively lower taxonomic resolutions.
+As richness declines, the proportion of routes having singletons declines (Table \@ref(tab:bird-table)), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
 The vast majority of counts have a greatest common divisor of one, except when taxonomic richness is low.
@@
 
-To test the sensitivity of the prevalence of counts without singletons to the count sum, I resample the counts with 400 or more observations to smaller count sums using a multivariate hypergeometric distribution.
+To test the sensitivity of the prevalence of routes without singletons to the count sum, I resample the routes with 400 or more observations to smaller count sums using a multivariate hypergeometric distribution.
 At the order level, the proportion of counts with singletons increases steeply until it reaches a maximum at about 200, thereafter is declines slowly (Fig. \@ref(fig:bird-recount)).
 The prevalence of counts with singletons is low for small counts because it is possible that only the common taxa are counted; with larger counts, the chance of counting a single individual of a rare taxon increases.
-With even larger counts the prevalence of counts with singletons decreases due to saturation as singletons become doubletons and few new taxa are found.
+With even larger counts the prevalence of routes with singletons decreases due to saturation as singletons become doubletons and few new taxa are found.
 
@@
 Some of the pollen datasets in the Neotoma database have an inexplicably high proportion of assemblages with a greatest common divisor greater than one. 
-Here I focus on four datasets produced by one research group. 
+Here I focus on four anonymised datasets produced by one research group. 
 All the datasets, which are relatively recent, include information on the spike used to calculate pollen concentration, so cannot have been digitised.

### 2020-05-13 response_to_reviewers.Rmd
.

# Response to reviewers

The referees have recommended publication after some minor, but significant, revisions to your manuscript.  Therefore, I invite you to respond to the referees' comments and revise your manuscript accordingly. Perhaps the most important point raised - notably by reviewer #1 - is that while your analysis is useful in helping identify errors and anomalies in data sets, this does not mean that falsification of data is commonplace in Quaternary palaeoecology, as you imply. Like you, I am a keen proponent of "open data" and, at my prompting, JQS has just moved from "authors are encouraged" to "authors are expected" to make their data openly available. Ensuring that authors do this is, to my mind, a far bigger problem than the few cases where data have been shown to be falsified.  It would therefore be helpful to frame your paper in this wider context of encouraging open data sharing, with data provided in their primary form (e.g. raw counts not %) to allow re-use.  Your paper is currently quite short and might benefit from more thorough explanations of the data sets used, etc.

We require a marked version of the revised manuscript. The marked version should indicate all changes that have been made in response to reviewer comments. This document can be created by highlighting areas of text or, by using the 'track changes' function in word.

##  Reviewer 1

__I appreciate the opportunity to review this paper, which describes useful methods for identifying potential errors in reported count sums in paleoecology. Overall, I found the paper to be clearly presented and accurate. It is deeply troubling to think that there are misreported data out there in my field, but I appreciate the author making tools available to identify those cases. It’s an important job and this paper should be of international interest. __

__I do take exception to two of the author’s conclusions (see the two starred items below) and believe he should reconsider them or more convincingly justify them. My other comments below are either very minor, or focused on clarifying a few take-home points and clarifying which datasets are referred to throughout the text. (An added table might help with that last issue.)__

__Lines 33-36: I don’t think the paper is strengthened with the statements that “mistakes happen and metadata such as count sums can be forgotten once percentages are calculated. In addition, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there may be an incentive to misreport the minimum count sum.” The former is vague and as a paleoecologist I can’t really image count sums “being forgotten” per se. I appreciate that the author is acknowledging that not every misreported number is intentional fraud; mistakes do happen! But this and the latter statement feel like speculation, and just not needed here. The author can simply state that this important data type has been demonstrably misreported in some past literature, per his next points with citations. __

This quote actually refers to a published paper. The paper reported that all counts were larger than 50 but some were as low as 13. 
Somewhere between the taxonomist who had counted the microfossils and the data analyst that instruction that some samples needed amalgamating got lost or forgotten. 
I have changes this to reflect the risk of secondary data analysis.

Without a survey of those who have mis-reported count sums, we can only speculate about their motive. 
I contend that poor preservation and low concentration are likely to increase the risk of misreported counts - I am assuming people are struggling with difficult samples.
I have edited this sentence to remove the phrase about incentives.

__Line 58: percentS (plural)?__

Fixed.

__Line 170: Suggest “four anonymized datasets” here to remind the reader that you are intentionally anonymizing datasets where you find irregularities.__

Fixed.

__And “the same research group” is ambigious (do you mean Axford from your previous example?). For clarity, I suggest “a single research group” or “one research group.”__

Fixed. “one research group”

__Line 183: “one pollen assemblage” of the four you analyzed as a set? I find the use of anonymized data rather hard to keep track of. I understand the choice to anonymize the “irregular” datasets here, and probably some awkwardness is an unavoidable outcome of anonymizing – but suggest another round of minor editing to make it as clear as possible throughout the paper which datasets are being referred to where. Suggest adding a TABLE listing all of the anonymized datasets referred to in the paper – with only the relevant properties of the datasets reported (dataset pollen3, proxy pollen, source Neotoma, # samples x, species richness x, etc). And then the text could refer to “dataset pollen3 (Table 3).” Most of this info is already in the text, but the author can make it easier for readers to keep track of the anonymized datasets.__

#TODO

__For Figs 1, 2 and 3 and Tables: For clarity and ease of use, I strongly suggest adding text that summarizes the take-home point(s) of each figure/table.  For example, rather than demonstrating problems with datasets, figs demonstrate major features of typical, ostensibly reliable datasets – and the utility of your methods. Make their take-homes clear in the captions.__

    #TODO

__**Line 251: “it appears 251 that a non-negligible fraction of the literature is affected.” Really? The global database of paleoecological data is HUGE.  What is meant by “non-negligible”?  This statement seems to condemn an entire field as rife with misconduct, when the results of this study of limited scope can NOT establish THAT.  To make this statement requires more evidence.  Instead, the author HAS demonstrated that there are probably examples of misreported count sums in public databases, and that the tools he is sharing can help in detecting those problems.__

<!-- (S4, S2, F1, Z2, C1, A1, T9, T1, L1, S1) -->
This comment hinges on the necessarily ill-defined nature of the term "non-negligible".

I am aware of 23 chironomid papers by five authors where the count sums are either smaller than the reported minimum reported or are smaller than 10 and the minimum is not reported. Several of these are widely cited. Assuming there are about 400 chironomid papers, this would represent about 5% of papers. I would contend that, for this subfield, this non-negligible.

I had hoped to be able to make quantitative estimates of the prevalence of under-counting from the Neotoma dataset. Unfortunately, for many Neotoma datasets, the metadata incorrectly report counts derived from stratigraphic diagrams as raw counts. Several strange datasets were found (perhaps 1%) but it was difficult to diagnose what had happened.

Since I submitted this manuscript, I have reviewed two manuscripts where count sums were below that reported.

I have changed the text to read "small but non-negligible".

__Lines 265-270: Does this finding contrast with reported, published count sums for these chironomid samples?  It’s unclear whether this is meant as an example of misconduct, or rather as a nice, simple example of data structure in low count sums. Clarify so as not to accidentally impugn these researchers if it’s the latter.__

TODO

__**Line 299: “the reader should be able to assume that the standard minimum count sum for the taxonomic group has been used.” I agree this would be great, but I disagree that this is true. Thus I strongly disagree that not meeting this expectation constitutes falsification. I can easily imagine a student new to the field making this mistake honestly; one would hope they archived their count sums so they are easy to reconstruct. Alternatively one could use the very tools presented here to sort it out. As a substitute, it would certainly be true (and alas also obvious, thus probably not worth stating) to say that “the reader should be able to assume that any reported minimum count sum is accurate and was strictly adhered to.”__

The cases I am thinking of here are papers written by people who are well aware of the typical minimum count sums for their taxon group.

    TODO

## Reviewer 2

__The paper deals with the count sum in microfossil-based palaeoecological research. In such research, the results (data) are normally presented as percentage data, and the author test some simple numerical tools to investigate whether the original count sum from which the percentages were calculated can be estimated. The conclusion is that the in some cases the real count sums are lower than reported in the papers, which will make any interpretations based on the data less reliable and useful.__

__Although the topic of the paper is no doubt important, there is a problem with the paper. The author has selected some datasets from the available large databases for analysis. In most cases, the papers from where the data are taken are not reported in the study. For example, in chapter 3.3 pollen data are presented with text “Here I focus on four datasets produced by the same research group”. No citations are given in text or in the caption of Table 2 where the data are analyzed. The same is true with diatom1 and diatom2 and marine1 datasets, which are analyzed but without citations. The author writes that “Datasets with possible misreporting are anonymised”. This is understandable, but clearly a problem from the point of view of normal scientific reporting, where the original sources are reported so that the readers can check and verify the authenticity of the data used.__

TODO

__Whether using anonymised data in the study can be accepted or not, is a wider question which I believe any journal needs to make. As said, I understand the author's decision, but I also see it as problematic.__

I understand the reviewer's concerns, and realise that using anonymised sources is an unusual decision. It is difficult to balance the need for the manuscript to be transparent with the need not to unfairly malign scientists for what may be honest errors. By anonymising sources, I am following the precident set by Brown and Heathers (2016; doi:10.1177/1948550616673876) and concur with their justification (in their conclusion).

As I wrote in my initial letter, I am prepared to share with the reviewers the information needed to decrypt the file containing the references so they can reproduce my analyses. I can also, equivalently give them the references directly.

__My other main issue is that I suggest that the author would make the “Methods” chapter clearer in the paper. Two techniques, called the minimum percentage method and direct search method, are used for estimating the count sums. It would be good to clearly present these methods in “Methods” instead of just mentioning them in the introduction, as is done in the current version.__

TODO

__Some more precise comments__

__-“We” and “I” are used in the text__

TODO

__-page 3. “Datasets…having under-reported count sums…” This should be “over-reported count sums”__

This sentence has been clarified. "having misreported the minimum count sum"

__-page 3. The bird example is hard to conceive. What does the “bird count” here mean? Does it mean the number of breeding bird species (two adults) in a certain area or a number of birds observed (so not necessarily breeding) in a certain area? Moreover, it is not totally clear why the analysis of the bird data was included in the study. One would assume that this type of diversity/evenness studies with bird data have been conducted earlier__

The methods developed in the manuscript are equally applicable to ecological and palaeoecological count data, so it is important to include data from both fields. The North American breeding birds is chosen because it is enormous, spans diverse ecological gradient, has a wide range of diversity at different taxonomic levels. This allows the sensitivity of the methods developed in the manuscript to be tested, for example, in relationship to richness. Richness is only discussed in so much as it is relevant to the presence of singletons, not as an ecological indicator in its own right.

The manuscript now describes this dataset better, and justifies its inclusion.

__-page 11 Telford, R.J. 2019a Is this a published or submitted paper? No journal is indicated__

Fixed - article now published

### 2020-06-10 make_manuscript.R
add instuctions on how to use renv

@@ -1,3 +1,7 @@
+#Use renv to install correct version of all necessary package
+#renv::restore()
+
+#load packages
 library("drake")
 
 plan <- "plan_count_check_ms_with_drake.R"

### 2020-06-10 Rmd/count_check_MS.Rmd
remove "we"

@@
 Deviations from the distribution of digits expected from the Newcomb-Benford law has been used to detect issues with scientific and financial data [@Barabesi2017].
-Carlisle [-@Carlisle2017] identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data.
-Brown et al. [-@Brown2017] developed the granularity-related inconsistency of means (GRIM) test which assesses whether means of integer data are consistent with the reported sample size.
+@Carlisle2017 identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data.
+@Brown2017 developed the granularity-related inconsistency of means (GRIM) test which assesses whether means of integer data are consistent with the reported sample size.
 However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data.
@@
 It could also happen knowingly, especially when microfossil preservation is poor or concentrations are low.
-This risk that count sums will be misreported is demonstrated by Larocque-Tobler et al [@Larocque-Tobler2016] who acknowledged that the chironomid count sums in [-@Larocque-Tobler2015] were not at least fifty head capsules as reported but as low as nineteen.
-Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.
+This risk that count sums will be misreported is demonstrated by @Larocque-Tobler2016 who acknowledged that the chironomid count sums in @Larocque-Tobler2015 were not at least fifty head capsules as reported but as low as nineteen.
+@Telford2019 reports some other cases where count sums may be much lower than reported.
 
@@
 In general, it is not possible to determine the sum from which percentages were calculated, but the properties of community and assemblage counts make it possible to estimate it.
-Given that we expect the rarest taxon to be a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
+Given the expectation that the rarest taxon is a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
 This method will fail for assemblages without singletons.
-We also expect the percent _p_ to be calculated from integer counts, therefore there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
+It is also expected that the percent _p_ is calculated from integer counts, therefore there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
 Possible values for _N_ can be found by a direct search algorithm over the range of plausible values of _N_.
@@
 Datasets with percent data where the minimum count sum was not reported in the associated paper were excluded.
-Datasets discussed by Telford [-@Telford2019] as possibly having misreported the minimum count sum were also excluded to avoid double reporting.
+Datasets discussed by @Telford2019 as possibly having misreported the minimum count sum were also excluded to avoid double reporting.
 Datasets with possible misreporting are anonymised, but no attempt is made to diagnose whether errors or misconduct are responsible.

### 2020-06-10 Rmd/count_check_MS.Rmd
more packages to cite, more methods

@@
 
+## Data
 Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea, and other sources.
@@
 
-All analyses were done in R version `r getRversion()` [@R] and used the packages extraDistr version `r packageVersion("extraDistr")` [@extraDistr], numbers version `r packageVersion("numbers")` [@numbers], and countSum `r packageVersion("countSum")` [@countSum].
+## Numerical methods
+
+For count data, the prevalence of counts without singletons and the 
+greatest common divisor of each count were calculated and related to the count sum.
+For percent data, the count sum was estimated using the minimum percent method and the direct search methods described above.
+
+All analyses were done in R v. `r getRversion()` [@R] and used the packages 
+extraDistr v. `r packageVersion("extraDistr")` [@extraDistr],
+numbers v. `r packageVersion("numbers")` [@numbers], 
+countSum v. `r packageVersion("countSum")` [@countSum], 
+neotoma v. `r packageVersion("neotoma")` [@neotoma], 
+drake v. `r packageVersion("drake")` [@drake], 
+and tidyverse v. `r packageVersion("tidyverse")` [@tidyverse].
 Code to replicate all the analyses is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms).
@@
 
-The North American breeding bird survey is a very large, annual monitoring program to track the status of North American bird populations [@NAmerican_Breeding_Bird2018].
+The North American breeding bird survey is a very large, annual monitoring programme to track the status of North American bird populations [@NAmerican_Breeding_Bird2018].
 All birds seen in point counts along ~40 km survey routes at the height of the breeding season are counted.

### 2020-06-10 plan_count_check_ms_with_drake.R

@@ -55,7 +55,7 @@ analyses <- drake_plan(
   
   #add extra packages to bibliography
   biblio2 = package_citations(
-    packages = c("extraDistr", "countSum", "numbers"), 
+    packages = c("extraDistr", "countSum", "numbers", "neotoma", "drake", "tidyverse"), 
     old_bib = file_in("Rmd/extra/countMS.bib"), 
     new_bib = file_out("Rmd/extra/countMS2.bib")),
   

### 2020-06-10 R/download_birds.R
reformats bird_recount_singleton_plot code

@@ -76,8 +76,12 @@ bird_plan <- drake_plan(
     summarise(m = mean(no_singletons > 0) * 100) %>% 
     summarise(m_no_singletons = mean(m), sd = sd(m), se = sd/sqrt(n())),
   
-  bird_recount_singleton_plot = bird_recount_singletons %>% 
-    ggplot(aes(x = new_count_sum, y = m_no_singletons, ymin = m_no_singletons - 1.96 * se, ymax = m_no_singletons + 2 * se)) + 
+  bird_recount_singleton_plot = ggplot(
+      data = bird_recount_singletons, 
+      aes(x = new_count_sum, y = m_no_singletons, 
+          ymin = m_no_singletons - 1.96 * se, 
+          ymax = m_no_singletons + 2 * se)
+    ) + 
     geom_errorbar() + 
     geom_point(size = 0.5) +
     labs(x = "Count sum", y = "Percent with singletons")

### 2020-06-10 R/download_chironomids.R
fix problem on unnest

@@ -11,11 +11,12 @@ chironomid_plan <- drake_plan(
   
   last_chance_direct_search_plot = last_chance_n %>% 
     filter(age_calBP == 1099) %>% 
+    select(age_calBP, totcaps, direct_search) %>% 
     unnest(direct_search) %>% 
+    filter(n <= 300) %>% 
     ggplot(aes(x  = n, y = score)) + 
     geom_point() +
-    labs(x = "Putative count sum", y = "Score") + 
-    xlim(NA, 300),
+    labs(x = "Putative count sum", y = "Score"),
   
   last_chance_plot = last_chance_n %>% 
     unnest(direct_search_est) %>%   

### 2020-06-11 Rmd/count_check_MS.Rmd
add hershop bog

@@
 loadd(last_chance_summ)
-loadd(pollen_summ, pollenA, pollen1234)
+loadd(pollen_summ, hershop_pollen_summ, pollenA, pollen1234)
 loadd(owens_summ1, owens_summ, owens_est_summ)
@@
 Some datasets with a high proportion of assemblages with a greatest common divisor above one are older datasets which may, despite the available metadata, have been digitised with the loss of rare taxa and precision.
-For example, one dataset includes `r pollenA$n_samp` assemblages with a median richness of `r pollenA$taxa_median` taxa.
-Count sums vary between `r pollenA$count_min` and `r pollenA$count_max`.
-The greatest common divisor is three for all assemblages, that is all `r pollenA$n_count` counts in the dataset are divisible by three.
+For example, all but `r hershop_pollen_summ$has_singleton %>% word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two).
+Discrepancies between the archived data and the original paper [@Larson_1972] including an incomplete species list and the inclusion of fern spores in the count sum of two hundred grains strongly suggest that these data have been digitised.
+For most purposes, such digitised data are perfectly adequate.
+
 Excluding such datasets would further increase the proportion of assemblages with singletons. 
-Some other Neotoma datasets are discussed is a subsequent section.
+Some other Neotoma datasets are discussed in section \@ref(unexpected-count-data) of this paper.
 
@@
 Some of the pollen datasets in the Neotoma database have an inexplicably high proportion of assemblages with a greatest common divisor greater than one. 
-Here I focus on four anonymised datasets produced by one research group. 
+Here I focus on five anonymised datasets.
+
+Dataset pollen1, published includes `r pollenA$n_samp` assemblages with a median richness of `r pollenA$taxa_median` taxa.
+Count sums vary between `r pollenA$count_min` and `r pollenA$count_max`.
+The greatest common divisor is three for all assemblages, that is all `r pollenA$n_count` counts in the dataset are divisible by three.
+
 All the datasets, which are relatively recent, include information on the spike used to calculate pollen concentration, so cannot have been digitised.

### 2020-06-11 R/download_neotoma.R

@@ -1,6 +1,7 @@
 
 #source functions
 source("R/get_sites_meta.R")
+source("R/hershop_pollen_plan.R")
 
 #### drake plan ####
 neotoma_plan <- drake_plan(
@@ -68,7 +69,8 @@ neotoma_plan <- drake_plan(
 
   #ecological groups
   pollen_wanted = c("TRSH", "UPHE", "SUCC", "PALM", "MANG"),
-
+  
+  
   ##bad examples####
   #presumed digitised dataset
   pollenA_data = if(has_password){
@@ -136,4 +138,11 @@ neotoma_plan <- drake_plan(
     list(table = table, figure = figure, summ = summ)
   }
 )#end of drake plan
+
+# bind pollen plans
+
+pollen_plan <- bind_plans(
+  neotoma_plan, 
+  hershop_plan
+) 
 

### 2020-06-11 R/hershop_pollen_plan.R

@@ -1,0 +1,74 @@
+#Hershop bog, neotoma datasetID 3132
+
+hershop_plan <- drake_plan(
+  hershop_pollen = counts(pollen_data[["3132"]]) %>% 
+    rownames_to_column(var = "sampleID") %>% 
+    pivot_longer(-sampleID, names_to = "taxa", values_to = "count") %>%
+    filter(count > 0),
+    
+  hershop_pollen_summ1 = hershop_pollen %>% 
+    group_by(sampleID) %>% 
+    summarise(
+      pollen_spore_count = sum(count), 
+      pollen_count = sum(count[taxa != "Unknown (trilete)"]), 
+      gcd = numbers::mGCD(count)), 
+  
+  hershop_pollen_summ = hershop_pollen_summ1 %>%
+    ungroup() %>% 
+    summarise(
+    n = n(),
+    has_singleton = sum(gcd == 1), 
+    pollen_gt_200 = mean(pollen_count >= 200),
+    pollen_spore_gt_200 = mean(pollen_spore_count >= 200),
+    pollen_spore_200 = mean(pollen_spore_count == 200)
+    )
+)#end plan
+
+# hershop_pollen %>% distinct(taxa) %>% arrange(taxa) %>% print(n = Inf)
+#  
+# hershop_pollen %>% 
+#   group_by(sampleID) %>% 
+#   mutate(percent = count/sum(count) * 100) %>% 
+#   group_by(taxa) %>% 
+#   filter(n() > 1, max(count) > 4) %>% 
+#   ggplot(aes(x = -as.numeric(sampleID), y = percent)) +
+#   geom_area() +
+#   geom_point(alpha = 0.5, colour = "red") + 
+#   coord_flip() +
+#   facet_wrap(~taxa, nrow = 1) +
+#   theme(strip.text = element_text(angle = 90))
+
+# taxa                         
+# <chr>                        
+# 1 Acer                  #paper explicitly states not present       
+# 2 Alnus                   10     
+# 3 Amaranthaceae         22       
+# 4 Apiaceae              20       
+# 5 Asteraceae            18       
+# 6 Betula                2       
+# 7 Carya                 5       
+# 8 Celtis                8       
+# 9 Cyperaceae            19       
+# 10 Ericaceae            25        
+# 11 Fraxinus             13        
+# 12 Ilex                 15        
+# 13 Juglans            7          
+# 14 Juniperus           3         
+# 15 Liquidambar            16      
[…]

### 2020-06-11 plan_count_check_ms_with_drake.R

@@ -109,7 +109,7 @@ plans <- bind_rows(
   bird_plan, 
   testate_plan, 
   chironomid_plan, 
-  neotoma_plan,
+  pollen_plan,
   diatom_plan,
   marine_plan,
   analyses)

### 2020-06-11 Rmd/count_check_MS.Rmd
Lake A (x3) moved and expanded

@@
 Count sums vary between `r pollenA$count_min` and `r pollenA$count_max`.
-The greatest common divisor is three for all assemblages, that is all `r pollenA$n_count` counts in the dataset are divisible by three.
+The greatest common divisor is three for all assemblages: all `r pollenA$n_count` counts in the dataset are divisible by three. 
+Assuming (optimistically) the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $1/k^n$.
+For a dataset this large, the probability of having all counts divisible by three by chance is 10^`r ceiling(- pollenA$n_count * log10(3))`^.
+Given this very small probability, it seems reasonable to conclude that the archived data are not the original counts, but I have no idea how they were generated.
 

### 2020-06-11 Rmd/count_check_MS.Rmd
edit hershop bog with Patty thesis

@@
 The assemblages with a greatest common divisor above one are not randomly distributed among the datasets: only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
-Some datasets with a high proportion of assemblages with a greatest common divisor above one are older datasets which may, despite the available metadata, have been digitised with the loss of rare taxa and precision.
-For example, all but `r hershop_pollen_summ$has_singleton %>% word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two).
-Discrepancies between the archived data and the original paper [@Larson_1972] including an incomplete species list and the inclusion of fern spores in the count sum of two hundred grains strongly suggest that these data have been digitised.
-For most purposes, such digitised data are perfectly adequate.
+Some datasets with a high proportion of assemblages with a greatest common divisor above one are older datasets which may, despite the available metadata, have been back-transformed from percent data or digitised with the loss of rare taxa and precision.
+For example, all but `r hershop_pollen_summ$has_singleton %>% word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two).
+The original data [@Patty1968] were counted to exactly 200 pollen grains and fern spores. 
+It seems likely that the data were converted to percent, rounded and then back-tranformed to counts.
+For most purposes, such back-transformed or digitised data are perfectly adequate.
 

### 2020-06-11 R/hershop_pollen_plan.R
remove comparison with Larson et al. Patty 1968 is source

@@ -23,52 +23,3 @@ hershop_plan <- drake_plan(
     pollen_spore_200 = mean(pollen_spore_count == 200)
     )
 )#end plan
-
-# hershop_pollen %>% distinct(taxa) %>% arrange(taxa) %>% print(n = Inf)
-#  
-# hershop_pollen %>% 
-#   group_by(sampleID) %>% 
-#   mutate(percent = count/sum(count) * 100) %>% 
-#   group_by(taxa) %>% 
-#   filter(n() > 1, max(count) > 4) %>% 
-#   ggplot(aes(x = -as.numeric(sampleID), y = percent)) +
-#   geom_area() +
-#   geom_point(alpha = 0.5, colour = "red") + 
-#   coord_flip() +
-#   facet_wrap(~taxa, nrow = 1) +
-#   theme(strip.text = element_text(angle = 90))
-
-# taxa                         
-# <chr>                        
-# 1 Acer                  #paper explicitly states not present       
-# 2 Alnus                   10     
-# 3 Amaranthaceae         22       
-# 4 Apiaceae              20       
-# 5 Asteraceae            18       
-# 6 Betula                2       
-# 7 Carya                 5       
-# 8 Celtis                8       
-# 9 Cyperaceae            19       
-# 10 Ericaceae            25        
-# 11 Fraxinus             13        
-# 12 Ilex                 15        
-# 13 Juglans            7          
-# 14 Juniperus           3         
-# 15 Liquidambar            16      
-# 16 Myrica               12        
-# 17 Onagraceae             26      
-# 18 Pinus                 4       
-# 19 Poaceae              17        
-# 20 Polygonaceae         24        
-# 21 Quercus            1          
-# 22 Salix                 9       
-# 23 Spermatophyta undiff. (trees)
-# 24 Typha                  21      
-# 25 Ulmus                 6       
-# 26 Unknown (trilete)  
-
-# #missing
-# #populus 11
-# #Rhus 14
-# #Vitis 27
-# #Nyctaginaceae 23

### 2020-06-11 Rmd/count_check_MS.Rmd
pollen2 pollen3 text

@@
 The greatest common divisor is three for all assemblages: all `r pollenA$n_count` counts in the dataset are divisible by three. 
-Assuming (optimistically) the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $1/k^n$.
+Assuming (optimistically) the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $k^{-n}$.
 For a dataset this large, the probability of having all counts divisible by three by chance is 10^`r ceiling(- pollenA$n_count * log10(3))`^.
-Given this very small probability, it seems reasonable to conclude that the archived data are not the original counts, but I have no idea how they were generated.
+Given this very small probability, it seems reasonable to conclude that the archived data are not the original counts.
 
-All the datasets, which are relatively recent, include information on the spike used to calculate pollen concentration, so cannot have been digitised.
-The four datasets have some assemblages with a greatest common divisor of two interspersed, sometimes regularly, amongst assemblages with the expected greatest common divisor of one.
-The assemblages with a greatest common divisor of one typically have several singletons (Table \@ref(tab:neotoma-table)).
+In datasets pollen2 and pollen3, presented in the same paper, # % and # % of assemblages do not have singletons and have a greatest common divisor greater than one.
+Both datasets match the published pollen stratigraphies.
+For most of both datasets, assemblages with and without singletons are arranged alternately.
+Samples without singletons are on the 10 cm intervals, and those with on the intervening 5 cm intervals. 
+Closer inspection reveals that the assemblages between two assemblages without singletons have counts that are the mean of their neighbours.
+On request, the authors sent the original count data. 
+Comparison of these with the archived data showed that the archived assemblages without singletons had had their counts multiplied by two.
+The intervening assemblages did not exist in the original data. 
+In the published and archived data, they had been interpolated from their neighbours (Fig. \@ref(fig:archived-original-plot-plot)).
+
+fig - plot of 1 or 2 mid-common taxa by depth in archived and original
 
@@
 
-Assemblages with a greatest common divisor of one typically have a higher taxonomic richness than assemblages with a greatest common divisor of two (Fig. \@ref(fig:neotoma-plot)).
-In the datasets with a wide range of count sums, the richness of the assemblages with a greatest common divisor of two is comparable with the richness of the assemblages with a greatest common divisor of one that have a count sum that is half as large.
-
-```{r neotoma-plot, fig.height = 70/24.5, fig.width=80/24.5, fig.cap="Number of taxa in against count sum for assemblages in two pollen datasets with a greatest common divisor of one (red circles) or two (blue triangles).
+```{r archived-original-plot, fig.height = 70/24.5, fig.width=80/24.5, fig.cap="Number of taxa in against count sum for assemblages in two pollen datasets with a greatest common divisor of one (red circles) or two (blue triangles).
 Lines are loess with a span of 1.
 All four pollen datasets show the same pattern, but it is clearest in datasets 1 and 2 which have the highest proportion of assemblages with a greatest common divisor of two.", message = FALSE}
-pollen1234$figure %>% print()
+readd(archived_original_plot)
 ```

### 2020-06-12 Rmd/count_check_MS.Rmd
datasets pollen23 plots and tables

@@
 
-In datasets pollen2 and pollen3, presented in the same paper, # % and # % of assemblages do not have singletons and have a greatest common divisor greater than one.
+In datasets pollen2 and pollen3, presented in the same paper, # % and # % of assemblages do not have singletons and have a greatest common divisor greater than one (Table \@ref(tab:unexpected-table))).
 Both datasets match the published pollen stratigraphies.
@@
 The intervening assemblages did not exist in the original data. 
-In the published and archived data, they had been interpolated from their neighbours (Fig. \@ref(fig:archived-original-plot-plot)).
+In the published and archived data, they had been interpolated from their neighbours (Fig. \@ref(fig:interpolated-plot)).
 
-fig - plot of 1 or 2 mid-common taxa by depth in archived and original
-
-```{r neotoma-table}
-pollen1234$table %>% knitr::kable(caption = "The number of assemblages, median number of taxa, percent of assemblages with a greatest common divisor (GCD) of one, and the median number of singletons in assemblages with a greatest common divisor of one, in four pollen datasets")
+```{r unexpected-table}
+readd(unexpected_table) %>% knitr::kable(caption = "The number of assemblages, median number of taxa, percent of assemblages with a greatest common divisor (GCD) of one, and the median number of singletons in assemblages with a greatest common divisor of one, in four pollen datasets")
 ```
 
-```{r archived-original-plot, fig.height = 70/24.5, fig.width=80/24.5, fig.cap="Number of taxa in against count sum for assemblages in two pollen datasets with a greatest common divisor of one (red circles) or two (blue triangles).
-Lines are loess with a span of 1.
-All four pollen datasets show the same pattern, but it is clearest in datasets 1 and 2 which have the highest proportion of assemblages with a greatest common divisor of two.", message = FALSE}
-readd(archived_original_plot)
+```{r interpolated-plot, warning = FALSE, fig.height = 100/24.5, fig.width=180/24.5, fig.cap="A) Plot of selected taxa by depth from dataset pollen2.
+Counts with blue triangles are from assemblages with a greatest common divisor of one.
+These appear to have been linearly interpolated between the adjacent counts from assemblages with a greatest common divisor of two (red circles).
+B) Percent of taxa in each assemblage where the count is the mean of the counts in the adjacent assemblages, excluding runs of zero counts.
+Most of the assemblages with a greatest common divisor of one are exactly linearly interpolated between adjacent assemblages.
+The main exception is the assemblage at 210 cm which has a single singleton and otherwise has a greatest common divisor of two, other exceptions have a few extra singletons not predicted from the adjacent assembages.", message = FALSE}
+readd(pollen23_plot)
 ```
 
-One assemblage with a greatest common divisor of two has `r pollen1234$summ$max_richness` taxa. 
-Assuming the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $1/k^n$: for a single assemblage with this many taxa, the probability is 1 in 10^`r floor(log10(2^pollen1234$summ$max_richness))`^.
+One assemblage with a greatest common divisor of two has `r "pollen1234$summ$max_richness"` taxa. 
+Assuming the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $1/k^n$: for a single assemblage with this many taxa, the probability is 1 in 10^`r "floor(log10(2^pollen1234$summ$max_richness))"`^.
 The probability that these counts reflect the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled at some stage in some way.

### 2020-06-12 R/unexpected_pollen_plan.R

@@ -1,0 +1,119 @@
+
+unexpected_pollen_plan <- drake_plan(
+  ##bad examples####
+  #multiple 3
+  pollenA_data = if(has_password){
+    datasetID <- neotoma_secrets %>% 
+      filter(dataset == 1) %>% 
+      select(-problem)
+    pollen %>% 
+      inner_join(datasetID) %>% 
+      mutate(datasetID = dataset) %>% 
+      select(-dataset)
+  } else {
+    read_csv("data_backup/pollenA_data.csv") %>% 
+      group_by(sampleID)
+  },
+  
+  pollenA_summ = pollenA_data %>% 
+    pollen_summarise() %>% 
+    ungroup() %>% 
+    assert(in_set(3), gcd),# check multiples of three
+  
+  pollenA = pollenA_summ %>% 
+    summarise(
+      n_samp = n(),
+      n_count = sum(n_taxa),
+      count_min = min(count_sum),
+      count_max = max(count_sum),
+      taxa_median = median(n_taxa)
+    ),
+  
+  
+  ## interpolated
+  pollen23_data = if(has_password){
+      datasetID <- neotoma_secrets %>% filter(dataset %in% 2:3) %>% 
+        select(-problem)
+      
+      depths <- pollen_data[as.character(pull(datasetID, datasetID))] %>% 
+        map_df("sample.meta") %>% 
+        select(depth, sample.id)
+      
+      pollen %>% 
+        ungroup() %>% 
+        inner_join(datasetID) %>% 
+        left_join(depths, by = c("sampleID" = "sample.id")) %>% 
+        mutate(datasetID = dataset) %>% 
+        select(-dataset) %>% 
+        mutate(taxa = paste("Taxon", as.numeric(factor(taxa)))) # anonymise taxa
+    } else {
+      read_csv("data_backup/pollen23_data.csv")
+    },
+  
+  #### dataset 2/3 plot ####
+  pollen23_plot = {
+    plot_data <- pollen23_data %>% 
+      filter(datasetID == 2) %>% 
+      ungroup() %>% 
+      select(datasetID, depth, taxa, count) %>%
+      complete(taxa, nesting(datasetID, depth), fill = list(count = 0)) %>%
+      group_by(datasetID, depth) %>% 
+      mutate(gcd = factor(numbers::mGCD(count))) 
+
+    count_depth <- plot_data %>%
+      filter(taxa %in% c("Taxon 41",
+                         "Taxon 38",
[…]

### 2020-06-12 plan_count_check_ms_with_drake.R

@@ -1,6 +1,7 @@
-#import package
+#import packages
 library("drake")
 library("tidyverse")
+library("patchwork")
 library("readxl")
 #remotes::install_github("richardjtelford/countSum")
 library("countSum")

### 2020-06-12 Rmd/count_check_MS.Rmd
improve pollen2/3

@@
 
-In datasets pollen2 and pollen3, presented in the same paper, # % and # % of assemblages do not have singletons and have a greatest common divisor greater than one (Table \@ref(tab:unexpected-table))).
-Both datasets match the published pollen stratigraphies.
-For most of both datasets, assemblages with and without singletons are arranged alternately.
+Many assemblages in datasets pollen2 and pollen3 lack singletons and have a greatest common divisor greater than one (Table \@ref(tab:unexpected-table))).
+For most of both datasets, which come from the same paper and match the published pollen stratigraphies, assemblages with and without singletons are arranged alternately.
 Samples without singletons are on the 10 cm intervals, and those with on the intervening 5 cm intervals. 
@@
 In the published and archived data, they had been interpolated from their neighbours (Fig. \@ref(fig:interpolated-plot)).
+The effect of these manipulations is to increase the resolution of the pollen stratigraphy.
 

### 2020-06-12 R/download_neotoma.R

@@ -2,6 +2,7 @@
 #source functions
 source("R/get_sites_meta.R")
 source("R/hershop_pollen_plan.R")
+source("R/unexpected_pollen_plan.R")
 
 #### drake plan ####
 neotoma_plan <- drake_plan(
@@ -70,79 +71,13 @@ neotoma_plan <- drake_plan(
   #ecological groups
   pollen_wanted = c("TRSH", "UPHE", "SUCC", "PALM", "MANG"),
   
-  
-  ##bad examples####
-  #presumed digitised dataset
-  pollenA_data = if(has_password){
-    datasetID <- neotoma_secrets %>% filter(dataset == "A") 
-    pollen %>% 
-      semi_join(datasetID)
-  } else {
-    read_csv("data_backup/pollenA_data.csv") %>% 
-      group_by(sampleID)
-  },
-  
-  pollenA_summ = pollenA_data %>% 
-    pollen_summarise() %>% 
-    ungroup() %>% 
-    assert(in_set(3), gcd),# check multiples of three
-    
-  pollenA = pollenA_summ %>% 
-    summarise(
-      n_samp = n(),
-      n_count = sum(n_taxa),
-      count_min = min(count_sum),
-      count_max = max(count_sum),
-      taxa_median = median(n_taxa)
-    ),
-  
-  ## unexpected gcd == 2
-  pollen1234 = {
-    datasets = neotoma_secrets %>% filter(dataset %in% 1:4) 
-    #check has spike
-    datasets %>% pull(datasetID) %>% map(process_rare_data, pollen_data = pollen_data)
-    
-    #table of results
-    table <- pollen_summ1 %>% 
-      inner_join(datasets, by  = "datasetID") %>% 
-      rename(Dataset = dataset) %>% 
-      group_by(Dataset) %>% 
-      summarise(`No. assemblages` = n(), 
-                `Median no. taxa` = median(n_taxa), 
-                `% GCD = 1` = mean(gcd == 1) * 100, 
-                `Median no. singletons (GCD = 1)` = median(n_singletons[gcd == 1])) 
-
-    #figure    
-    figure <- pollen_summ1 %>% 
-      inner_join(datasets, by = "datasetID") %>%
-      filter(dataset %in% 1:2) %>% 
-      mutate(gcd = factor(gcd)) %>% 
-      ggplot(aes(x = count_sum, y = n_taxa, colour = gcd, shape = gcd)) +
[…]

### 2020-06-12 R/unexpected_pollen_plan.R

@@ -114,6 +114,8 @@ unexpected_pollen_plan <- drake_plan(
         `Median richness` = median(richness), 
         `Median countsum` = median(countsum),
         `Percent GCD > 1` = mean(gcd > 1) * 100
-                )
+                ) %>% 
+    mutate(datasetID = paste0("pollen", datasetID)) %>% 
+    rename(Dataset = datasetID)
 )#end of plan
 

### 2020-06-14 Rmd/count_check_MS.Rmd
unify unexpected pollen code

@@
 
-Dataset pollen1, published includes `r pollenA$n_samp` assemblages with a median richness of `r pollenA$taxa_median` taxa.
-Count sums vary between `r pollenA$count_min` and `r pollenA$count_max`.
-The greatest common divisor is three for all assemblages: all `r pollenA$n_count` counts in the dataset are divisible by three. 
+One assemblage with a greatest common divisor of two has `r "pollen1234$summ$max_richness"` taxa. 
+Assuming the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $1/k^n$: for a single assemblage with this many taxa, the probability is 1 in 10^`r "floor(log10(2^pollen1234$summ$max_richness))"`^.
+The probability that these counts reflect the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled at some stage in some way.
+
+Dataset pollen3, published includes `r pollen3$n_samp` assemblages with a median richness of `r pollen3$taxa_median` taxa.
+Count sums vary between `r pollen3$count_min` and `r pollen3$count_max`.
+The greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three. 
 Assuming (optimistically) the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $k^{-n}$.
-For a dataset this large, the probability of having all counts divisible by three by chance is 10^`r ceiling(- pollenA$n_count * log10(3))`^.
+For a dataset this large, the probability of having all counts divisible by three by chance is 10^`r ceiling(- pollen3$n_count * log10(3))`^.
 Given this very small probability, it seems reasonable to conclude that the archived data are not the original counts.
 
-Many assemblages in datasets pollen2 and pollen3 lack singletons and have a greatest common divisor greater than one (Table \@ref(tab:unexpected-table))).
+Many assemblages in datasets pollen4 and pollen5 lack singletons and have a greatest common divisor greater than one (Table \@ref(tab:unexpected-table))).
 For most of both datasets, which come from the same paper and match the published pollen stratigraphies, assemblages with and without singletons are arranged alternately.
@@
 The main exception is the assemblage at 210 cm which has a single singleton and otherwise has a greatest common divisor of two, other exceptions have a few extra singletons not predicted from the adjacent assembages.", message = FALSE}
-readd(pollen23_plot)
+readd(pollen4_plot)
 ```
-
-One assemblage with a greatest common divisor of two has `r "pollen1234$summ$max_richness"` taxa. 
-Assuming the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $1/k^n$: for a single assemblage with this many taxa, the probability is 1 in 10^`r "floor(log10(2^pollen1234$summ$max_richness))"`^.
-The probability that these counts reflect the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled at some stage in some way.
 

### 2020-06-14 R/unexpected_pollen_plan.R

@@ -1,26 +1,34 @@
 
 unexpected_pollen_plan <- drake_plan(
   ##bad examples####
-  #multiple 3
-  pollenA_data = if(has_password){
+  pollen_unexpected_data = if(has_password){
     datasetID <- neotoma_secrets %>% 
-      filter(dataset == 1) %>% 
       select(-problem)
+    
+    depths <- pollen_data[as.character(pull(datasetID, datasetID))] %>% 
+      map_df("sample.meta") %>% 
+      select(depth, sample.id)
+    
     pollen %>% 
-      inner_join(datasetID) %>% 
+      ungroup() %>% 
+      inner_join(datasetID) %>%
+      left_join(depths, by = c("sampleID" = "sample.id")) %>% 
       mutate(datasetID = dataset) %>% 
-      select(-dataset)
+      select(-dataset) %>% 
+      mutate(taxa = paste("Taxon", as.numeric(factor(taxa)))) # anonymise taxa
   } else {
-    read_csv("data_backup/pollenA_data.csv") %>% 
+    read_csv("data_backup/pollen_unexpected_data.csv") %>% 
       group_by(sampleID)
   },
   
-  pollenA_summ = pollenA_data %>% 
+  pollen3_summ = pollen_unexpected_data %>% 
+    filter(datasetID == 3) %>% 
+    group_by(sampleID) %>% 
     pollen_summarise() %>% 
     ungroup() %>% 
     assert(in_set(3), gcd),# check multiples of three
   
-  pollenA = pollenA_summ %>% 
+  pollen3 = pollen3_summ %>% 
     summarise(
       n_samp = n(),
       n_count = sum(n_taxa),
@@ -31,29 +39,11 @@ unexpected_pollen_plan <- drake_plan(
   
   
   ## interpolated
-  pollen23_data = if(has_password){
-      datasetID <- neotoma_secrets %>% filter(dataset %in% 2:3) %>% 
-        select(-problem)
-      
-      depths <- pollen_data[as.character(pull(datasetID, datasetID))] %>% 
-        map_df("sample.meta") %>% 
-        select(depth, sample.id)
-      
-      pollen %>% 
-        ungroup() %>% 
-        inner_join(datasetID) %>% 
-        left_join(depths, by = c("sampleID" = "sample.id")) %>% 
-        mutate(datasetID = dataset) %>% 
-        select(-dataset) %>% 
[…]

### 2020-06-14 plan_count_check_ms_with_drake.R

@@ -92,14 +92,11 @@ analyses <- drake_plan(
         mutate(taxon = fct_anon(factor(taxon), prefix = "diatom_")) %>%
         write_csv(path = "data_backup/diatom2.csv")
       
-      #pollen A. (multiple 3)
-      pollenA_data %>% 
+      #unexpected pollen 
+      pollen_unexpected_data %>% 
         ungroup() %>% 
-        mutate(
-          datasetID = "pollen1",
-          sampleID = fct_anon(factor(sampleID), prefix = "sample_"),
-          taxa = fct_anon(factor(taxa), prefix = "pollen_")) %>%
-        write_csv("data_backup/pollenA_data.csv")
+        mutate(sampleID = fct_anon(factor(sampleID), prefix = "sample_")) %>%
+        write_csv("data_backup/pollen_unexpected_data.csv")
       
       }
     }

### 2020-06-15 Rmd/count_check_MS.Rmd
wildhorse

@@
 loadd(last_chance_summ)
-loadd(pollen_summ, hershop_pollen_summ, pollenA, pollen1234)
+loadd(pollen_summ, hershop_pollen_summ, wildhorse, pollen_summ1, pollen_summ2, pollen3, pollen1234)
 loadd(owens_summ1, owens_summ, owens_est_summ)
@@
 The pollen data from the Neotoma database included over `r pollen_summ$n_assemblages` assemblages in `r pollen_summ$n_datasets %>% format()` datasets. 
-Nearly all assemblages have singletons (`r pollen_summ$p_singletons`%) and a greatest common divisor of one (`r pollen_summ$p_gcd1`%). 
-The assemblages with a greatest common divisor above one are not randomly distributed among the datasets: only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
-Some datasets with a high proportion of assemblages with a greatest common divisor above one are older datasets which may, despite the available metadata, have been back-transformed from percent data or digitised with the loss of rare taxa and precision.
-For example, all but `r hershop_pollen_summ$has_singleton %>% word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two).
-The original data [@Patty1968] were counted to exactly 200 pollen grains and fern spores. 
-It seems likely that the data were converted to percent, rounded and then back-tranformed to counts.
-For most purposes, such back-transformed or digitised data are perfectly adequate.
-
-Excluding such datasets would further increase the proportion of assemblages with singletons. 
-Some other Neotoma datasets are discussed in section \@ref(unexpected-count-data) of this paper.
+Conforming to expectations, nearly all assemblages have singletons (`r pollen_summ$p_singletons`%) and a greatest common divisor of one (`r pollen_summ$p_gcd1`%). 
+The assemblages without singletons or with a greatest common divisor above one are not randomly distributed among the datasets.
+Half of the assemblages without singletons are in `r pollen_summ$half_no_sing` datasets.
+`r Words(pollen_summ$half_gcd)` datasets contain half of the `r pollen_summ$n_gcd1` assemblages with a greatest common divisor above one and only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
+Some Neotoma datasets are with a high proportion of assemblages without singletons or a with a greatest common divisor above one are discussed in section \@ref(unexpected-count-data) of this paper.
 
@@
 
-Some of the pollen datasets in the Neotoma database have an inexplicably high proportion of assemblages with a greatest common divisor greater than one. 
-Here I focus on five anonymised datasets.
+This section explores some of the pollen datasets in the Neotoma database that have a high proportion of assemblages without singletons or with a greatest common divisor greater than one.
+
+Some of these datasets are older datasets which may, despite the available metadata, have been back-transformed from percent data or digitised with the loss of rare taxa and precision.
+For example, all but `r hershop_pollen_summ$has_singleton %>% word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two).
+The original data [@Patty1968] were counted to exactly 200 pollen grains and fern spores. 
+It seems likely that the data were converted to percent, rounded and then back-transformed to counts.
+For most purposes, such back-transformed or digitised data are perfectly adequate.
+
+Some datasets with assemblages without singletons have few taxa.
+For example, the `r wildhorse$n_assemblage` assemblage dataset from Wildhorse Lake [@ref] includes only `r wildhorse$n_taxa_dataset` taxa (`r mean(pollen_summ2$n_taxa_dataset > wildhorse$n_taxa_dataset) * 100` % of neotoma pollen datasets have more than this), and relatively large count sums (median `r wildhorse$median_count`).
+While `r wildhorse$no_singletons * 100`% of assemblages in this dataset lack singletons, even more assemblage (`r wildhorse$no_zeros * 100`%) lack taxa with zero counts. 
+This is in contrast to most (`r pollen_summ$has_zeros * 100`%) neotoma pollen assemblages which have zero counts.  
+These patterns could arise either when there is a high sampling effort in a low diversity site, or when the archived data do not include the entire pollen assemblage, instead focusing on the subset of abundant taxa of most interest to the authors.
+
+For some other datasets, the reason that assemblages lack singletons or have a greatest common divisor above one is less clear. 
+These are treated anonymously.
 
@@
 
-Dataset pollen3, published includes `r pollen3$n_samp` assemblages with a median richness of `r pollen3$taxa_median` taxa.
-Count sums vary between `r pollen3$count_min` and `r pollen3$count_max`.
-The greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three. 
+In dataset pollen3, the greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three (Table \@ref(tab:unexpected-table))). 
 Assuming (optimistically) the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $k^{-n}$.

### 2020-06-15 R/download_neotoma.R

@@ -34,27 +34,37 @@ neotoma_plan <- drake_plan(
     filter(sum(count) >= 50, n() > 1) %>%
     #round slightly to correct near-integers backtransformed? from percent
     mutate(count = round(count, 3)) %>% 
-    #filter out two datasets with probable percent data
-    filter(!datasetID %in% c(15059, 25609)) %>% 
-    mutate(count = ceiling(count)), 
+    #filter out four datasets with probable percent data
+    filter(!datasetID %in% c(15059, 25609, 16209, 16210)) %>% 
+    mutate(count = ceiling(count)) %>% 
+    #remove unnecessary columns
+    select(datasetID, sampleID, taxa, count) %>% 
+    # sometimes spikes have not been correctly identified
+    filter(!(datasetID == 252 & taxa == "Alnus")) %>% 
+    #remove duplicate dataset == 21903
+    filter(datasetID != 47543), 
+  
 
 
   #process
   pollen_summ = tibble(
-    n_datasets = n_distinct(pollen$datasetID),
-    n_assemblages = (floor(n_distinct(pollen$sampleID)/1000) * 1000) %>% format(big.mark = ","),
+    n_datasets = nrow(pollen_summ2),
+    n_assemblages = (floor(nrow(pollen_summ1)/1000) * 1000) %>% format(big.mark = ","),
     p_singletons = mean(pollen_summ1$n_singletons > 0) * 100,
     p_gcd1 = mean(pollen_summ1$gcd == 1) * 100,
     n_gcd1 = sum(pollen_summ1$gcd > 1),
-    threshold = 0.5, 
-    n_gcd1_high = pollen_summ1 %>% 
-      summarise(dgcd = mean(gcd == 1),  n = n()) %>% 
-      filter(dgcd < threshold) %>% 
-      summarise(n = sum(n)),
-    p_gcd1_high = n_gcd1_high/n_gcd1 * 100,
+    dataset_gcd1 = mean(pollen_summ2$gcd1 < 1) * 100
     
-    dataset_gcd1 = mean(pollen_summ2$gcd1 != 1) * 100,
-    dataset_gcd.5 = sum(pollen_summ2$gcd1 < threshold)
+  ) %>% bind_cols(#number of datasets needed to get half gcd > 1 assemblages
+    pollen_summ1 %>% 
+      group_by(datasetID) %>% 
+      summarise(gcd = sum(gcd > 1), no_singletons = sum(min > 1), threshold = 0.5) %>% 
[…]

### 2020-06-15 R/get_sites_meta.R

@@ -42,8 +42,14 @@ process_rare_data <- function(ID, pollen_data){#browser()
 
 #dplyr function to summarise pollen counts
 pollen_summarise <- . %>% 
+  group_by(datasetID) %>% 
+  mutate(n_taxa_dataset = n_distinct(taxa)) %>% 
+  group_by(datasetID, sampleID) %>% 
   summarise(count_sum = sum(count),
             n_taxa = n(), 
             gcd = numbers::mGCD(count), 
             n_singletons = sum(count == 1), 
-            min = min(count))
+            min = min(count),
+            n_taxa_dataset = first(n_taxa_dataset),
+            has_zero = n_taxa_dataset > n_taxa
+            )

### 2020-06-15 R/hershop_pollen_plan.R

@@ -21,5 +21,23 @@ hershop_plan <- drake_plan(
     pollen_gt_200 = mean(pollen_count >= 200),
     pollen_spore_gt_200 = mean(pollen_spore_count >= 200),
     pollen_spore_200 = mean(pollen_spore_count == 200)
-    )
+    ),
+  
+  
+#### Wildhorse Lake ####  
+  wildhorse = pollen %>% filter(datasetID == 15894) %>% 
+  ungroup() %>% 
+  mutate(n_taxa_dataset = n_distinct(taxa)) %>% 
+  group_by(datasetID, sampleID) %>% 
+  summarise( #summarise samples
+    count_sum = sum(count), 
+    n_singletons = sum(count == 1), 
+    n_taxa = n(), 
+    n_taxa_dataset = first(n_taxa_dataset)) %>% 
+  summarise(#summarise dataset
+    n_assemblage = n(), 
+    median_count = median(count_sum), 
+    no_singletons = mean(n_singletons == 0), 
+    no_zeros = mean(n_taxa != n_taxa_dataset), 
+    n_taxa_dataset = first(n_taxa_dataset))
 )#end plan

### 2020-06-16 Rmd/count_check_MS.Rmd
new pollen2

@@
 
+Dataset pollen2 (Fig. \@ref(fig:multiples-plot))
+
+```{r multiples-plot, warning = FALSE, fig.height = 100/24.5, fig.width=180/24.5, fig.cap = "A) Distribution of the percent of counts that are divisible by _k_ in each pollen assemblage with at least ten taxa.
+The vertical dashed line represents the 99.5% percentile.
+B) Percent of count in each assemblage in dataset pollen2 that are divisible by _k_ against depth.
+The dashed lines are the quantiles shown in part A.
+", message = FALSE}
+multiples_plot
+```
+
 In dataset pollen3, the greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three (Table \@ref(tab:unexpected-table))). 

### 2020-06-16 R/unexpected_pollen_plan.R

@@ -20,7 +20,65 @@ unexpected_pollen_plan <- drake_plan(
     read_csv("data_backup/pollen_unexpected_data.csv") %>% 
       group_by(sampleID)
   },
+
+  #### multiples_plot ####
+  multiples_plot = {
+    multiples = pollen_unexpected_data %>%
+      group_by(depth) %>% 
+      filter(datasetID == 2) %>% 
+      summarise(n = n(), 
+                div_2 = mean(count %% 2 == 0),
+                div_3 = mean(count %% 3 == 0)) %>% 
+      pivot_longer(cols = starts_with("div"), names_to = "divisible", values_to = "prop") %>% 
+      mutate(
+        percent = prop * 100,
+        divisible = str_replace(string = divisible, pattern = ".*(\\d)", replacement = "italic(k)==\\1"))
+    
+    # null 
+    multiples_null <- pollen %>% 
+      filter(n() >= 10) %>% #remove low diversity assemb
+      summarise(
+        div_2 = mean(count %% 2 == 0),
+        div_3 = mean(count %% 3 == 0)
+      ) %>% 
+      pivot_longer(cols = starts_with("div"), names_to = "divisible", values_to = "prop")  %>% 
+      mutate(
+        percent = prop * 100,
+        divisible = str_replace(string = divisible, pattern = ".*(\\d)", replacement = "italic(k)==\\1"))
+    
+    multiples_99 <- multiples_null  %>% 
+      group_by(divisible) %>% 
+      summarise(q99 = quantile(percent, prob = 0.995)) 
+    
+    
+    div_null = ggplot(multiples_null, aes(x = percent, fill = divisible)) +
+      geom_histogram(boundary = 0) +
+      geom_vline(data = multiples_99, mapping = aes(xintercept = q99, colour = divisible), linetype = "dashed") +
+      scale_fill_brewer(palette = "Set1") +
+      scale_colour_brewer(palette = "Set1") +
+      scale_x_continuous(expand = c(0.02, 0)) +
+      facet_wrap(~divisible, ncol = 1, labeller = label_parsed) + 
+      labs(x = expression('%'~counts~divisible~by~italic(k)), y = "Frequency") +
+      theme(legend.position = "none")
+    
+    div_data = multiples %>% 
+      ggplot(aes(x = depth, y = percent, colour = divisible)) + 
+      geom_line() +
+      geom_point(size = 1) +
[…]

### 2020-06-16 Rmd/count_check_MS.Rmd
pollen1, pollen2

@@
 loadd(last_chance_summ)
-loadd(pollen_summ, hershop_pollen_summ, wildhorse, pollen_summ1, pollen_summ2, pollen3, pollen1234)
+loadd(pollen_summ, hershop_pollen_summ, wildhorse, pollen_summ1, pollen_summ2, pollen3, pollen1, pollen2)
 loadd(owens_summ1, owens_summ, owens_est_summ)
@@
 Some datasets with assemblages without singletons have few taxa.
-For example, the `r wildhorse$n_assemblage` assemblage dataset from Wildhorse Lake [@ref] includes only `r wildhorse$n_taxa_dataset` taxa (`r mean(pollen_summ2$n_taxa_dataset > wildhorse$n_taxa_dataset) * 100` % of neotoma pollen datasets have more than this), and relatively large count sums (median `r wildhorse$median_count`).
+For example, the `r wildhorse$n_assemblage` assemblage dataset from Wildhorse Lake [@Mehringer1985] includes only `r wildhorse$n_taxa_dataset` taxa (`r mean(pollen_summ2$n_taxa_dataset > wildhorse$n_taxa_dataset) * 100` % of Neotoma pollen datasets have more than this), and relatively large count sums (median `r wildhorse$median_count`).
 While `r wildhorse$no_singletons * 100`% of assemblages in this dataset lack singletons, even more assemblage (`r wildhorse$no_zeros * 100`%) lack taxa with zero counts. 
-This is in contrast to most (`r pollen_summ$has_zeros * 100`%) neotoma pollen assemblages which have zero counts.  
+This is in contrast to most (`r pollen_summ$has_zeros * 100`%) Neotoma pollen assemblages which have zero counts.  
 These patterns could arise either when there is a high sampling effort in a low diversity site, or when the archived data do not include the entire pollen assemblage, instead focusing on the subset of abundant taxa of most interest to the authors.
@@
 
-One assemblage with a greatest common divisor of two has `r "pollen1234$summ$max_richness"` taxa. 
-Assuming the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $1/k^n$: for a single assemblage with this many taxa, the probability is 1 in 10^`r "floor(log10(2^pollen1234$summ$max_richness))"`^.
-The probability that these counts reflect the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled at some stage in some way.
+`r Words(pollen1$n_gcd2)` of the `r pollen1$n_sample` assemblages in dataset pollen1 lack singletons and have a greatest common divisor of two.
+These assemblages have between `r pollen1$min_taxa` and `r pollen1$max_taxa` taxa.
+The probability of all _n_ counts being divisible by _k_ is $k^{-n}$, assuming the distribution of counts is locally uniform.
+That is, for _k_ = 2, odd counts are as common as even counts.
+For an assemblage with `r pollen1$min_taxa` or `r pollen1$max_taxa` taxa, the probability of having all counts divisible by two by chance are 10^`r ceiling(- pollen1$min_taxa * log10(2))`^ and 10^`r ceiling(- pollen1$max_taxa * log10(2))`^ respectively.
+These probabilities are not so small that it is unrealistic to expect to find assemblages like these in the Neotoma database, but to find so many in one dataset is unexpected.
+Furthermore, the assumption made here is optimistic, in reality for _k_ = 2, more odd counts are expected ones are expected than even ones in most datasets, as typical rank abundance curves have more singletons than doubletons and so on, which makes finding assemblages with a greatest common divisor above one even more unlikely.
 
-Dataset pollen2 (Fig. \@ref(fig:multiples-plot))
+Dataset pollen2 has `r pollen2$n_sample` assemblages, of which `r words(pollen2$n_gcd2)` have a greatest common divisor of two and a further `r words(pollen2$n_gcd3)` with a greatest common divisor of three, one of which has `r pollen2$max_taxa3` taxa.
+The probability of having all counts divisible by three in such a taxonomically rich assemblage is 10^`r ceiling(- pollen2$max_taxa3 * log10(3))`.
+In addition to the assemblages with a greatest common divisor above one, this dataset includes several assemblages that have a higher proportion of counts divisible by two or three than is typical in assemblages from Neotoma (Fig.
+\@ref(fig:multiples-plot)).
+It is unclear what processes could generate such counts.
 
@@
 ", message = FALSE}
-multiples_plot
+readd(multiples_plot)
 ```
@@
 In dataset pollen3, the greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three (Table \@ref(tab:unexpected-table))). 
-Assuming (optimistically) the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $k^{-n}$.
-For a dataset this large, the probability of having all counts divisible by three by chance is 10^`r ceiling(- pollen3$n_count * log10(3))`^.
+For a dataset this large, the probability of having all counts divisible by three by chance is 3^{-`r pollen3$n_count`}^ or 10^`r ceiling(- pollen3$n_count * log10(3))`^.
 Given this very small probability, it seems reasonable to conclude that the archived data are not the original counts.

### 2020-06-16 R/unexpected_pollen_plan.R

@@ -20,8 +20,37 @@ unexpected_pollen_plan <- drake_plan(
     read_csv("data_backup/pollen_unexpected_data.csv") %>% 
       group_by(sampleID)
   },
-
-  #### multiples_plot ####
+  
+  pollen_unexpected_summ = pollen_unexpected_data %>% 
+    group_by(datasetID, sampleID) %>% 
+    pollen_summarise() %>% 
+    ungroup(),
+  
+  ####pollen1 ####
+  pollen1 = pollen_unexpected_summ %>% 
+      filter(datasetID == 1) %>% 
+      ungroup() %>% 
+      summarise(
+        min_taxa = min(n_taxa[gcd == 2]),
+        max_taxa = max(n_taxa[gcd == 2]),
+        n_sample = n(),
+        n_gcd2 = sum(gcd == 2)
+    ),
+  
+  #### multiples_plot - pollen 2####
+  pollen2 = pollen_unexpected_summ %>% 
+    filter(datasetID == 2) %>% 
+    ungroup() %>% 
+    summarise(
+      n_sample = n(),
+      min_taxa2 = min(n_taxa[gcd == 2]),
+      max_taxa2 = max(n_taxa[gcd == 2]),
+      min_taxa3 = min(n_taxa[gcd == 3]),
+      max_taxa3 = max(n_taxa[gcd == 3]),
+      n_gcd2 = sum(gcd == 2),
+      n_gcd3 = sum(gcd == 3),
+    ),
+  
   multiples_plot = {
     multiples = pollen_unexpected_data %>%
       group_by(depth) %>% 
@@ -79,21 +108,21 @@ unexpected_pollen_plan <- drake_plan(
   
   
   #### pollen 3 multiples ####  
-  pollen3_summ = pollen_unexpected_data %>% 
-    filter(datasetID == 3) %>% 
-    group_by(sampleID) %>% 
-    pollen_summarise() %>% 
-    ungroup() %>% 
-    assert(in_set(3), gcd),# check multiples of three
   
-  pollen3 = pollen3_summ %>% 
+  pollen3 = {
+    pollen3_summ <- pollen_unexpected_summ %>% 
+      filter(datasetID == 3) %>% 
+      assert(in_set(3), gcd)# check multiples of three
+    
+    pollen3_summ %>% 
     summarise(
       n_samp = n(),
       n_count = sum(n_taxa),
       count_min = min(count_sum),
       count_max = max(count_sum),
[…]

### 2020-06-16 Rmd/count_check_MS.Rmd
details

@@
 
-The vast majority of the bird routes have singletons at the species level (Table \@ref(tab:bird-table))).
+The vast majority of the bird routes have singletons at the species level (Table \@ref(tab:bird-table)).
 To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the birds routes to progressively lower taxonomic resolutions.
@@
   select(`Taxonomic level` = taxonomic_level, `Percent with singletons` = p_singletons, `Percent GCD = 1` = GCD1, `Median richness` = median_richness) %>%
-  knitr::kable(digits = c(0, 6, 6, 3), caption = "Percent of bird counts with singletons and of greatest common divisor (GCD) of one, at different taxonomic levels.")
+  knitr::kable(digits = c(0, 6, 6, 3), caption = "Percent of bird counts from the North American breeding bird survey with singletons and of greatest common divisor (GCD) of one, at different taxonomic levels to mimic the effect of taxonomic richness.")
 options(digits = 2)
@@
 
-```{r bird-recount, fig.cap="Effect of count sum on the proportion of bird counts without singleton at the order level.
+```{r bird-recount, fig.cap="Effect of count sum on the proportion of bird counts from the North American breeding bird survey without singleton at the order level.
 Results are the mean of ten trials, error bars are two standard errors.", fig.width = 80/25.4, fig.height = 70/25.4}
@@
 
-In dataset pollen3, the greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three (Table \@ref(tab:unexpected-table))). 
+In dataset pollen3, the greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three (Table \@ref(tab:unexpected-table)). 
 For a dataset this large, the probability of having all counts divisible by three by chance is 3^{-`r pollen3$n_count`}^ or 10^`r ceiling(- pollen3$n_count * log10(3))`^.
@@
 
-Many assemblages in datasets pollen4 and pollen5 lack singletons and have a greatest common divisor greater than one (Table \@ref(tab:unexpected-table))).
+Many assemblages in datasets pollen4 and pollen5 lack singletons and have a greatest common divisor greater than one (Table \@ref(tab:unexpected-table)).
 For most of both datasets, which come from the same paper and match the published pollen stratigraphies, assemblages with and without singletons are arranged alternately.
-Samples without singletons are on the 10 cm intervals, and those with on the intervening 5 cm intervals. 
+Assemblages without singletons are on the 10 cm intervals, and those with on the intervening 5 cm intervals. 
 Closer inspection reveals that the assemblages between two assemblages without singletons have counts that are the mean of their neighbours.
@@
 ```{r unexpected-table}
-readd(unexpected_table) %>% knitr::kable(caption = "The number of assemblages, median number of taxa, percent of assemblages with a greatest common divisor (GCD) of one, and the median number of singletons in assemblages with a greatest common divisor of one, in four pollen datasets")
+readd(unexpected_table) %>% knitr::kable(caption = "The number of assemblages, median number of taxa and count sum, percent of assemblages with a greatest common divisor (GCD) of one, in five anonymised pollen datasets.")
 ```
@@
 
-The archived data include `r diatom1_summ$n_taxa` taxa that have a maximum abundance of at least 1.25%; this is about half of the 109 taxa reported in the paper.
+The archived data for diatom1 include `r diatom1_summ$n_taxa` taxa that have a maximum abundance of at least 1.25%; this is about half of the 109 taxa reported in the paper.
 This pruning of rare taxa will increase the risk of counts without singletons, which would cause the minimum percent method to fail, but will have little impact on the direct search method.
@@
 These tests require the assemblage data which are often not archived. 
-However, if the count sums are very low it is possible to recognise this from a stratigraphic diagram. 
-For example, a chironomid stratigraphy had an assemblage where all five taxa had a relative abundance of about 20%, suggesting a count sum of five, far below the reported fifty chironomids.
-Another assemblage had six taxa with a relative abundance of just over 10% and a seventh with twice as much, suggesting a count sum of eight chironomids.
-On request, the authors sent the data. 
-None of the `r chironomid2_summ$n_samp` assemblages in the dataset had more that fifty chironomids: the count sums varied between `r chironomid2_summ$count_min` and  `r chironomid2_summ$count_max`.
+However, if the count sums are very low (below about 20) it is possible to recognise this from a stratigraphic diagram.
 
@@
 Substantial undercounts will potentially seriously effect the precision of the results.
-Such undercounts might constitute a data handling error, for example, if samples with low counts were supposed to be merged but that step was forgotten.
+Such undercounts might constitute a data handling error, for example, if assemblages with low counts were supposed to be merged or omitted but that step was forgotten.
 Substantial and pervasive under-counting could be construed as scientific misconduct due to either negligence or falsification, and action to correct the literature is probably required.
@@
 50 for chironomids, several hundred for pollen and diatoms).
-If the actual count sums are materially below this, then this potentially constitutes falsification by omission [@Fanelli2013].
+The methods in this paper may help estimate what the count sums are.
+If the actual count sums are materially below this, then action to correct the literature may be necessary, and if the omission is deliberate it 
+potentially constitutes misconduct [@Fanelli2013].
 
@@
 
-A non-negligible fraction of the archived assemblage data includes counts that appear to have a count sum below that reported in the paper. 
+A small but non-negligible fraction of archived assemblage data includes counts that appear, using methods developed above, to have a count sum below that reported in the original paper. 
 Some of the apparent counts are small enough that the uncertainty of the count and derived statistics will be substantially larger than expected.
-These results highlight the importance of archiving both the raw data and the code required to process the data.
+These results highlight the importance of archiving both the raw microfossil count data and the code required to process the data.
 

### 2020-06-16 R/unexpected_pollen_plan.R

@@ -189,7 +189,7 @@ unexpected_pollen_plan <- drake_plan(
       summarise(
         `No. assemblages` = n(),
         `Median richness` = median(richness), 
-        `Median countsum` = median(countsum),
+        `Median count sum` = median(countsum),
         `Percent GCD > 1` = mean(gcd > 1) * 100
                 ) %>% 
     mutate(datasetID = paste0("pollen", datasetID)) %>% 

### 2020-06-16 plan_count_check_ms_with_drake.R
change to released v of dplyr

@@ -10,8 +10,6 @@ library("assertr")
 library("rjt.misc")
 library("english")
 library("neotoma")
-#remotes::install_github("richardjtelford/neotoma2tibble")
-library("neotoma2tibble")
 
 #extra packages required
 requireNamespace("visNetwork")

### 2020-08-19 R/download_testates.R
download only Amesbury data

@@ -5,30 +5,22 @@ testate_plan <- drake_plan(
   testate_datasets = get_dataset(datasettype = "testate amoebae surface sample"),
 
   #download_data
-  testate_data = testate_datasets %>% get_download(), #SLOW
+  testate_data = testate_datasets %>%
+    #zap two datasets with null submission date
+    discard(~is.null(.x$submission$submission.date)) %>% 
+    #keep only data from before 2019-05-01 - Amesbury's data
+    keep(~{lubridate::ymd(.x$submission$submission.date) < as.Date("2019-05-01")}) %>% 
+    map(get_download), #SLOW
   
-  #extract site information
-  testate_meta = testate_datasets %>% 
-    map("site.data") %>% 
-    map_df(as_tibble) %>% 
-    bind_cols(testate_datasets %>% 
-                map_df("dataset.meta")) %>% 
-    mutate(authors = testate_datasets %>% 
-             map("pi.data") %>% 
-             map("ContactName") %>% 
-             map_chr(paste, collapse = " "), 
-           submission = testate_datasets %>% 
-             map("submission") %>% 
-             map_chr("submission.date") %>% 
-             lubridate::ymd()),
-
   #some sites have only percent data
   testate_has_percent = testate_data %>%
+    map(1) %>% 
     map("taxon.list") %>%
     map_lgl(~any(.x$variable.units == "percent", na.rm = TRUE)),  
   
   #get counts
-  testate_counts = testate_data[!testate_has_percent & testate_meta$submission < lubridate::ymd("2019-05-01")] %>%#cut data without counts or recently added
+  testate_counts = testate_data[!testate_has_percent] %>%#cut data without counts
+    map(1) %>% #each element is a download_list - extract contents
     map(~({
       cnt <- counts(.x)
       tl <- .x$taxon.list %>% filter(variable.units == "NISP")
@@ -68,4 +60,4 @@ testate_plan <- drake_plan(
        n_gcd3 = sum(gcd == 3),
        n_gcd2_low_div = sum(gcd[n_taxa < 5] > 1)
      )
-)
+)#end of plan

### 2020-08-19 R/renv_extra_packages.R
list of dependencies not identified by renv

@@ -1,0 +1,4 @@
+#extra packages required but recognised by renv (probably suggests rather than imports field of package description file)
+requireNamespace("visNetwork")
+requireNamespace("rticles")
+requireNamespace("curl")

### 2020-08-19 plan_count_check_ms_with_drake.R
remove futures - not supported in Rstudio

@@ -10,16 +10,10 @@ library("assertr")
 library("rjt.misc")
 library("english")
 library("neotoma")
-
-#extra packages required
-requireNamespace("visNetwork")
-requireNamespace("rticles")
+library("neotoma2tibble")
 
 #drake configuration
 pkgconfig::set_config("drake::strings_in_dots" = "literals")
-#set up parallel processing for drake
-options(future.fork.enable = TRUE)
-future::plan(future::multiprocess) 
 
 ## askpass
 options(askpass = function(x)readLines("pw.txt"))
@@ -31,6 +25,7 @@ source("R/download_chironomids.R")
 source("R/download_neotoma.R")
 source("R/download_diatoms.R")
 source("R/download_marine.R")
+source("R/download_neotoma_diatoms.R")
 
 
 #construct drake plan
@@ -54,7 +49,7 @@ analyses <- drake_plan(
   
   #add extra packages to bibliography
   biblio2 = package_citations(
-    packages = c("extraDistr", "countSum", "numbers", "neotoma", "drake", "tidyverse"), 
+    packages = c("extraDistr", "countSum", "numbers", "neotoma", "drake", "tidyverse", "bookdown", "renv"), 
     old_bib = file_in("Rmd/extra/countMS.bib"), 
     new_bib = file_out("Rmd/extra/countMS2.bib")),
   
@@ -108,12 +103,13 @@ plans <- bind_rows(
   pollen_plan,
   diatom_plan,
   marine_plan,
+  neotoma_diatoms_plan,
   analyses)
 
 #quick network plot
 plot(plans)
 
 #configure and make drake plan
-config <- drake_config(plans, jobs = 3, parallelism = "future", keep_going = TRUE)
+config <- drake_config(plans, keep_going = TRUE)
 
 config

### 2020-08-19 Rmd/count_check_MS.Rmd
improved abstract - methods

@@
   Microfossil counts are a key data type in palaeoecology.
-Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so.
+Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so, and perhaps other aspects of the counts.
 This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions.
-Several assemblage datasets that fail these tests, and potentially have misreported count sums, are identified.
+Several assemblage datasets that fail these tests, and potentially have misreported counts or count sums, are identified.
 journal: "Journal of Quaternary Science"
@@
 
-If the genuine assemblage counts are archived for all taxa, it is trivial to identify undercounts.
-Unfortunately, count data might be falsified so that it appears to meet the reported count sum.
-A more common problem is that, regrettably, many palaeoecologists archive percent data without an indication of the count sum.
-This paper develops simple tests that can flag if the count data might have been misrepresented, or if the count sum of percent data is perhaps lower than reported.
+Given that count sums of percent data have been misreported, the possibility that other aspects of count data may be misreported needs to be considered.
+This paper aim to develop and test simple methods that can flag if the count data might have been misreported, or if the count sum of percent data is perhaps lower than reported.
 
-The key insight that allows inference about the count sum is that assemblage data are expected to follow the typical features of a rank abundance curve.
-In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual, hereafter singletons, unless the sampling effort is high relative to the taxonomic richness [@Coddington2009].
+The key insight that allows inference about community or assemblage count data is that they represent a species abundance distribution.
+The ubiquitous shape of species abundance distributions when plotted as as the number of species against abundance is a hyperbolic or hollow curve with many rare and few abundant species [@McGill_2007].
+This relationship is central to ecological theory. 
+Communities or assemblages data that differ markedly this expected relationship
+
+For example, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual, hereafter singletons, unless the sampling effort is high relative to the taxonomic richness [@Coddington2009].
+
 In the occasional assemblages without singletons, the greatest common divisor should usually be one, i.e.
@@
 
-In general, it is not possible to determine the sum from which percentages were calculated, but the properties of community and assemblage counts make it possible to estimate it.
-Given the expectation that the rarest taxon is a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
-This method will fail for assemblages without singletons.
-It is also expected that the percent _p_ is calculated from integer counts, therefore there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
-Possible values for _N_ can be found by a direct search algorithm over the range of plausible values of _N_.
-An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
-These tests are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.
+Some complications and caveats are discussed.
 
-This paper aims to test the methods presented above and present some cases with unexpected results.
-Some complications and caveats are discussed.
+<!-- If the genuine assemblage counts are archived for all taxa, it is trivial to identify undercounts. -->
+<!-- Unfortunately, count data might be falsified so that it appears to meet the reported count sum. -->
+<!-- A more common problem is that, regrettably, many palaeoecologists archive percent data without an indication of the count sum. -->
 
@@
 greatest common divisor of each count were calculated and related to the count sum.
-For percent data, the count sum was estimated using the minimum percent method and the direct search methods described above.
+
+Generalised additive model proportion of explained deviance
+Proportion of taxa divisible by _k_ where $k > 1$
+
+In general, it is not possible to determine the sum from which percentages were calculated, but the properties of community and assemblage counts make it possible to estimate it.
+Given the expectation that the rarest taxon is a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
+This method will fail for assemblages without singletons.
+It is also expected that the percent _p_ is calculated from integer counts, therefore there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
+Possible values for _N_ can be found by a direct search algorithm over the range of plausible values of _N_.
+An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
+These tests are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.
 
@@
 neotoma v. `r packageVersion("neotoma")` [@neotoma], 
-drake v. `r packageVersion("drake")` [@drake], 
 and tidyverse v. `r packageVersion("tidyverse")` [@tidyverse].
+Packages drake v. `r packageVersion("drake")` [@drake], 
+bookdown `r packageVersion("bookdown")` [@bookdown] and
+renv v. `r packageVersion("renv")` [@renv] were used to ensure reproducibility.
 Code to replicate all the analyses is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms).

### 2021-06-06 R/download_birds.R
use downloaded species list - AOU match

@@ -5,19 +5,19 @@ bird_plan <- drake_plan(
   #download data
   bird_download = {
     bbs <- get_bbs_data(bbs_version = 2020, sb_dir = "data/birds2020")
-    #remove unneeded components
-    bbs$weather <- NULL
-    bbs$routes <- NULL
+
     #select required columns
-    bbs$species_list <- bbs$species_list %>% 
-      select(order, family, genus, species, AOU)
-    bbs$observations <- bbs$observations %>% 
+    bbs$observations %>% 
       select(RouteDataID, AOU, SpeciesTotal)
-    bbs
     },
   
+  bird_species = read_table("data/birds2020/SpeciesList.txt", skip = 9) %>% 
+    slice(-1) %>% # row of dashes
+    select(-Seq, -French_Common_Name, -English_Common_Name, -Species) %>% 
+    rename(Species = Spanish_Common_Name, Order = ORDER), 
+  
   #aggregate to different taxonomic levels
-  bird_singletons = bird_process(bird_download$observations, bird_download$species_list),
+  bird_singletons = bird_process(bird_download, bird_species),
 
   bird_singleton_summary = bird_singletons %>%
     group_by(taxonomic_level) %>%
@@ -28,7 +28,8 @@ bird_plan <- drake_plan(
       n_routes = n(),
       GCD1 = mean(GCD == 1, na.rm = TRUE) * 100,
       GCD2plus = mean(GCD > 1, na.rm = TRUE) * 100,
-      GCD_max = max(GCD, na.rm = TRUE)
+      GCD_max = max(GCD, na.rm = TRUE), 
+      .groups = "drop_last"
     ) %>%
     mutate(taxonomic_level = factor(
       taxonomic_level,
@@ -36,20 +37,26 @@ bird_plan <- drake_plan(
     )) %>%
     arrange(taxonomic_level),
   
+  bird_summ = list(
+    n_without_singletons = bird_singletons %>% filter(taxonomic_level == "species", singletons == 0) %>% nrow(),
+    n_GCD_gt1 = bird_singletons %>% filter(taxonomic_level == "species", GCD > 1 | is.na(GCD))
+    
+  ),
+  
   #simulate counts with mv hypergeometric
-  bird_recount_singletons = bird_download$observations %>% 
+  bird_recount_singletons = bird_download %>% 
     group_by(RouteDataID) %>% 
[…]

### 2021-06-06 R/bird_functions.R
makes code more memory efficient

@@ -14,20 +14,24 @@ bird_process <- function(bird_data, bird_species) {
   joined <- bird_data %>%
     left_join(bird_species, by = "AOU")
   
+  agg_taxa <- . %>% 
+    group_by(RouteDataID, taxon) %>%
+    summarise(count = sum(SpeciesTotal), .groups = "drop_last") 
+  
   out <- bind_rows(
-    species = joined %>% rename(taxon = species),
-    genus = joined %>% rename(taxon = genus),
-    family = joined %>% rename(taxon = family),
-    order = joined %>% rename(taxon = order),
+    species = joined %>% select(RouteDataID, taxon = Species, count = SpeciesTotal),  # no need to aggregate
+    genus = joined %>% rename(taxon = Genus) %>% agg_taxa,
+    family = joined %>% rename(taxon = Family) %>% agg_taxa,
+    order = joined %>% rename(taxon = Order) %>% agg_taxa,
     .id = "taxonomic_level"
   ) %>%
-    group_by(RouteDataID, taxonomic_level, taxon) %>%
-    summarise(count = sum(SpeciesTotal), .groups = "drop_last") %>%
+    group_by(RouteDataID, taxonomic_level) %>%
     summarise(
       n_taxa = n(),
       singletons = sum(count == 1),
       count_sum = sum(count), 
-      GCD = ifelse(n_taxa > 1, numbers::mGCD(count), NA)#only find GCD where no singletons for speed ( and > 1 taxa). Using ifelse as it doesn't run TRUE code when it would crash
+      GCD = ifelse(n_taxa > 1, numbers::mGCD(count), NA),#only find GCD where no singletons for speed ( and > 1 taxa). Using ifelse as it doesn't run TRUE code when it would crash
+      .groups = "drop_last"
     )
   return(out)
 }

### 2021-06-06 R/planktic_forams_plan.R
count summary table

@@ -17,10 +17,27 @@ planktic_foram_plan <- drake_plan(
               n_singletons = sum(count == 1), 
               min = min(count)),
   
-  foram_summary = list(
-    n_samples = n_distinct(foram_data$sampleID),
-    n_taxa = n_distinct(foram_data$taxon)
+  
+  # 
+  count_summary = bind_rows(
+    `Planktic foraminifera` = foram_summ %>% 
+      summarise(
+        `No. Samples` = n(),
+        `Median count sum` = median(count_sum),
+        `Median richness` = median(n_taxa),
+        `Singletons %` = mean(n_singletons >= 1) * 100,
+        `GCD1 %` = mean(gcd == 1) * 100
+      ), 
+    `Testate amoebae` = testate_summ1 %>% 
+      summarise(
+        `No. Samples` = n(),
+        `Median count sum` = median(count_sum),
+        `Median richness` = median(n_taxa),
+        `Singletons %` = mean(n_singletons >= 1) * 100,
+        `GCD1 %` = mean(gcd == 1) * 100
+      ), 
     
+    .id = "Dataset"
   )
   
 )# end of plan

### 2021-06-06 Rmd/count_check_MS.Rmd
improved methods, bird data

@@
 
+library(drake)
+library(tidyverse)
+library(english)
+library(rjt.misc)
+
 #set ggplot theme
@@
 
-loadd(bird_singleton_summary)
+loadd(bird_singleton_summary, bird_summ)
 loadd(testate_summ)
@@
 Given that count sums of percent data have been misreported, the possibility that other aspects of count data may be misreported needs to be considered.
-This paper aim to develop and test simple methods that can flag if the count data might have been misreported, or if the count sum of percent data is perhaps lower than reported.
 
 The key insight that allows inference about community or assemblage count data is that they represent a species abundance distribution.
-The ubiquitous shape of species abundance distributions when plotted as as the number of species against abundance is a hyperbolic or hollow curve with many rare and few abundant species [@McGill_2007].
-This relationship is central to ecological theory. 
-Communities or assemblages data that differ markedly this expected relationship
+The ubiquitous shape of species abundance distributions when plotted as the number of species against abundance is a hyperbolic or hollow curve with many rare and few abundant species and is central to ecological theory [@McGill_2007].
+From this, two predictions can be made: that counts of an entire community or assemblage should have many rare taxa, and that the distribution of abundances should be smooth.
+Count data that differ markedly from these predictions might not accurately reflect the original community or assemblage.
 
-For example, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual, hereafter singletons, unless the sampling effort is high relative to the taxonomic richness [@Coddington2009].
+This paper aims to develop and test simple methods to test the abundance of rare taxa and the smoothness of the species abundance distribution that can  use to flag cases if the count data might have been misreported.
+Next, as, regrettably, many palaeoecological data sets are archived as percent data, these tests are then extended to identify percent data where the count sum might be lower than reported.
+Finally, some complications and caveats are discussed.
 
-In the occasional assemblages without singletons, the greatest common divisor should usually be one, i.e.
+<!-- For example, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual, hereafter singletons, unless the sampling effort is high relative to the taxonomic richness [@Coddington2009].
+-->
+
+<!-- In the occasional assemblages without singletons, the greatest common divisor should usually be one, i.e.
 few assemblages should have all counts divisible by an integer larger than one, especially if species richness is high.
-
-Some complications and caveats are discussed.
-
-<!-- If the genuine assemblage counts are archived for all taxa, it is trivial to identify undercounts. -->
-<!-- Unfortunately, count data might be falsified so that it appears to meet the reported count sum. -->
-<!-- A more common problem is that, regrettably, many palaeoecologists archive percent data without an indication of the count sum. -->
+-->
 
@@
 ## Data
+
 Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea, and other sources.
 A range of ecological and palaeoecological data were sought to allow for differences in typical count sums and species richness.
-Fossil pollen assemblage in the Neotoma database [@neotoma_2018] (downloaded `r readd(pollen_download_date)`) with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded.
-Datasets with percent data where the minimum count sum was not reported in the associated paper were excluded.
+
+Count data analysed include the North American Breeding Bird survey [@bbsAssistant2]; 
+the Brown University planktic foraminiferal database [@Prell1999];
+the MOLTEN diatom database [@Elinor]; 
+the North American testate amoeba training set [@AMESBURY2018];
+and fossil pollen assemblage from the Neotoma database [@neotoma_2018]. 
+Pollen data were downloaded on `r readd(pollen_download_date)`, samples with with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded.
+
+Examples of percent data where the minimum count sum was reported in the associated paper were sought.
 Datasets discussed by @Telford2019 as possibly having misreported the minimum count sum were also excluded to avoid double reporting.
-Datasets with possible misreporting are anonymised, but no attempt is made to diagnose whether errors or misconduct are responsible.
+Datasets with possible misreporting are anonymised; no attempt is made to diagnose whether errors or misconduct are responsible.
 This paper does not attempt to be an exhaustive survey of all the data available.
@@
 
-For count data, the prevalence of counts without singletons and the 
-greatest common divisor of each count were calculated and related to the count sum.
+### Test for count data
 
-Generalised additive model proportion of explained deviance
-Proportion of taxa divisible by _k_ where $k > 1$
+To test for rare taxa, the prevalence of samples without singletons was calculated for each data set.
+
+To test if the species abundance distribution is smooth, I used two tests.
+First, for each assemblage or community with at least ` word(min_richness)` taxa, the proportion of taxa with counts divisible by $k$ where $k$ is a small integer larger than one is calculated.
+Samples where this proportion is high have a saw-toothed pattern of counts. 
+In extreme cases, where all taxa have counts divisible by $k$, the sample has a greatest common divisor of $k$.
+Data sets with a many samples with a greatest common divisor greater than one or an unusually high proportion, relative to the majority of pollen data set in the Neotoma database, of counts divisible by $k$ were flagged for closer inspection.
+
+The second test of smoothness is to fit a generalised additive model with Poisson errors to the counts aggregated over samples in each data set. 
+Only counts of ` word(gam_max)` or fewer were considered due to low power in the relatively infrequent larger counts.
+Data sets where the proportion of explained deviance is low relative to the majority of pollen data set in the Neotoma database were flagged for closer inspection.
+
+### Test for percent data
 
@@
 An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
-These tests are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.
+These tests for percent data are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.
+
+### Software
 
@@
 neotoma v. `r packageVersion("neotoma")` [@neotoma], 
+bbsAssistant v. `r packageVersion("bbsAssistant")` [@bbsAssistant1], 
 and tidyverse v. `r packageVersion("tidyverse")` [@tidyverse].
 Packages drake v. `r packageVersion("drake")` [@drake], 
-bookdown `r packageVersion("bookdown")` [@bookdown] and
+bookdown `r packageVersion("bookdown")` [@bookdown1] and
 renv v. `r packageVersion("renv")` [@renv] were used to ensure reproducibility.
-Code to replicate all the analyses is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms).
+Code to reproduce all the analyses is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms) and [zenodo.org](zenodo.org).
+All issues identified in the Neotoma database were reported to the maintainers.
 
@@
 
-The North American breeding bird survey is a very large, annual monitoring programme to track the status of North American bird populations [@NAmerican_Breeding_Bird2018].
+The North American breeding bird survey is a very large, annual monitoring programme to track the status of North American bird populations [@bbsAssistant2].
 All birds seen in point counts along ~40 km survey routes at the height of the breeding season are counted.
@@
 
-The vast majority of the bird routes have singletons at the species level (Table \@ref(tab:bird-table)).
-To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the birds routes to progressively lower taxonomic resolutions.
+The vast majority of the bird routes have singletons at the species level.
+Only `r bird_summ$n_without_singletons` routes do not have singletons, of these only `r bird_summ$n_GCD_gt1 %>% nrow() %>% words()`, all with very low richness, have a greatest common divisor greater than one.
+
+To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the data to progressively lower taxonomic resolutions.
 As richness declines, the proportion of routes having singletons declines (Table \@ref(tab:bird-table)), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
@@
 
-To test the sensitivity of the prevalence of routes without singletons to the count sum, I resample the routes with 400 or more observations to smaller count sums using a multivariate hypergeometric distribution.
+To test the sensitivity of the prevalence of routes without singletons to the count sum, I resample the routes with 400 or more individuals to smaller count sums using a multivariate hypergeometric distribution.
 At the order level, the proportion of counts with singletons increases steeply until it reaches a maximum at about 200, thereafter is declines slowly (Fig. \@ref(fig:bird-recount)).
@@
 
-The `r nrow(owens_summ1)` diatom counts from Owen's Lake span a wide range of count sums (`r min(owens_summ1$count_sum)` -- `r max(owens_summ1$count_sum)`) because, if concentrations were low, the requirement that at least 300 valves were counted was replaced by a minimum area of slide to count [@Platt1997].
-Discounting the `r sum(owens_summ1$n_taxa == 1)` mono-specific assemblages (all with very low count sums`r max(owens_summ1$count_sum[owens_summ1$n_taxa == 1]) %>% insist(.
-<= 10) %>% invisible()`), `r owens_summ$p_singletons`% of assemblages have singletons with a median number of singletons per assemblage of `r owens_summ$median_sing %>% words()`.
-Assemblages with count sums above 50 have a higher chance of having singletons (`r owens_summ$no_single_gt_50`% vs `r owens_summ$no_single50_`%).
-`r owens_summ$gcd2`% of assemblages have a greatest common divisor of one, the remainder have a greatest common divisor of `r owens_summ$gcd_max %>% words()`.
-Assemblages with a greatest common divisor above one all have low taxonomic richness`r owens_summ1 %>% filter(gcd > 1) %>% pull(n_taxa) %>% max() %>% insist(. <= 6) %>% invisible()`.
+The ` nrow(owens_summ1)` diatom counts from Owen's Lake span a wide range of count sums (` min(owens_summ1$count_sum)` -- ` max(owens_summ1$count_sum)`) because, if concentrations were low, the requirement that at least 300 valves were counted was replaced by a minimum area of slide to count [@Platt1997].
+Discounting the ` sum(owens_summ1$n_taxa == 1)` mono-specific assemblages (all with very low count sums ` max(owens_summ1$count_sum[owens_summ1$n_taxa == 1]) %>% insist(.
+<= 10) %>% invisible()`), ` owens_summ$p_singletons`% of assemblages have singletons with a median number of singletons per assemblage of ` owens_summ$median_sing %>% words()`.
+Assemblages with count sums above 50 have a higher chance of having singletons (` owens_summ$no_single_gt_50`% vs ` owens_summ$no_single50_`%).
+` owens_summ$gcd2`% of assemblages have a greatest common divisor of one, the remainder have a greatest common divisor of ` owens_summ$gcd_max %>% words()`.
+Assemblages with a greatest common divisor above one all have low taxonomic richness` owens_summ1 %>% filter(gcd > 1) %>% pull(n_taxa) %>% max() %>% insist(. <= 6) %>% invisible()`.
 
@@
 Of the `r testate_summ$n_gcd1 %>% words()` assemblages with a greatest common divisor above one, `r testate_summ$n_gcd2_low_div %>% words()` have fewer than five taxa; `r testate_summ$n_gcd2 %>% words()` have a greatest common divisor of two, and the other `r testate_summ$n_gcd3 %>% words()` has a greatest common divisor of three.
+
+```{r multitaxa}
+options(digits = 4)
+readd(count_summary) %>% 
+  knitr::kable(caption = "Properties of different data sets")
+options(digits = 2)
+```
 
@@
 The Owen's Lake dataset [@Platt1997] includes percentages for each taxa, allowing the count sums estimated from the percent data to be verified.
-Excluding mono-specific assemblages, for which no meaningful estimate of the count sum is possible, the minimum percent method and the direct search method correctly estimate the count sum for, respectively, `r owens_est_summ$minpc`% and `r owens_est_summ$direct`% of the assemblages.
+Excluding mono-specific assemblages, for which no meaningful estimate of the count sum is possible, the minimum percent method and the direct search method correctly estimate the count sum for, respectively, ` owens_est_summ$minpc`% and ` owens_est_summ$direct`% of the assemblages.
 The few assemblages that fail the direct search method are species poor and have low count sums.

### 2022-01-03 R/bird_functions.R
summarise counts function

@@ -26,13 +26,8 @@ bird_process <- function(bird_data, bird_species) {
     .id = "taxonomic_level"
   ) %>%
     group_by(RouteDataID, taxonomic_level) %>%
-    summarise(
-      n_taxa = n(),
-      singletons = sum(count == 1),
-      count_sum = sum(count), 
-      GCD = ifelse(n_taxa > 1, numbers::mGCD(count), NA),#only find GCD where no singletons for speed ( and > 1 taxa). Using ifelse as it doesn't run TRUE code when it would crash
-      .groups = "drop_last"
-    )
+    summarise_counts()
+
   return(out)
 }
 

### 2022-01-03 R/download_birds.R

@@ -22,13 +22,13 @@ bird_plan <- drake_plan(
   bird_singleton_summary = bird_singletons %>%
     group_by(taxonomic_level) %>%
     summarise(
-      p_singletons = mean(singletons > 0) * 100,
-      n_singletons = sum(singletons == 0),
+      p_singletons = mean(n_singletons > 0) * 100,
+      n_singletons = sum(n_singletons == 0),
       median_richness = median(n_taxa),
       n_routes = n(),
-      GCD1 = mean(GCD == 1, na.rm = TRUE) * 100,
-      GCD2plus = mean(GCD > 1, na.rm = TRUE) * 100,
-      GCD_max = max(GCD, na.rm = TRUE), 
+      GCD1 = mean(gcd == 1, na.rm = TRUE) * 100,
+      GCD2plus = mean(gcd > 1, na.rm = TRUE) * 100,
+      GCD_max = max(gcd, na.rm = TRUE), 
       .groups = "drop_last"
     ) %>%
     mutate(taxonomic_level = factor(
@@ -38,8 +38,8 @@ bird_plan <- drake_plan(
     arrange(taxonomic_level),
   
   bird_summ = list(
-    n_without_singletons = bird_singletons %>% filter(taxonomic_level == "species", singletons == 0) %>% nrow(),
-    n_GCD_gt1 = bird_singletons %>% filter(taxonomic_level == "species", GCD > 1 | is.na(GCD))
+    n_without_singletons = bird_singletons %>% filter(taxonomic_level == "species", n_singletons == 0) %>% nrow(),
+    n_GCD_gt1 = bird_singletons %>% filter(taxonomic_level == "species", gcd > 1 | is.na(gcd))
     
   ),
   

### 2022-01-03 R/download_testates.R

@@ -35,11 +35,7 @@ testate_plan <- drake_plan(
   
   #run analyses
   testate_summ1 = testate_counts %>% 
-    summarise(count_sum = sum(count), 
-              n_taxa = n(), 
-              gcd = numbers::mGCD(count), 
-              n_singletons = sum(count == 1), 
-              min = min(count)) %>% 
+    summarise_counts() %>% 
     assert(within_bounds(1, 3), gcd),
 
  

### 2022-01-03 R/functions/general_functions.R

@@ -1,0 +1,41 @@
+## functional sequence to calculate GCD and singleton data per sample
+
+summarise_counts <- . %>% 
+  summarise(
+    count_sum = sum(count), 
+    n_taxa = n(), 
+    min = min(count),
+    n_singletons = sum(count == 1), 
+    # GCD is trivially one if there are singletons so no need to calculate
+    gcd = fast_mGCD(n_taxa = n_taxa, count = count, n_singletons = n_singletons, min = min), 
+    .groups = "drop_last"
+  )
+
+#only find GCD where no singletons for speed
+# When n_taxa == 1, mGCD will give an error, so return min
+fast_mGCD <- function(n_taxa, count, n_singletons, min){
+  if (n_taxa == 1){# mGCD will crash. Result is trivially min
+    return(min)
+  }
+  else if (n_singletons > 0){ #GCD is trivially 1
+    return(1)
+  }
+  else {
+    numbers::mGCD(count)
+  }
+}
+
+## functional sequence to summarise GCD and singleton data per data set
+
+summarise_singletons <- . %>% 
+  filter(count_sum >= 50, n_taxa > 1) %>% 
+  summarise(
+    `Nr. Samples` = n(),
+    `Median count sum` = median(count_sum),
+    `Median richness` = median(n_taxa),
+    `Without singletons` = glue("{sum(n_singletons == 0)} ({round(mean(n_singletons == 0) * 100, 3)}%)"),
+    `GCD > 1` = glue("{sum(gcd > 1)} ({round(mean(gcd > 1) * 100, 3)}%)"),
+    .groups = "drop"
+  ) %>% 
+  mutate(across(c(`Without singletons`, `GCD > 1`), as.character))
+

### 2022-01-03 R/planktic_forams_plan.R

@@ -11,33 +11,67 @@ planktic_foram_plan <- drake_plan(
     
   foram_summ = foram_data %>% 
     group_by(sampleID) %>% 
-    summarise(count_sum = sum(count), 
-              n_taxa = n(), 
-              gcd = numbers::mGCD(count), 
-              n_singletons = sum(count == 1), 
-              min = min(count)),
+    summarise_counts(),
   
   
-  # 
+  # summary table of all count data
   count_summary = bind_rows(
+    `Birds - species` = bird_singletons %>% 
+      ungroup() %>%
+      filter(taxonomic_level == "species") %>% 
+      summarise_singletons(),
+    `Birds - genus` = bird_singletons %>% 
+      ungroup() %>%
+      filter(taxonomic_level == "genus") %>% 
+      summarise_singletons(),
+    `Birds - family` = bird_singletons %>% 
+      ungroup() %>%
+      filter(taxonomic_level == "family") %>% 
+      summarise_singletons(),
+    `Birds - order` = bird_singletons %>% 
+      ungroup() %>%
+      filter(taxonomic_level == "order") %>% 
+      summarise_singletons(),
+    `Coastal diatoms` = molten_singletons %>% 
+      summarise_singletons(),
     `Planktic foraminifera` = foram_summ %>% 
-      summarise(
-        `No. Samples` = n(),
-        `Median count sum` = median(count_sum),
-        `Median richness` = median(n_taxa),
-        `Singletons %` = mean(n_singletons >= 1) * 100,
-        `GCD1 %` = mean(gcd == 1) * 100
-      ), 
+      summarise_singletons(), 
     `Testate amoebae` = testate_summ1 %>% 
-      summarise(
-        `No. Samples` = n(),
-        `Median count sum` = median(count_sum),
-        `Median richness` = median(n_taxa),
-        `Singletons %` = mean(n_singletons >= 1) * 100,
-        `GCD1 %` = mean(gcd == 1) * 100
-      ), 
-    
+      summarise_singletons(), 
+    `Pollen` = pollen_summ1 %>% 
+      ungroup() %>% 
+      summarise_singletons(),
     .id = "Dataset"
-  )
+  ),
   
+  richness_singleton_plot = bind_rows(
+    `Birds - species` = bird_singletons %>% 
+      ungroup() %>%
+      filter(taxonomic_level == "species"),
[…]

### 2022-01-03 R/download_diatoms.R
use molten data

@@ -1,69 +1,40 @@
 #diatoms
 diatom_plan <- drake_plan(
-  ####Owens Lake####
-  owens = read_file(file = "http://diatom.ansp.org/dpdc/download.asp?type=count&DatasubsetID=51") %>% 
-    gsub(pattern = "<.*?>", replacement = "", x = .) %>% 
-    read_delim(delim = "\t"),
+  #### MOLTEN ####
   
-  owens_summ1 = owens %>% 
-  group_by(SampleId) %>% 
-  summarise(mn = min(Count), n_sing = sum(Count == 1), count_sum = sum(Count), n_taxa = n(), gcd = ifelse(n_taxa > 1, numbers::mGCD(Count), NA)),
-  
-  owens_summ = owens_summ1 %>% 
-    filter(n_taxa > 1) %>% 
-    summarise(
-      n = n(),
-      count_min = min(count_sum),
-      count_max = max(count_sum),
-      median_sing = median(n_sing),
-      p_singletons  = mean(n_sing  > 0) * 100,
-      no_singletons = mean(mn != 1) * 100,
-      no_single_gt_50 = mean(mn[count_sum > 50] == 1) * 100, 
-      no_single50_ = mean(mn[count_sum <= 50] == 1) * 100, 
-      gcd2 = mean(gcd == 1) * 100,
-      gcd_max = max(gcd, na.rm = TRUE)),
-  
-  ####owens diatoms percent####
-  owens_est = owens %>% 
-    estimate_n(ID_cols = c("SampleId", "SampleCode", "CountSum"), percent_col = "Percent", taxon_col = "TaxonShortName", digits = 4) %>%
-    assertr::verify(map_int(direct_search_est, nrow) == 1) %>% #check only one row in each direct_search_est
-    unnest(direct_search_est) %>% 
-    assertr::verify(score == 1),#check score ==1
-  
-  owens_est_summ = owens_est %>% 
-    select(est_n_minpc, est_n_direct, CountSum, n_taxa) %>% 
-    filter(n_taxa > 1) %>% 
-    ungroup() %>% 
-    summarise(
-      minpc = mean(round(est_n_minpc, 3) == CountSum) * 100, 
-      direct = mean(est_n_direct == CountSum) * 100
-    ),
-  
-  owens_fail = owens_est %>% 
-    select(est_n_minpc, est_n_direct, CountSum, n_taxa) %>%
-    filter(n_taxa > 1, est_n_direct != CountSum),
-  
-  owens_fail_minpc = owens_est %>% 
[…]

### 2022-01-03 make_manuscript.R
eof

@@ -16,3 +16,4 @@ if(length(drake_failed()) == 0){
 
 #show dependency graph
 r_vis_drake_graph(source = plan, targets_only = TRUE, main = "count ms dependency graph")
+

### 2022-01-03 plan_count_check_ms_with_drake.R
rounding plan, coment out  diatoms

@@ -6,11 +6,11 @@ library("readxl")
 library("countSum")
 library("assertr")
 library("rjt.misc")
-library("english")
 library("neotoma")
 library("neotoma2tibble")
 library("bbsAssistant")
 library("janitor")
+library("glue")
 
 #drake configuration
 pkgconfig::set_config("drake::strings_in_dots" = "literals")
@@ -25,9 +25,13 @@ source("R/download_chironomids.R")
 source("R/download_neotoma.R")
 source("R/download_diatoms.R")
 source("R/download_marine.R")
-source("R/download_neotoma_diatoms.R")
+#source("R/download_neotoma_diatoms.R")
 source("R/planktic_forams_plan.R")
+source("R/percent_rounding_plan.R")
 
+
+# Import functions
+source("R/functions/general_functions.R")
 
 #construct drake plan
 analyses <- drake_plan(
@@ -50,7 +54,8 @@ analyses <- drake_plan(
   
   #add extra packages to bibliography
   biblio2 = package_citations(
-    packages = c("extraDistr", "countSum", "numbers", "neotoma", "drake", "tidyverse", "bookdown", "renv", "bbsAssistant"), 
+    packages = c("extraDistr", "countSum", "numbers", "neotoma", "bbsAssistant",
+                 "drake", "tidyverse", "bookdown", "renv"), 
     old_bib = file_in("Rmd/extra/countMS.bib"), 
     new_bib = file_out("Rmd/extra/countMS2.bib")),
   
@@ -104,8 +109,9 @@ plans <- bind_rows(
   pollen_plan,
   diatom_plan,
   marine_plan,
-  neotoma_diatoms_plan,
+#  neotoma_diatoms_plan,
   planktic_foram_plan,
+  percent_rounding_plan,
   analyses)
 
 #quick network plot

### 2022-01-03 R/percent_rounding_plan.R
rounding plan

@@ -1,0 +1,28 @@
+
+percent_rounding_plan <- drake_plan(
+  percent_rounding_sim = {
+    dat <- tibble(spp = c("a", "b", "c"), count = c(1, 5, 19)) %>% 
+      crossing(count_sum = 50:1000) %>% 
+      mutate(percent = count / count_sum * 100)
+  
+    1:4 %>% 
+      set_names() %>% 
+      map_dfr(~{
+        mutate(dat, percent = round(percent, .x)) %>% 
+        estimate_n(percent_col = "percent", taxon_col = "spp", ID_cols = "count_sum",digits = .x, nmax = 1500) %>% 
+          select(-direct_search, -data, -n_taxa) %>% 
+          unnest(cols = direct_search_est)
+    }, .id = "round")
+  },
+  
+  percent_rounding_plot =  percent_rounding_sim %>% 
+    pivot_longer(c(est_n_direct, est_n_minpc), names_to = "method", values_to = "estimate") %>% 
+    mutate(method = factor(method, levels = c("est_n_minpc", "est_n_direct"), labels = c("Minimum percent", "Direct search"))) %>% 
+    ggplot(aes(x = count_sum, y = round(estimate) - count_sum, colour = round)) +
+    geom_line() + 
+    scale_colour_viridis_d() +
+    labs(x = "Count sum", y = "Estimated - actual count sum", colour = "Digits") +
+    facet_wrap(~method) +
+    theme(legend.position = c(0.01, .99), legend.justification = c(0, 1))
+
+)#end of plan

### 2022-01-03 Rmd/count_check_MS.Rmd
updates, -diatoms

@@
 
-library(drake)
-library(tidyverse)
-library(english)
-library(rjt.misc)
+# Packages loaded for convenience when knitting without running r_make
+library("drake")
+library("tidyverse")
+library("rjt.misc")
+
+#packages only loaded in this file
+library("kableExtra")
+library("english")
 
@@
 
-Count data analysed include the North American Breeding Bird survey [@bbsAssistant2]; 
-the Brown University planktic foraminiferal database [@Prell1999];
-the MOLTEN diatom database [@Elinor]; 
-the North American testate amoeba training set [@AMESBURY2018];
-and fossil pollen assemblage from the Neotoma database [@neotoma_2018]. 
-Pollen data were downloaded on `r readd(pollen_download_date)`, samples with with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded.
+Count data analysed include the 1) North American Breeding Bird survey [@bbsAssistant2], a very large, annual monitoring programme to track the status of North American bird populations.
+All birds seen in point counts along ~40 km survey routes at the height of the breeding season are recorded. 
+2) The Brown University planktic foraminiferal database [@Prell1999].
+3) The MOLTEN diatom database of modern coastal diatom assemblages [@Clarke2006].
+4) The North American testate amoeba training set [@AMESBURY2018].
+5) Fossil pollen assemblage from the Neotoma database [@neotoma_2018] downloaded on `r readd(pollen_download_date)`.
+
+Counts with with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded from all datasets.
 
@@
 
-To test for rare taxa, the prevalence of samples without singletons was calculated for each data set.
+The prevalence of counts without singletons, and counts with a greatest common divisor greater than one, was calculated for each data set.
 
@@
 
-The North American breeding bird survey is a very large, annual monitoring programme to track the status of North American bird populations [@bbsAssistant2].
-All birds seen in point counts along ~40 km survey routes at the height of the breeding season are counted.
-The very large number of routes, over `r format(floor(bird_singleton_summary$n_routes[1]/1000) * 1000, big.mark = ",")`, typically with large numbers of individuals, allows the ideas developed above to be tested: do most routes have singletons and is the greatest common divisor one, and identify circumstances for which they fail.
-This can be tested for the observed data, and by simulating smaller counts and lower richness.
+Across all the datasets, only a small minority of samples lack singletons, and even fewer have a greatest common divisor larger than one (Table \@ref(tab:multitaxa).
+Most of the samples without singletons, or with a greatest common divisor larger than one, have low richness (Fig. \@ref(fig:richness-singleton-plot).
 
-The vast majority of the bird routes have singletons at the species level.
-Only `r bird_summ$n_without_singletons` routes do not have singletons, of these only `r bird_summ$n_GCD_gt1 %>% nrow() %>% words()`, all with very low richness, have a greatest common divisor greater than one.
+Although the proportion of pollen assemblages that lack singletons or have a greatest common divisor of one are low, such assemblages are not randomly distributed among the `r pollen_summ$n_datasets %>% format()` pollen datasets.
 
-To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the data to progressively lower taxonomic resolutions.
-As richness declines, the proportion of routes having singletons declines (Table \@ref(tab:bird-table)), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
-The vast majority of counts have a greatest common divisor of one, except when taxonomic richness is low.
+Half of the assemblages without singletons are in just `r pollen_summ$half_no_sing` datasets.
+`r Words(pollen_summ$half_gcd)` datasets contain half of the assemblages with a greatest common divisor above one and only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
+Some pollen datasets with unexpected properties are discussed in section \@ref(unexpected-count-data).
 
-```{r bird-table}
-options(digits = 4)
-bird_singleton_summary %>%
-  select(`Taxonomic level` = taxonomic_level, `Percent with singletons` = p_singletons, `Percent GCD = 1` = GCD1, `Median richness` = median_richness) %>%
-  knitr::kable(digits = c(0, 6, 6, 3), caption = "Percent of bird counts from the North American breeding bird survey with singletons and of greatest common divisor (GCD) of one, at different taxonomic levels to mimic the effect of taxonomic richness.")
-options(digits = 2)
+```{r multitaxa}
+readd(count_summary) %>% 
+  #manually escape % as cannot use kbl(escape = TRUE) with linebreak
+  mutate(across(where(is.character), str_replace, pattern = "%", replacement = "\\\\%")) %>% 
+  kbl(x = ., booktabs = TRUE, escape = FALSE,
+      caption = "Properties of different data sets", 
+      col.names = linebreak(str_wrap(names(.), width = 10))
+    ) %>% 
+  kable_styling(latex_options = c("basic", "hold_position"))
 ```
 
-To test the sensitivity of the prevalence of routes without singletons to the count sum, I resample the routes with 400 or more individuals to smaller count sums using a multivariate hypergeometric distribution.
+```{r richness-singleton-plot, fig.cap = "Relationship betwen the presence of singletons and taxonomic richness for the different datasets.
+Samples with a greatest common divisor greater than one are shown in dark blue.
+Curve is fitted with a generalised additive model", fig.height = 120/24.5, fig.width=160/24.5}
+readd(richness_singleton_plot)
+```
+
+### Effect of richness on the prevalence of singletons
+
+To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the bird data to progressively lower taxonomic resolutions.
+As richness declines, the proportion of counts without singletons increases (Table \@ref(tab:multitaxa)), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
+Even when richness is low, only a small minority of counts have a greatest common divisor greater than one.
+
+### Effect of count sum on the prevalence of singletons
+
+To test the sensitivity of the prevalence of counts without singletons to the count sum, I resample the bird counts with 400 or more individuals to smaller count sums using a multivariate hypergeometric distribution.
 At the order level, the proportion of counts with singletons increases steeply until it reaches a maximum at about 200, thereafter is declines slowly (Fig. \@ref(fig:bird-recount)).
@@
 
-The ` nrow(owens_summ1)` diatom counts from Owen's Lake span a wide range of count sums (` min(owens_summ1$count_sum)` -- ` max(owens_summ1$count_sum)`) because, if concentrations were low, the requirement that at least 300 valves were counted was replaced by a minimum area of slide to count [@Platt1997].
-Discounting the ` sum(owens_summ1$n_taxa == 1)` mono-specific assemblages (all with very low count sums ` max(owens_summ1$count_sum[owens_summ1$n_taxa == 1]) %>% insist(.
-<= 10) %>% invisible()`), ` owens_summ$p_singletons`% of assemblages have singletons with a median number of singletons per assemblage of ` owens_summ$median_sing %>% words()`.
-Assemblages with count sums above 50 have a higher chance of having singletons (` owens_summ$no_single_gt_50`% vs ` owens_summ$no_single50_`%).
-` owens_summ$gcd2`% of assemblages have a greatest common divisor of one, the remainder have a greatest common divisor of ` owens_summ$gcd_max %>% words()`.
-Assemblages with a greatest common divisor above one all have low taxonomic richness` owens_summ1 %>% filter(gcd > 1) %>% pull(n_taxa) %>% max() %>% insist(. <= 6) %>% invisible()`.
+### Smooth distributions
 
-The `r testate_summ$n_assemblages` assemblages in the North American testate amoeba training set with available count data [@AMESBURY2018] have count sums between `r testate_summ$count_min` and `r testate_summ$count_max` tests (median = `r testate_summ$count_median`), and taxonomic richness ranges between `r testate_summ$tax_min` and `r testate_summ$tax_max` (median = `r testate_summ$tax_median`) taxa.
-`r testate_summ$p_singletons`% of the assemblages have at least one singleton (median `r testate_summ$median_singletons`) and `r testate_summ$p_gcd1`% of assemblages have a greatest common divisor of one.
-Of the `r testate_summ$n_gcd1 %>% words()` assemblages with a greatest common divisor above one, `r testate_summ$n_gcd2_low_div %>% words()` have fewer than five taxa; `r testate_summ$n_gcd2 %>% words()` have a greatest common divisor of two, and the other `r testate_summ$n_gcd3 %>% words()` has a greatest common divisor of three.
+## Estimating the count sum from percent
 
-```{r multitaxa}
-options(digits = 4)
-readd(count_summary) %>% 
-  knitr::kable(caption = "Properties of different data sets")
-options(digits = 2)
+The precision of both methods for estimating the count sum from percent data depend on count sum and on the precision to which the percent are recorded (Fig. \@ref(fig:percent-est-plot).
+In general, the direct search method is more precise and can give useful estimates when the assemblages have no singletons.
+
+```{r percent-est-plot, fig.cap = "Accuracy of estimates of the count sum calculated with two different methods with percent data have been rounded to different numbers of digits, based on some simulated data", fig.width = 80/25.4, fig.height = 80/25.4}
+readd(percent_rounding_plot) + theme(legend.position = "bottom")
 ```
-
-The pollen data from the Neotoma database included over `r pollen_summ$n_assemblages` assemblages in `r pollen_summ$n_datasets %>% format()` datasets. 
-Conforming to expectations, nearly all assemblages have singletons (`r pollen_summ$p_singletons`%) and a greatest common divisor of one (`r pollen_summ$p_gcd1`%). 
-The assemblages without singletons or with a greatest common divisor above one are not randomly distributed among the datasets.
-Half of the assemblages without singletons are in `r pollen_summ$half_no_sing` datasets.
-`r Words(pollen_summ$half_gcd)` datasets contain half of the `r pollen_summ$n_gcd1` assemblages with a greatest common divisor above one and only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
-Some Neotoma datasets are with a high proportion of assemblages without singletons or a with a greatest common divisor above one are discussed in section \@ref(unexpected-count-data) of this paper.
-
-## Estimating the percent sum
-
-The Owen's Lake dataset [@Platt1997] includes percentages for each taxa, allowing the count sums estimated from the percent data to be verified.
-Excluding mono-specific assemblages, for which no meaningful estimate of the count sum is possible, the minimum percent method and the direct search method correctly estimate the count sum for, respectively, ` owens_est_summ$minpc`% and ` owens_est_summ$direct`% of the assemblages.
-The few assemblages that fail the direct search method are species poor and have low count sums.
 
@@
 Counts with blue triangles are from assemblages with a greatest common divisor of one.
-These appear to have been linearly interpolated between the adjacent counts from assemblages with a greatest common divisor of two (red circles).
+These have been linearly interpolated between the adjacent counts from assemblages with a greatest common divisor of two (red circles).
 B) Percent of taxa in each assemblage where the count is the mean of the counts in the adjacent assemblages, excluding runs of zero counts.

### 2022-12-31 manuscript/count_check_MS.Rmd
rename rmd dir

--- 2022-01-03 Rmd/count_check_MS.Rmd
@@
 Six percent of papers published in _Molecular and Cellular Biology_ show evidence of inappropriate image duplication [@Bik2018] either due to error or, more rarely, misconduct.
-It would be reckless to assume that the palaeoecological literature does not have equivalent problems: the low rate of retractions in the ecological and geological literature [@Grieneisen2012] may partially reflect our limited ability to detect errors and misconduct rather than their prevalence.
+It would, especially after the spate of retractions in behavioural ecology [@Pennisi2021], be reckless to assume that the ecological and palaeoecological literature is devoid of errors and misconduct.
+The low rate of retractions in the ecological and geological literature [@Grieneisen2012] may partially reflect our limited ability to detect errors and misconduct rather than their prevalence.
 
 Several numerical tools have been developed to identify questionable data.
-Deviations from the distribution of digits expected from the Newcomb-Benford law has been used to detect issues with scientific and financial data [@Barabesi2017].
+Deviations from the distribution of digits expected from the Newcomb-Benford law have been used to detect issues with scientific and financial data [@Barabesi2017].
 @Carlisle2017 identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data.
@@
 It could also happen knowingly, especially when microfossil preservation is poor or concentrations are low.
-This risk that count sums will be misreported is demonstrated by @Larocque-Tobler2016 who acknowledged that the chironomid count sums in @Larocque-Tobler2015 were not at least fifty head capsules as reported but as low as nineteen.
+This risk that count sums could be misreported is demonstrated by @Larocque-Tobler2016 who acknowledged that the chironomid count sums in @Larocque-Tobler2015 were not at least fifty head capsules as reported but as low as nineteen.
 @Telford2019 reports some other cases where count sums may be much lower than reported.
 
-Given that count sums of percent data have been misreported, the possibility that other aspects of count data may be misreported needs to be considered.
+Given that count sums have been misreported, the possibility that other aspects of count data may be misreported needs to be considered.
 
-The key insight that allows inference about community or assemblage count data is that they represent a species abundance distribution.
-The ubiquitous shape of species abundance distributions when plotted as the number of species against abundance is a hyperbolic or hollow curve with many rare and few abundant species and is central to ecological theory [@McGill_2007].
-From this, two predictions can be made: that counts of an entire community or assemblage should have many rare taxa, and that the distribution of abundances should be smooth.
+The key insight that allows inference about community or assemblage count data is that they can be represented by a species abundance distribution.
+The ubiquitous shape of species abundance distributions when plotted as the number of species against abundance is a hyperbolic curve with many rare and few abundant species and is central to ecological theory [@McGill_2007].
+From this, two predictions can be made: that counts of an entire community or assemblage should have many rare taxa, and that the distribution of abundances should have a smooth and monotonic decline.
 Count data that differ markedly from these predictions might not accurately reflect the original community or assemblage.
 
-This paper aims to develop and test simple methods to test the abundance of rare taxa and the smoothness of the species abundance distribution that can  use to flag cases if the count data might have been misreported.
+This paper aims to develop simple methods to test the abundance of rare taxa and the smoothness of the species abundance distribution that can be used to flag cases where the count data might have been misreported.
 Next, as, regrettably, many palaeoecological data sets are archived as percent data, these tests are then extended to identify percent data where the count sum might be lower than reported.
@@
 
-<!-- For example, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual, hereafter singletons, unless the sampling effort is high relative to the taxonomic richness [@Coddington2009].
+<!-- For example, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will  unless the sampling effort is high relative to the taxonomic richness [@Coddington2009].
 -->
@@
 Count data analysed include the 1) North American Breeding Bird survey [@bbsAssistant2], a very large, annual monitoring programme to track the status of North American bird populations.
-All birds seen in point counts along ~40 km survey routes at the height of the breeding season are recorded. 
+The survey reports all birds seen in point counts along ~40 km survey routes at the height of the breeding season. 
 2) The Brown University planktic foraminiferal database [@Prell1999].
@@
 
-The prevalence of counts without singletons, and counts with a greatest common divisor greater than one, was calculated for each data set.
+Following the assumptions derived from the shape of species abundance distributions that there should be many rare taxa, I test for the presence of taxa represented by a single individual, hereafter singletons, in each sample.
 
-To test if the species abundance distribution is smooth, I used two tests.
-First, for each assemblage or community with at least ` word(min_richness)` taxa, the proportion of taxa with counts divisible by $k$ where $k$ is a small integer larger than one is calculated.
-Samples where this proportion is high have a saw-toothed pattern of counts. 
-In extreme cases, where all taxa have counts divisible by $k$, the sample has a greatest common divisor of $k$.
-Data sets with a many samples with a greatest common divisor greater than one or an unusually high proportion, relative to the majority of pollen data set in the Neotoma database, of counts divisible by $k$ were flagged for closer inspection.
+Following the assumption that the species abundance distribution should show a smooth monotonic decline, I use two related tests. 
+First, I calculate the greatest common divisor greater for each sample. 
+The greatest common divisor is expected to be one, but would be two if all counts can be divided by two, giving a saw-toothed species abundance distribution.
+The second test is a relaxed version of the first.
+Rather than finding samples where all the counts have a greatest common divisor greater than one, it finds the the proportion of taxa with counts divisible by $k$ where $k$ is a small integer larger than one is calculated.
 
-The second test of smoothness is to fit a generalised additive model with Poisson errors to the counts aggregated over samples in each data set. 
-Only counts of ` word(gam_max)` or fewer were considered due to low power in the relatively infrequent larger counts.
-Data sets where the proportion of explained deviance is low relative to the majority of pollen data set in the Neotoma database were flagged for closer inspection.
+Data sets with a many samples without singletons, a greatest common divisor greater than one, or an unusually high proportion, relative to the majority of pollen data set in the Neotoma database, of counts divisible by $k$ were flagged for closer inspection.
 
@@
 It is also expected that the percent _p_ is calculated from integer counts, therefore there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
-Possible values for _N_ can be found by a direct search algorithm over the range of plausible values of _N_.
+Possible values for _N_ can be found by a direct search over the range of plausible values of _N_.
 An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
@@
 Packages drake v. `r packageVersion("drake")` [@drake], 
-bookdown `r packageVersion("bookdown")` [@bookdown1] and
+bookdown v. `r packageVersion("bookdown")` [@bookdown1] and
 renv v. `r packageVersion("renv")` [@renv] were used to ensure reproducibility.
@@
 
-Across all the datasets, only a small minority of samples lack singletons, and even fewer have a greatest common divisor larger than one (Table \@ref(tab:multitaxa).
-Most of the samples without singletons, or with a greatest common divisor larger than one, have low richness (Fig. \@ref(fig:richness-singleton-plot).
+Across all the datasets, only a small minority of samples lack singletons, and even fewer have a greatest common divisor larger than one (Table \@ref(tab:multitaxa)).
+Most of the samples without singletons, or with a greatest common divisor larger than one, have low richness (Fig. \@ref(fig:richness-singleton-plot)).
 
 Although the proportion of pollen assemblages that lack singletons or have a greatest common divisor of one are low, such assemblages are not randomly distributed among the `r pollen_summ$n_datasets %>% format()` pollen datasets.
-
 Half of the assemblages without singletons are in just `r pollen_summ$half_no_sing` datasets.
@@
 To test the sensitivity of the prevalence of counts without singletons to the count sum, I resample the bird counts with 400 or more individuals to smaller count sums using a multivariate hypergeometric distribution.
-At the order level, the proportion of counts with singletons increases steeply until it reaches a maximum at about 200, thereafter is declines slowly (Fig. \@ref(fig:bird-recount)).
-The prevalence of counts with singletons is low for small counts because it is possible that only the common taxa are counted; with larger counts, the chance of counting a single individual of a rare taxon increases.
-With even larger counts the prevalence of routes with singletons decreases due to saturation as singletons become doubletons and few new taxa are found.
+At the order level, the proportion of counts without singletons decreases steeply until it reaches a minimum at when the count sum is about 200, thereafter it increases slowly (Fig.
+\@ref(fig:bird-recount)).
+The prevalence of counts without singletons is high for small counts because it is possible that only the common taxa are counted; with larger counts, the chance of counting a single individual of a rare taxon increases.
+With even larger counts the prevalence of routes without singletons increases due to saturation as singletons become doubletons and few new taxa are found.
+The latter effect may largely be an artefact as in the simulation, the list of possible taxa to sample is short, whereas in the wild, there are many rare taxa which might be observed.
 
 ```{r bird-recount, fig.cap="Effect of count sum on the proportion of bird counts from the North American breeding bird survey without singleton at the order level.
-Results are the mean of ten trials, error bars are two standard errors.", fig.width = 80/25.4, fig.height = 70/25.4}
+Results are the mean of ten trials with 95% confidence interval.", fig.width = 80/25.4, fig.height = 70/25.4}
 readd(bird_recount_singleton_plot)
@@
 
-### Smooth distributions
-
 ## Estimating the count sum from percent
-
-The precision of both methods for estimating the count sum from percent data depend on count sum and on the precision to which the percent are recorded (Fig. \@ref(fig:percent-est-plot).
-In general, the direct search method is more precise and can give useful estimates when the assemblages have no singletons.
-
-```{r percent-est-plot, fig.cap = "Accuracy of estimates of the count sum calculated with two different methods with percent data have been rounded to different numbers of digits, based on some simulated data", fig.width = 80/25.4, fig.height = 80/25.4}
-readd(percent_rounding_plot) + theme(legend.position = "bottom")
-```
 
@@
 Half counts make the estimated count too high by a factor of two, so do not risk incorrectly flagging count sums as being too small.
-The direct search method is more precise than the minimum percent method as the rounding error is relatively smaller on the larger percent values the method uses.
 
@@
 readd(last_chance_plot)
+```
+
+The precision of both methods for estimating the count sum from percent data depend on count sum and on the precision to which the percent are recorded (Fig. \@ref(fig:percent-est-plot).
+In general, the direct search method is more precise and can give useful estimates when the assemblages have no singletons. 
+It can fail if data have been incorrectly rounded, for example by truncation. 
+If many digits are given, ignoring the last can solve this.
+
+```{r percent-est-plot, fig.cap = "Accuracy of estimates of the count sum calculated with two different methods with percent data have been rounded to different numbers of digits.
+Percent data are simulated by resampling a bird data route with 30 species to different count sums.", fig.width = 80/25.4, fig.height = 80/25.4}
+readd(percent_rounding_plot)
 ```
@@
 It seems likely that the data were converted to percent, rounded and then back-transformed to counts.
-For most purposes, such back-transformed or digitised data are perfectly adequate.
+For most purposes, such back-transformed or digitised data are perfectly adequate, but the Chao 1 richness estimator calculated using the number of singletons and doubletons, would fail [@Chao1987].
 

### 2022-12-31 run.R
adds visnetwork

@@ -8,3 +8,8 @@
 targets::tar_make()
 # targets::tar_make_clustermq(workers = 2) # nolint
 # targets::tar_make_future(workers = 2) # nolint
+
+
+targets::tar_visnetwork()
+
+

### 2022-12-31 manuscript/count_check_MS.qmd
rmd -> qmd

--- 2022-12-31 manuscript/count_check_MS.Rmd
@@
 Several assemblage datasets that fail these tests, and potentially have misreported counts or count sums, are identified.
-journal: "Journal of Quaternary Science"
-date: "`r format(Sys.Date(), '%d %B %Y')`"
+date: today
 header-includes:
@@
    - \newcommand{\elandscape}{\end{landscape}}   
-output:
-  bookdown::pdf_book:
-    base_format: 
-      rticles::elsevier_article
-    number_sections: true
+format: pdf
 documentclass: article
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = TRUE, message = TRUE, error = TRUE)
+```{r}
+#| label: setup
+#| include: false
 
 # Packages loaded for convenience when knitting without running r_make
-library("drake")
 library("tidyverse")
@@
 
-loadd(bird_singleton_summary, bird_summ)
-loadd(testate_summ)
-loadd(last_chance_summ)
-loadd(pollen_summ, hershop_pollen_summ, wildhorse, pollen_summ1, pollen_summ2, pollen3, pollen1, pollen2)
-loadd(owens_summ1, owens_summ, owens_est_summ)
-loadd(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ, chironomid2_summ)
+tar_load(bird_singleton_summary, bird_summ)
+tar_load(testate_summ)
+tar_load(last_chance_summ)
+tar_load(pollen_summ, hershop_pollen_summ, wildhorse, pollen_summ1, pollen_summ2, pollen3, pollen1, pollen2)
+tar_load(owens_summ1, owens_summ, owens_est_summ)
+tar_load(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ, chironomid2_summ)
 
@@
 4) The North American testate amoeba training set [@AMESBURY2018].
-5) Fossil pollen assemblage from the Neotoma database [@neotoma_2018] downloaded on `r readd(pollen_download_date)`.
+5) Fossil pollen assemblage from the Neotoma database [@neotoma_2018] downloaded on `r tar_read(pollen_download_date)`.
 
@@
 
-Across all the datasets, only a small minority of samples lack singletons, and even fewer have a greatest common divisor larger than one (Table \@ref(tab:multitaxa)).
-Most of the samples without singletons, or with a greatest common divisor larger than one, have low richness (Fig. \@ref(fig:richness-singleton-plot)).
+Across all the datasets, only a small minority of samples lack singletons, and even fewer have a greatest common divisor larger than one (@tbl-multitaxa).
+Most of the samples without singletons, or with a greatest common divisor larger than one, have low richness (@fig-richness-singleton-plot).
 
-Although the proportion of pollen assemblages that lack singletons or have a greatest common divisor of one are low, such assemblages are not randomly distributed among the `r pollen_summ$n_datasets %>% format()` pollen datasets.
+Although the proportion of pollen assemblages that lack singletons or have a greatest common divisor of one are low, such assemblages are not randomly distributed among the `r pollen_summ$n_datasets |> format()` pollen datasets.
 Half of the assemblages without singletons are in just `r pollen_summ$half_no_sing` datasets.
 `r Words(pollen_summ$half_gcd)` datasets contain half of the assemblages with a greatest common divisor above one and only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
-Some pollen datasets with unexpected properties are discussed in section \@ref(unexpected-count-data).
+Some pollen datasets with unexpected properties are discussed in section @sec-unexpected-count-data.
 
-```{r multitaxa}
-readd(count_summary) %>% 
+```{r}
+#| label: tbl-multitaxa
+#| tbl-cap: Properties of different data sets.
+tar_read(count_summary) |> 
   #manually escape % as cannot use kbl(escape = TRUE) with linebreak
-  mutate(across(where(is.character), str_replace, pattern = "%", replacement = "\\\\%")) %>% 
+  mutate(across(where(is.character), str_replace, pattern = "%", replacement = "\\\\%")) |> 
   kbl(x = ., booktabs = TRUE, escape = FALSE,
-      caption = "Properties of different data sets", 
       col.names = linebreak(str_wrap(names(.), width = 10))
-    ) %>% 
+    ) |> 
   kable_styling(latex_options = c("basic", "hold_position"))
@@
 
-```{r richness-singleton-plot, fig.cap = "Relationship betwen the presence of singletons and taxonomic richness for the different datasets.
-Samples with a greatest common divisor greater than one are shown in dark blue.
-Curve is fitted with a generalised additive model", fig.height = 120/24.5, fig.width=160/24.5}
-readd(richness_singleton_plot)
+```{r}
+#| label: fig-richness-singleton-plot
+#| fig.cap: Relationship betwen the presence of singletons and taxonomic richness for
+#|   the different datasets. Samples with a greatest common divisor greater than one
+#|   are shown in dark blue. Curve is fitted with a generalised additive model
+#| fig.height: !expr 120/24.5
+#| fig.width: !expr 160/24.5
+tar_read(richness_singleton_plot)
 ```
@@
 To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the bird data to progressively lower taxonomic resolutions.
-As richness declines, the proportion of counts without singletons increases (Table \@ref(tab:multitaxa)), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
+As richness declines, the proportion of counts without singletons increases (@tbl-multitaxa), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.
 Even when richness is low, only a small minority of counts have a greatest common divisor greater than one.
@@
 To test the sensitivity of the prevalence of counts without singletons to the count sum, I resample the bird counts with 400 or more individuals to smaller count sums using a multivariate hypergeometric distribution.
-At the order level, the proportion of counts without singletons decreases steeply until it reaches a minimum at when the count sum is about 200, thereafter it increases slowly (Fig.
-\@ref(fig:bird-recount)).
+At the order level, the proportion of counts without singletons decreases steeply until it reaches a minimum at when the count sum is about 200, thereafter it increases slowly (@fig-bird-recount).
 The prevalence of counts without singletons is high for small counts because it is possible that only the common taxa are counted; with larger counts, the chance of counting a single individual of a rare taxon increases.
@@
 
-```{r bird-recount, fig.cap="Effect of count sum on the proportion of bird counts from the North American breeding bird survey without singleton at the order level.
-Results are the mean of ten trials with 95% confidence interval.", fig.width = 80/25.4, fig.height = 70/25.4}
-readd(bird_recount_singleton_plot)
+```{r}
+#| label: fig-bird-recount
+#| fig.cap: Effect of count sum on the proportion of bird counts from the North American
+#|   breeding bird survey without singleton at the order level. Results are the mean
+#|   of ten trials with 95% confidence interval.
+#| fig.width: !expr 80/25.4
+#| fig.height: !expr 70/25.4
+tar_read(bird_recount_singleton_plot)
 ```
@@
 The percentages are given to two decimal places: to account for rounding errors, the countSum package uses the smallest and largest values consistent with the reported percentage.
-With the minimum percent method the estimated count sums are, within error, either identical to the reported count sum or twice as much (Fig. \@ref(fig:last-chance-plot)).
+With the minimum percent method the estimated count sums are, within error, either identical to the reported count sum or twice as much (@fig-last-chance-plot).
 With the direct search method, the estimated count sums are all exactly twice the reported count sum.
@@
 
-```{r last-chance-plot, fig.cap = "Estimated and reported chironomid count sums by the minumum percent method (solid symbols with error bars) and the direct search method (red crosses) from Last Chance Lake [@Axford2017].
-Lines show the 1:1 (solid) and 2:1 (dashed) relationships.", fig.width = 80/25.4, fig.height = 80/25.4}
-readd(last_chance_plot)
+```{r}
+#| label: fig-last-chance-plot
+#| fig.cap: Estimated and reported chironomid count sums by the minumum percent method
+#|   (solid symbols with error bars) and the direct search method (red crosses) from
+#|   Last Chance Lake [@Axford2017]. Lines show the 1:1 (solid) and 2:1 (dashed) relationships.
+#| fig.width: !expr 80/25.4
+#| fig.height: !expr 80/25.4
+tar_read(last_chance_plot)
 ```
 
-The precision of both methods for estimating the count sum from percent data depend on count sum and on the precision to which the percent are recorded (Fig. \@ref(fig:percent-est-plot).
+The precision of both methods for estimating the count sum from percent data depend on count sum and on the precision to which the percent are recorded (@fig-percent-est-plot).
 In general, the direct search method is more precise and can give useful estimates when the assemblages have no singletons. 
@@
 
-```{r percent-est-plot, fig.cap = "Accuracy of estimates of the count sum calculated with two different methods with percent data have been rounded to different numbers of digits.
-Percent data are simulated by resampling a bird data route with 30 species to different count sums.", fig.width = 80/25.4, fig.height = 80/25.4}
-readd(percent_rounding_plot)
+```{r}
+#| label: fig-percent-est-plot
+#| fig.cap: Accuracy of estimates of the count sum calculated with two different methods
+#|   with percent data have been rounded to different numbers of digits. Percent data
+#|   are simulated by resampling a bird data route with 30 species to different count
+#|   sums.
+#| fig.width: !expr 80/25.4
+#| fig.height: !expr 80/25.4
+tar_read(percent_rounding_plot)
 ```
@@
 Some of these datasets are older datasets which may, despite the available metadata, have been back-transformed from percent data or digitised with the loss of rare taxa and precision.
-For example, all but `r hershop_pollen_summ$has_singleton %>% word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two).
+For example, all but `r hershop_pollen_summ$has_singleton |> word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two).
 The original data [@Patty1968] were counted to exactly 200 pollen grains and fern spores. 
@@
 The probability of having all counts divisible by three in such a taxonomically rich assemblage is 10^`r ceiling(- pollen2$max_taxa3 * log10(3))`.
-In addition to the assemblages with a greatest common divisor above one, this dataset includes several assemblages that have a higher proportion of counts divisible by two or three than is typical in assemblages from Neotoma (Fig.
-\@ref(fig:multiples-plot)).
+In addition to the assemblages with a greatest common divisor above one, this dataset includes several assemblages that have a higher proportion of counts divisible by two or three than is typical in assemblages from Neotoma (@fig-multiples-plot).
 It is unclear what processes could generate such counts.
 
-```{r multiples-plot, warning = FALSE, fig.height = 100/24.5, fig.width=180/24.5, fig.cap = "A) Distribution of the percent of counts that are divisible by _k_ in each pollen assemblage with at least ten taxa.
-The vertical dashed line represents the 99.5% percentile.
-B) Percent of count in each assemblage in dataset pollen2 that are divisible by _k_ against depth.
-The dashed lines are the quantiles shown in part A.
-", message = FALSE}
-readd(multiples_plot)
+```{r}
+#| label: fig-multiples-plot
+#| warning: false
+#| fig.height: !expr 100/24.5
+#| fig.width: !expr 180/24.5
+#| fig.cap: 'A) Distribution of the percent of counts that are divisible by _k_ in each
+#|   pollen assemblage with at least ten taxa. The vertical dashed line represents the
+#|   99.5% percentile. B) Percent of count in each assemblage in dataset pollen2 that
+#|   are divisible by _k_ against depth. The dashed lines are the quantiles shown in
+#|   part A. '
+#| message: false
+tar_read(multiples_plot)
 ```
 
-In dataset pollen3, the greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three (Table \@ref(tab:unexpected-table)). 
+In dataset pollen3, the greatest common divisor is three for all assemblages: all `r pollen3$n_count` counts in the dataset are divisible by three (@tbl-unexpected-table). 
 For a dataset this large, the probability of having all counts divisible by three by chance is 3^{-`r pollen3$n_count`}^ or 10^`r ceiling(- pollen3$n_count * log10(3))`^.
@@
 
-Many assemblages in datasets pollen4 and pollen5 lack singletons and have a greatest common divisor greater than one (Table \@ref(tab:unexpected-table)).
+Many assemblages in datasets pollen4 and pollen5 lack singletons and have a greatest common divisor greater than one (@tbl-unexpected-table).
 For most of both datasets, which come from the same paper and match the published pollen stratigraphies, assemblages with and without singletons are arranged alternately.
@@
 The intervening assemblages did not exist in the original data. 
-In the published and archived data, they had been interpolated from their neighbours (Fig. \@ref(fig:interpolated-plot)).
+In the published and archived data, they had been interpolated from their neighbours (@fig-interpolated-plot).
 The effect of these manipulations is to increase the resolution of the pollen stratigraphy.
 
-```{r unexpected-table}
-readd(unexpected_table) %>% knitr::kable(caption = "The number of assemblages, median number of taxa and count sum, percent of assemblages with a greatest common divisor (GCD) of one, in five anonymised pollen datasets.")
+```{r}
+#| label: tbl-unexpected-table
+#| tbl-cap: The number of assemblages, median number of taxa and count sum, percent of assemblages with a greatest common divisor (GCD) of one, in five anonymised pollen datasets.
+tar_read(unexpected_table) |> knitr::kable()
 ```
 
-```{r interpolated-plot, warning = FALSE, fig.height = 100/24.5, fig.width=180/24.5, fig.cap="A) Plot of selected taxa by depth from dataset pollen2.
-Counts with blue triangles are from assemblages with a greatest common divisor of one.
-These have been linearly interpolated between the adjacent counts from assemblages with a greatest common divisor of two (red circles).
-B) Percent of taxa in each assemblage where the count is the mean of the counts in the adjacent assemblages, excluding runs of zero counts.
-Most of the assemblages with a greatest common divisor of one are exactly linearly interpolated between adjacent assemblages.
-The main exception is the assemblage at 210 cm which has a single singleton and otherwise has a greatest common divisor of two, other exceptions have a few extra singletons not predicted from the adjacent assembages.", message = FALSE}
-readd(pollen4_plot)
+```{r}
+#| label: fig-interpolated-plot
+#| warning: false
+#| fig.height: !expr 100/24.5
+#| fig.width: !expr 180/24.5
+#| fig.cap: A) Plot of selected taxa by depth from dataset pollen2. Counts with blue
+#|   triangles are from assemblages with a greatest common divisor of one. These have
+#|   been linearly interpolated between the adjacent counts from assemblages with a greatest
+#|   common divisor of two (red circles). B) Percent of taxa in each assemblage where
+#|   the count is the mean of the counts in the adjacent assemblages, excluding runs
+#|   of zero counts. Most of the assemblages with a greatest common divisor of one are
+#|   exactly linearly interpolated between adjacent assemblages. The main exception is
+#|   the assemblage at 210 cm which has a single singleton and otherwise has a greatest
+#|   common divisor of two, other exceptions have a few extra singletons not predicted
+#|   from the adjacent assembages.
+#| message: false
+tar_read(pollen4_plot)
 ```
@@
 ## Unexpected percent data
-```{r line169}
+```{r}
+#| label: line169
 
@@
 The associated paper reports that at least 400 valves were counted from each assemblage, which means that singletons should have a relative abundance of 0.25% or less.
-However, the rarest taxon in two assemblages have a relative abundance of `r diatom1_summ$maxpc`%, and the relative abundance of all other taxa in these assemblages are, within rounding error, integer multiples of this`r diatom1_summ$prop_integer %>% insist(.
-== 1) %>% invisible()`.
+However, the rarest taxon in two assemblages have a relative abundance of `r diatom1_summ$maxpc`%, and the relative abundance of all other taxa in these assemblages are, within rounding error, integer multiples of this`r diatom1_summ$prop_integer |> insist(.
+== 1) |> invisible()`.
 If the count sums actually are at least 400, this would imply that each species count in these assemblages is an integer multiple of at least `r diatom1_summ$integer_mult`.
@@
 Alternatively, the actual count sum for these assemblages could be as low as `r diatom1_summ$low_n` valves.
-In total, `r diatom1_summ$n_low_n %>% words()` assemblages appear to have a count sum below 400 valves. 
+In total, `r diatom1_summ$n_low_n |> words()` assemblages appear to have a count sum below 400 valves. 
 The results section of the paper acknowledges that counts were below the standard 400 valves for part of the Late Glacial.
@@
 
-```{r line178}
+```{r}
+#| label: line178
 
@@
 The associated paper reports that the diatom counts included 400–500 valves per assemblage except for four diatom-poor assemblages with at least 100 valves.
-The direct search method estimates that `r diatom2_summ$n400 %>% words()` assemblages have a count sum of less than 400, and `r diatom2_summ$n100 %>% words()` have less than 100.
-The direct search and minimum percent methods agree, within rounding error, `r diatom2_summ$meth_agree %>% insist(isTRUE(.)) %>% invisible()`on the estimated count sums, implying that either all assemblages have singletons and count sums are low in some assemblages, or that the greatest common divisor of the raw counts is greater than one.
-This means that for the assemblage where the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc %>% words()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.
+The direct search method estimates that `r diatom2_summ$n400 |> words()` assemblages have a count sum of less than 400, and `r diatom2_summ$n100 |> words()` have less than 100.
+The direct search and minimum percent methods agree, within rounding error, `r diatom2_summ$meth_agree |> insist(isTRUE(.)) |> invisible()`on the estimated count sums, implying that either all assemblages have singletons and count sums are low in some assemblages, or that the greatest common divisor of the raw counts is greater than one.
+This means that for the assemblage where the minimum percent is `r diatom2_summ$maxpc`, which has `r diatom2_summ$rich_maxpc |> words()` taxa, all counts would need to be multiples of `r diatom2_summ$integer_mult` for the count sum to be at least 100.
 
 Dataset chironomid1 has `r chironomid1_summ$nsamples` assemblages.
-Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 %>% words` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
-These assemblages are diverse (richness between `r chironomid1_summ$richness %>% paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.
+Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 |> words` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
+These assemblages are diverse (richness between `r chironomid1_summ$richness |> paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.
 
@@
 The associated paper reports that assemblages with count sums below 100 were omitted.
-The direct search method estimates that `r marine1_summ$n100 %>% words()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa %>% words()` mono-specific assemblage), with count sums as low as `r marine1_summ$est_n_min`.
+The direct search method estimates that `r marine1_summ$n100 |> words()` assemblages have count sums below 100 (this excludes `r marine1_summ$n_one_taxa |> words()` mono-specific assemblage), with count sums as low as `r marine1_summ$est_n_min`.
 With this dataset, the direct search and the minimum percent method give identical results for some assemblages but highly divergent estimates for others, with the direct search method sometimes giving estimates of several thousand, in one case over two orders of magnitude higher than the minimum percent method.
@@
 
-```{r}
-#set digits to default
-options(digits = 7)
-```
-
 # References{-}

### 2022-12-31 R/percent_rounding_plan.R
to targets. Changes route ID as previous was missing

@@ -1,37 +1,40 @@
+# effect of rounding on precision of estimates
 
-percent_rounding_plan <- drake_plan(
-  percent_rounding_sim = {
+simulate_percent_rounding <- function(bird_download) {
     test_data <- bird_download |> 
-      filter(RouteDataID == 6169900) |> 
-      select(spp = AOU, SpeciesTotal) |> 
+      filter(RouteDataID == 6168100) |> 
+      select(AOU, SpeciesTotal) |> 
       uncount(SpeciesTotal)
     
-    counts <- map_dfr(.x = 50:1000, 
+    counts <- map(.x = 50:1000, 
             ~ test_data |> 
               slice_sample(n = .x) |> 
-              count(spp) |> 
+              count(AOU) |> 
               mutate(percent = n/.x * 100, count_sum = .x)
-            )
+            ) |> 
+      list_rbind()
   
-    1:3 %>% 
-      set_names() %>% 
-      map_dfr(~{
-        mutate(counts, percent = round(percent, .x)) %>% 
-        estimate_n(percent_col = "percent", taxon_col = "spp", 
-                   ID_cols = "count_sum", digits = .x, nmax = 1500) %>% 
-          select(-direct_search, -data, -n_taxa) %>% 
+    1:3 |> 
+      set_names() |> 
+      map(~{
+        mutate(counts, percent = round(percent, .x)) |> 
+        estimate_n(percent_col = "percent", taxon_col = "AOU", 
+                   ID_cols = "count_sum", digits = .x, nmax = 1500) |> 
+          select(-direct_search, -data, -n_taxa) |> 
           unnest(cols = direct_search_est)
-      }, .id = "round")
-  },
+      }) |> 
+      list_rbind(names_to = "round")
+  }
   
-  percent_rounding_plot = percent_rounding_sim %>% 
-    pivot_longer(c(est_n_direct, est_n_minpc), names_to = "method", values_to = "estimate") %>% 
-    mutate(method = factor(method, levels = c("est_n_minpc", "est_n_direct"), labels = c("Minimum percent", "Direct search"))) %>% 
-    ggplot(aes(x = count_sum, y = round(estimate) - count_sum, colour = round)) +
+plot_percent_rounding <- function(percent_rounding_sim){
+  percent_rounding_sim |> 
+    pivot_longer(c(est_n_direct, est_n_minpc), names_to = "method", values_to = "estimate") |> 
[…]

### 2023-01-01 make_manuscript.R
unneded drake file

@@ -1,19 +1,0 @@
-#Use renv to install correct version of all necessary package
-#renv::restore()
-
-#load packages
-library("drake")
-
-plan <- "plan_count_check_ms_with_drake.R"
-
-#make ms
-r_make(source = plan) # Build the right things.
-drake_failed()
-
-if(length(drake_failed()) == 0){
-  fs::file_show("Rmd/count_check_MS.pdf")#display pdf
-}
-
-#show dependency graph
-r_vis_drake_graph(source = plan, targets_only = TRUE, main = "count ms dependency graph")
-

### 2023-01-01 R/download_diatoms.R
diatom to targets

@@ -1,114 +1,107 @@
 #diatoms
-diatom_plan <- drake_plan(
   #### MOLTEN ####
   
-  molten_download = target({
+download_molten <- function() {
       sqlite_file <- "data/define.sqlite"
       download.file(
         url = "https://github.com/richardjtelford/kustkarnor/blob/master/data/define.sqlite?raw=true", 
         destfile = sqlite_file)
       sqlite_file
-    },
-    format = "file"),
-  molten_data = {
-    con <- DBI::dbConnect(RSQLite::SQLite(), dbname = molten_download)
+    }
+  
+load_molten_data <- function(molten_file) {
+    con <- DBI::dbConnect(RSQLite::SQLite(), dbname = molten_file)
     
-    tbl(con, "counts") %>%
+    tbl(con, "counts") |>
       #merge synonyms
-      left_join(tbl(con, "Synonyms"), by = c("taxonCode" = "synonymCode")) %>%
-      mutate(taxonCode = coalesce(correctCode, taxonCode)) %>%
+      left_join(tbl(con, "Synonyms"), by = c("taxonCode" = "synonymCode")) |>
+      mutate(taxonCode = coalesce(correctCode, taxonCode)) |>
       #get country code for merges & drop samples with no chem - "RIB16102b"
-      inner_join(tbl(con, "fchem") %>% select(siteId, countryId), by  = "siteId") %>%
-      left_join(tbl(con, "Merges"), by = c("taxonCode" = "oldtaxonCode", "countryId")) %>%
+      inner_join(tbl(con, "fchem") |> select(siteId, countryId), by  = "siteId") |>
+      left_join(tbl(con, "Merges"), by = c("taxonCode" = "oldtaxonCode", "countryId")) |>
       mutate(
         taxonCode = coalesce(mergedTaxonCode, taxonCode)
-      ) %>%
+      ) |>
       #remove excluded taxa
-      anti_join(tbl(con, "ExcludedTaxa"), by = "taxonCode") %>%
+      anti_join(tbl(con, "ExcludedTaxa"), by = "taxonCode") |>
       #select required columns
-      select(siteId, taxonCode, count) %>%
+      select(siteId, taxonCode, count) |>
       #sum merged taxa
-      group_by(siteId, taxonCode) %>%
-      collect() %>%
-      assert(not_na, everything()) %>%
+      group_by(siteId, taxonCode) |>
+      collect() |>
[…]

### 2023-01-01 R/download_marine.R
fix

@@ -33,7 +33,7 @@ estimate_marine1_n <- function(marine1) {
 }
 
   #summarise
-summarise_marine1 <- function(marine1_est_n){
+summarise_marine1 <- function(marine1, marine1_est_n){
   list(
     nsamples = nrow(marine1),
     median_taxa = median(marine1_est_n$n_taxa),

### 2023-01-01 R/download_chironomids.R
chironomids to targets

@@ -1,91 +1,114 @@
-#chironomids
+# chironomids
 
-chironomid_plan <- drake_plan(
-  #data for Last Chance Lake
-  last_chance = read.table("https://www1.ncdc.noaa.gov/pub/data/paleo/insecta/chironomidae/northamerica/greenland/last-chance2017chironomid.txt", header = TRUE),
-  
-  last_chance_n = last_chance %>% 
-    pivot_longer(cols = c(-age_calBP, -totcaps), names_to = "taxon", values_to = "percent") %>% 
-    filter(percent > 0) %>% 
-    estimate_n(ID_cols = c("age_calBP", "totcaps"), digits = 2),
-  
-  last_chance_direct_search_plot = last_chance_n %>% 
-    filter(age_calBP == 1099) %>% 
-    select(age_calBP, totcaps, direct_search) %>% 
-    unnest(direct_search) %>% 
-    filter(n <= 300) %>% 
-    ggplot(aes(x  = n, y = score)) + 
-    geom_point() +
-    labs(x = "Putative count sum", y = "Score"),
-  
-  last_chance_plot = last_chance_n %>% 
-    unnest(direct_search_est) %>%   
-    ggplot(aes(x = totcaps, y = est_n_minpc)) + 
-    geom_abline(aes(slope = sl, intercept = 0, linetype = sf), data = tibble(sl = 1:2, sf = factor(sl)), show.legend = FALSE, colour = "grey50") +
-    geom_point() +
-    geom_errorbar(aes(ymax = est_max_minpc, ymin = est_min_minpc)) +
-    geom_point(mapping = aes(x = totcaps, y = est_n_direct), shape = "x", colour = "red") +
-    coord_equal() +
-    labs(x = "Reported count sum", y = "Estimated count sum"),
-  
-  last_chance_summ = list(
+
+# data for Last Chance Lake
+download_last_chance <- function() {
+  last_chance <- read.table(
+    file = "https://www1.ncdc.noaa.gov/pub/data/paleo/insecta/chironomidae/northamerica/greenland/last-chance2017chironomid.txt",
+    header = TRUE
+  )
+
+  last_chance
+}
+
+# estimate n
+estimate_last_chance_n <- function(last_chance) {
+  print(class(last_chance))
+  print(filter)
+  last_chance <- last_chance |>
+    pivot_longer(
+      cols = c(-age_calBP, -totcaps),
+      names_to = "taxon", values_to = "percent"
+    ) |>
[…]

### 2023-01-01 run.R
shows pdf if no errors

@@ -12,4 +12,6 @@ targets::tar_make()
 
 targets::tar_visnetwork()
 
-
+if(length(drake_errored()) == 0){
+  fs::file_show("manuscript/count_check_MS.pdf") # display pdf is no errors
+}

### 2023-01-01 manuscript/count_check_MS.qmd
tar_load

@@
 csl: extra/elsevier-harvard_rjt.csl
+execute: 
+  echo: false
+  error: false
 ---
@@
 library("tidyverse")
-library("rjt.misc")
+#library("rjt.misc")
 
@@
 
-tar_load(bird_singleton_summary, bird_summ)
-tar_load(testate_summ)
-tar_load(last_chance_summ)
-tar_load(pollen_summ, hershop_pollen_summ, wildhorse, pollen_summ1, pollen_summ2, pollen3, pollen1, pollen2)
-tar_load(owens_summ1, owens_summ, owens_est_summ)
-tar_load(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ, chironomid2_summ)
+tar_load("summarised_bird_singletons")
+tar_load("summarised_bird")
+tar_load("testate_summ")
+tar_load("last_chance_summ")
+tar_load("pollen_summ")
+tar_load("hershop_pollen_summ")
+tar_load("wildhorse")
+tar_load("pollen_summ1")
+tar_load("pollen_summ2")
+tar_load("pollen3")
+tar_load("pollen1")
+tar_load("pollen2")
+tar_load("owens_summ1")
+tar_load("owens_summ")
+tar_load("owens_est_summ")
+tar_load("diatom1_summ")
+tar_load("diatom2_summ")
+tar_load("chironomid1_summ")
+tar_load("marine1_summ")
+tar_load("chironomid2_summ")
 
@@
 #| fig.height: !expr 70/25.4
-tar_read(bird_recount_singleton_plot)
+tar_read(recounted_bird_singletons_plot)
 ```
@@
 Dataset chironomid1 has `r chironomid1_summ$nsamples` assemblages.
-Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 |> words` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
+Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 |> words()` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
 These assemblages are diverse (richness between `r chironomid1_summ$richness |> paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.

### 2023-01-01 _targets.R
diatoms and chironomids

@@ -10,7 +10,7 @@ library("tarchetypes")
 # Set target options:
 tar_option_set(
   packages = c("tidyverse", "patchwork", "readxl", "countSum", "assertr"), 
-  impports = "countSum"
+  imports = "countSum",
   format = "rds" # default storage format
   # Set other options as needed.
 )
@@ -59,11 +59,11 @@ list(
   ),
   tar_target(
     name = secrets,
-    command = get_secrets(secrets_file, has_passord)
+    command = get_secrets(secrets_file, has_password)
   ),
   tar_target(
     name = neotoma_secrets,
-    command = get_neotoma_secrets(neotoma_secrets_file, has_passord)
+    command = get_neotoma_secrets(neotoma_secrets_file, has_password)
   ),
   
   # birds ####
@@ -140,6 +140,39 @@ list(
     command = summarise_counts(molten_data)
   ),
 
+ ## diatoms ####
+## diatom1
+  tar_target(
+    # download and import
+    name = diatom1,
+    command = download_diatom1(has_password, secrets)
+  ),
+  tar_target(
+    # estimate n
+    name = diatom1_est_n,
+    command = estimate_diatom_n(diatom1, digits = 2)
+  ),
+  tar_target(
+    #summarise diatom1
+    name = diatom1_summ,
+    command = summarise_diatom(diatom1_est_n, diatom1),
+  ),
+  ## diatom2
+  tar_target(
+    # download and import
+    name = diatom2,
+    command = download_diatom2(has_password, secrets)
+  ),
+  tar_target(
+    # estimate n
+    name = diatom2_est_n,
+    command = estimate_diatom_n(diatom2, digits = 2)
+  ),
+  tar_target(
+    #summarise diatom1
+    name = diatom2_summ,
+    command = summarise_diatom(diatom2_est_n, diatom2),
+  ),
   # last chance chronomids ####
   tar_target(
    # download and import last chance
@@ -173,9 +206,21 @@ list(
   ),
   tar_target(
     name = marine1_summ,
-    command = summarise_marine1(marine1_est_n)
+    command = summarise_marine1(marine1, marine1_est_n)
   ),
-
+## chironomid1 ####
+  tar_target(
+    name = chironomid1,
+    command = download_chironomid1(has_password, secrets)
+  ),
+  tar_target(
+    name = chironomid1_est_n,
+    command = estimate_chironomid1_n(chironomid1)
+  ),
+  tar_target(
[…]

### 2023-06-26 _targets.R
pollen & testate import

@@ -9,11 +9,13 @@ library("tarchetypes")
 
 # Set target options:
 tar_option_set(
-  packages = c("tidyverse", "patchwork", "readxl", "countSum", "assertr"), 
+  packages = c("tidyverse", "patchwork", "readxl", "countSum", "assertr", "neotoma2", "conflicted", "glue"), 
   imports = "countSum",
   format = "rds" # default storage format
   # Set other options as needed.
 )
+
+conflicted::conflict_prefer_all(winner = "dplyr", quiet = TRUE)
 
 ## askpass
 options(askpass = function(x) readLines("pw.txt"))
@@ -222,10 +224,65 @@ list(
     command = summarise_chironomid1(chironomid1, chironomid1_est_n)
   ),
 
+### pollen ####
+  tar_target(
+    name = pollen_datasets,
+    command = neotoma2::get_datasets(
+      datasettype = "pollen", 
+      all_data = TRUE)
+  ),
+  tar_target(
+    name = pollen_data,
+    command = get_download_batch(pollen_datasets)
+  ),
+  tar_target(
+    name = pollen,
+    command = get_pollen_table(pollen_data)
+  ),
+  tar_target(
+    name = pollen_assem_res,
+    command = summarise_counts(pollen)
+  ),
+  tar_target(
+    name = pollen_dataset_res,
+    command = summarise_singletons(pollen_assem_res)
+  ),
+  tar_target(
+    name = pollen_results,
+    command = make_pollen_results(pollen_assem_res, pollen_dataset_res)
+  ),
+  
+### forams ####
+  tar_target(
+    name = forams,
+    command = import_forams()
+  ),
+  tar_target(
+    name = foram_assem_res,
+    command = summarise_counts(forams)
+  ),
+   
+## testate ####
+  tar_target(
+    name = testate_datasets,
+    command = neotoma2::get_datasets(
+      datasettype = "testate amoebae surface sample", 
+      all_data = TRUE)
+  ),
+  tar_target(
+    name = testate_data,
+    command = get_download_batch(testate_datasets, offset = 10)
+  ),
+  tar_target(
+    name = testate,
+    command = get_testate_table(testate_data)
+  ),
+
+
 
   ## manuscript
   tar_quarto(
     name = manuscript,
-    path = "manuscript/count_check_MS.qmd"
+    path = "count_check_MS.qmd"
   )
 )

### 2023-06-26 R/download_diatoms.R
archived data has new col name

@@ -95,7 +95,7 @@ summarise_diatom <- function(est_n, data){
       filter(dataset == "diatom2") |>
       pull(secret)
     read_delim(target, delim = "\t", skip = 217) |>
-      select(-(Latitude:`Depth [m]`)) |>
+      select(-(Latitude:`Depth sed [m]`)) |>
       rename(sampleID = Event) |> 
       pivot_longer(cols = -sampleID, names_to = "taxon", values_to = "percent") |>
       filter(percent > 0)

### 2023-06-26 R/general_functions.R
batch download fun

@@ -1,6 +1,18 @@
+# Download neotoma data in batches
+
+get_download_batch <- function(datasets, offset = 50) {
+  starts <- seq(1, length(datasets), offset)
+  downloads <- map(starts, \(x) {
+    get_downloads(datasets[x:min(x + offset - 1, length(datasets))])
+  })
+  reduce(downloads, .f = c)
+}
+
+
+
 ##  calculate GCD and singleton data per sample
 
-summarise_counts <- function(df){
+summarise_counts <- function(df) {
   df |> 
   summarise(
     count_sum = sum(count), 
@@ -15,7 +27,7 @@ summarise_counts <- function(df){
 
 #only find GCD where no singletons for speed
 # When n_taxa == 1, mGCD will give an error, so return min
-fast_mGCD <- function(n_taxa, count, n_singletons, min){
+fast_mGCD <- function(n_taxa, count, n_singletons, min) {
   if (n_taxa == 1){# mGCD will crash. Result is trivially min
     return(min)
   }
@@ -29,7 +41,7 @@ fast_mGCD <- function(n_taxa, count, n_singletons, min){
 
 ## summarise GCD and singleton data per data set
 
-summarise_singletons <- function(df){
+summarise_singletons <- function(df) {
   df |> 
   filter(count_sum >= 50, n_taxa > 1) |> 
   summarise(
@@ -43,3 +55,8 @@ summarise_singletons <- function(df){
   mutate(across(c(`Without singletons`, `GCD > 1`), as.character))
 
 }
+
+floor1000 <- function(x, base = 1000) {
+  floor(x / base) * base
+}
+

### 2023-06-26 R/multitaxa_funtions.R
functions for multiple taxa groups

@@ -1,0 +1,64 @@
+summarise_all_counts = function(bird_singletons, molten_singletons, foram_summ, foram_assem_res, pollen_assem_res){
+  bind_rows(
+    `Birds - species` = bird_singletons |> 
+      ungroup() |>
+      filter(taxonomic_level == "species") |> 
+      summarise_singletons(),
+    `Birds - genus` = bird_singletons |> 
+      ungroup() |>
+      filter(taxonomic_level == "genus") |> 
+      summarise_singletons(),
+    `Birds - family` = bird_singletons |> 
+      ungroup() |>
+      filter(taxonomic_level == "family") |> 
+      summarise_singletons(),
+    `Birds - order` = bird_singletons |> 
+      ungroup() |>
+      filter(taxonomic_level == "order") |> 
+      summarise_singletons(),
+    `Coastal diatoms` = molten_singletons |> 
+      summarise_singletons(),
+    `Planktic foraminifera` = foram_assem_res |> 
+      summarise_singletons(), 
+    `Testate amoebae` = testate_summ1 |> 
+      summarise_singletons(), 
+    `Pollen` = pollen_assem_res |> 
+      ungroup() |> 
+      summarise_singletons(),
+    .id = "Dataset"
+  )
+}
+
+
+make_richness_singleton_plot = function(bird_singletons, molten_singletons, foram_assem_res, testate_summ1, pollen_assem_res) {
+  bind_rows(
+  `Birds - species` = bird_singletons |> 
+    ungroup() |>
+    filter(taxonomic_level == "species"),
+  `Birds - order` = bird_singletons |> 
+    ungroup() |>
+    filter(taxonomic_level == "order"),
+  `Coastal diatoms` = molten_singletons,
+  `Planktic foraminifera` = foram_assem_res, 
+  `Testate amoebae` = testate_summ1, 
+  `Pollen` = pollen_assem_res |> ungroup(),
+  .id = "dataset"
+)  |> 
+  filter(n_taxa != 1) |> # remove monospecific samples - only birds mostly at order level
+  mutate(
+    dataset = factor(
+      x = dataset, 
+      levels = c("Birds - species", "Birds - order", "Coastal diatoms", 
+                 "Planktic foraminifera", "Testate amoebae", "Pollen")),
+    singletons = if_else(n_singletons > 0, 1, 0)
+  ) |>
+  arrange(gcd) |> 
[…]

### 2023-06-26 count_check_MS.qmd
moved otherwise cannot find targets cache

--- 2023-01-01 manuscript/count_check_MS.qmd
@@
 classoption: a4paper
-bibliography: extra/countMS2.bib
-csl: extra/elsevier-harvard_rjt.csl
+bibliography: manuscript_extra/countMS2.bib
+csl: manuscript_extra/elsevier-harvard_rjt.csl
 execute: 
@@
 
-# Packages loaded for convenience when knitting without running r_make
+# Packages loading
+library("targets")
 library("tidyverse")
@@
 
-#packages only loaded in this file
-library("kableExtra")
+library("gt")
 library("english")
@@
 
-tar_load("summarised_bird_singletons")
-tar_load("summarised_bird")
-tar_load("testate_summ")
+# tar_load("summarised_bird_singletons")
+# tar_load("summarised_bird")
+# tar_load("testate_summ")
  tar_load("last_chance_summ")
-tar_load("pollen_summ")
+# tar_load("pollen_assem_res")
+# tar_load("pollen_dataset_res")
+ tar_load("pollen_results")
  tar_load("hershop_pollen_summ")
  tar_load("wildhorse")
-tar_load("pollen_summ1")
-tar_load("pollen_summ2")
  tar_load("pollen3")
@@
  tar_load("pollen2")
-tar_load("owens_summ1")
-tar_load("owens_summ")
-tar_load("owens_est_summ")
+# tar_load("owens_summ1")
+# tar_load("owens_summ")
+# tar_load("owens_est_summ")
  tar_load("diatom1_summ")
@@
  tar_load("marine1_summ")
-tar_load("chironomid2_summ")
+# tar_load("chironomid2_summ")
 
@@
 4) The North American testate amoeba training set [@AMESBURY2018].
-5) Fossil pollen assemblage from the Neotoma database [@neotoma_2018] downloaded on `r tar_read(pollen_download_date)`.
+5) Fossil pollen assemblage from the Neotoma database [@neotoma_2018] downloaded on `r targets::tar_meta(pollen_data) |> pull(time) |> format("%d %B %Y")`.
 
@@
 countSum v. `r packageVersion("countSum")` [@countSum], 
-neotoma v. `r packageVersion("neotoma")` [@neotoma], 
+neotoma2 v. `r packageVersion("neotoma2")` [@neotoma2], 
 bbsAssistant v. `r packageVersion("bbsAssistant")` [@bbsAssistant1], 
 and tidyverse v. `r packageVersion("tidyverse")` [@tidyverse].
-Packages drake v. `r packageVersion("drake")` [@drake], 
-bookdown v. `r packageVersion("bookdown")` [@bookdown1] and
+Packages targets v. `r packageVersion("targets")` [@targets], 
+quarto v. `r packageVersion("quarto")` [@bookdown1] and
 renv v. `r packageVersion("renv")` [@renv] were used to ensure reproducibility.
@@
 
-Although the proportion of pollen assemblages that lack singletons or have a greatest common divisor of one are low, such assemblages are not randomly distributed among the `r pollen_summ$n_datasets |> format()` pollen datasets.
-Half of the assemblages without singletons are in just `r pollen_summ$half_no_sing` datasets.
-`r Words(pollen_summ$half_gcd)` datasets contain half of the assemblages with a greatest common divisor above one and only `r pollen_summ$dataset_gcd1`% of the datasets have any such assemblages.
+Although the proportion of pollen assemblages that lack singletons or have a greatest common divisor of one are low, such assemblages are not randomly distributed among the `r pollen_results$n_datasets |> format()` pollen datasets.
+Half of the assemblages without singletons are in just `r pollen_results$half_no_sing` datasets.
+`r Words(pollen_results$half_gcd)` datasets contain half of the assemblages with a greatest common divisor above one and only `r pollen_results$dataset_gcd1`% of the datasets have any such assemblages.
 Some pollen datasets with unexpected properties are discussed in section @sec-unexpected-count-data.
@@
   mutate(across(where(is.character), str_replace, pattern = "%", replacement = "\\\\%")) |> 
-  kbl(x = ., booktabs = TRUE, escape = FALSE,
-      col.names = linebreak(str_wrap(names(.), width = 10))
-    ) |> 
-  kable_styling(latex_options = c("basic", "hold_position"))
+  gt(x = _)
+# , booktabs = TRUE, escape = FALSE,
+#       col.names = linebreak(str_wrap(names(.), width = 10))
+#     ) |> 
+#   kable_styling(latex_options = c("basic", "hold_position"))
 ```

### 2024-01-12 _targets.R
more targets

@@ -195,6 +195,10 @@ list(
   tar_target(
     name = last_chance_plot,
     command = plot_last_chance(last_chance_estimate_n)
+  ),  
+  tar_target(
+    name = last_chance_summ,
+    command = summarise_last_chance(last_chance_data)
   ),
 
   ## marine1
@@ -251,6 +255,12 @@ list(
     name = pollen_results,
     command = make_pollen_results(pollen_assem_res, pollen_dataset_res)
   ),
+
+## hershop
+  tar_target(
+    name = hershop_pollen,
+    command = process_hershop_pollen(pollen_data)
+  ),
   
 ### forams ####
   tar_target(
@@ -277,7 +287,24 @@ list(
     name = testate,
     command = get_testate_table(testate_data)
   ),
+  tar_target(
+    name = testate_assem_res,
+    command = summarise_counts(testate)
+  ),
 
+## multi taxa tables/figures
+  tar_target(
+    name = count_summary,
+    command = summarise_all_counts(bird_singletons, molten_singletons,
+                                   foram_summ, foram_assem_res,
+                                   testate_assem_res, pollen_assem_res) 
+  ),
+  tar_target(
+    name = richness_singleton_plot,
+    command = make_richness_singleton_plot(bird_singletons, molten_singletons,
+                                           foram_assem_res, testate_assem_res,
+                                           pollen_assem_res)
+  ),
 
 
   ## manuscript

### 2024-01-12 count_check_MS.qmd
gt

@@
  tar_load("pollen_results")
- tar_load("hershop_pollen_summ")
+ tar_load("hershop_pollen")
  tar_load("wildhorse")
@@
 tar_read(count_summary) |> 
-  #manually escape % as cannot use kbl(escape = TRUE) with linebreak
-  mutate(across(where(is.character), str_replace, pattern = "%", replacement = "\\\\%")) |> 
-  gt(x = _)
-# , booktabs = TRUE, escape = FALSE,
-#       col.names = linebreak(str_wrap(names(.), width = 10))
-#     ) |> 
-#   kable_styling(latex_options = c("basic", "hold_position"))
+  gt()
 ```
@@
 Some of these datasets are older datasets which may, despite the available metadata, have been back-transformed from percent data or digitised with the loss of rare taxa and precision.
-For example, all but `r hershop_pollen_summ$has_singleton |> word()` of the `r hershop_pollen_summ$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two).
+For example, all but `r hershop_pollen$has_singleton |> word()` of the `r hershop_pollen$n` assemblages in the archived pollen data from Hershop Bog, Texas have a greatest common divisor of two or more (the exception has a single singleton and is otherwise divisible by two).
 The original data [@Patty1968] were counted to exactly 200 pollen grains and fern spores. 
 It seems likely that the data were converted to percent, rounded and then back-transformed to counts.
-For most purposes, such back-transformed or digitised data are perfectly adequate, but the Chao 1 richness estimator calculated using the number of singletons and doubletons, would fail [@Chao1987].
+For most purposes, such back-transformed or digitised data are perfectly adequate, but the Chao 1 richness estimator [@Chao1987] calculated using the number of singletons and doubletons, would fail.
 

### 2024-01-15 R/download_pollen.R
update renv

@@ -8,7 +8,7 @@ get_pollen_table <- function(pollen_data) {
   map(start, \(x){
     pd <- pollen_data[[x:min(x + offset - 1, length(pollen_data))]] 
     pd <- pd |> 
-      neotoma2::filter(!datasetid %in% c(
+      filter(!datasetid %in% c( ##!neotoma2::
         55055, 15059, 25609, 39474, 54508, # percent 
         16209, 16210, 15696 # possible percent (not counts anyway)
       ))

### 2024-01-15 _targets.R

@@ -9,7 +9,7 @@ library("tarchetypes")
 
 # Set target options:
 tar_option_set(
-  packages = c("tidyverse", "patchwork", "readxl", "countSum", "assertr", "neotoma2", "conflicted", "glue"), 
+  packages = c("tidyverse", "patchwork", "readxl", "countSum", "assertr", "conflicted", "glue"), 
   imports = "countSum",
   format = "rds" # default storage format
   # Set other options as needed.
@@ -231,7 +231,7 @@ list(
 ### pollen ####
   tar_target(
     name = pollen_datasets,
-    command = neotoma2::get_datasets(
+    command = get_datasets(
       datasettype = "pollen", 
       all_data = TRUE)
   ),
@@ -275,7 +275,7 @@ list(
 ## testate ####
   tar_target(
     name = testate_datasets,
-    command = neotoma2::get_datasets(
+    command = get_datasets(
       datasettype = "testate amoebae surface sample", 
       all_data = TRUE)
   ),

### 2024-01-15 run.R

@@ -12,7 +12,7 @@ targets::tar_make()
 
 targets::tar_visnetwork()
 
-if(length(drake_errored()) == 0){
+if(TRUE){
   fs::file_show("manuscript/count_check_MS.pdf") # display pdf is no errors
 }
 

### 2024-03-01 _targets.R
adds bibliography

@@ -308,6 +308,26 @@ list(
 
 
   ## manuscript
+
+
+  tar_target(
+    # bibliography
+    name = raw_bibliography,
+    command = "manuscript_extra/countMS.bib",
+    format = "file" 
+  ), 
+  tar_target(
+    # bibliography
+    name = bibliography,
+    command = package_citations(
+          packages = c("extraDistr", "countSum", "numbers", "bbsAssistant",
+                       "targets", "tidyverse", "quarto", "renv"),
+          old_bib = raw_bibliography,
+          new_bib = "manuscript_extra/countMS2.bib"),
+    format = "file" 
+  ), 
+
+ 
   tar_quarto(
     name = manuscript,
     path = "count_check_MS.qmd"

### 2024-03-01 _targets_packages.R
update packages

@@ -1,7 +1,10 @@
 # Generated by targets::tar_renv(): do not edit by hand
 library(assertr)
+library(conflicted)
 library(countSum)
+library(glue)
 library(patchwork)
+library(quarto)
 library(readxl)
 library(tidyverse)
 library(visNetwork)

### 2024-03-01 run.R

@@ -16,4 +16,4 @@ if(TRUE){
   fs::file_show("manuscript/count_check_MS.pdf") # display pdf is no errors
 }
 
-targets::tar_renv(extras = c("visNetwork"))
+targets::tar_renv(extras = c("quarto", "visNetwork"))

### 2024-03-01 R/download_birds.R
update bbs data

@@ -2,44 +2,44 @@
 
 # download data
 download_bird <- function() {
-  sb_id <- "625f151ed34e85fa62b7f926"
-  bbs_dir <- "data/birds2022"
+  sb_id <- "64ad9c3dd34e70357a292cee"
+  bbs_dir <- "data/birds2023"
+  
+  # make dir if
+  if (!fs::dir_exists(bbs_dir)) {
+    fs::dir_create(bbs_dir)
+  }
   
   bbsAssistant::download_bbs_data(
     sb_id = sb_id,
     bbs_dir = bbs_dir, 
-    overwrite = TRUE
+    overwrite = FALSE
   )
-  # remove 3 lines from species list to 
-  # work around bug in read_fwf
 
-  sp_path <- file.path(bbs_dir, "SpeciesList.txt")
-  fs::file_copy(sp_path,  paste0(sp_path, 2), overwrite = TRUE)
-  sp_list <- readLines(sp_path)
-  write_lines(sp_list[-(1:3)], sp_path)
-  
-  bbs <- bbsAssistant::import_bbs_data(bbs_dir = bbs_dir, sb_id = sb_id)
-  #replace original
-  fs::file_copy(paste0(sp_path, 2), sp_path, overwrite = TRUE)
+  bbs <- read_csv(file.path(bbs_dir, "50-StopData.zip"))
+
 
   # select required columns
-  bbs$observations |>
+  bbs |>
     (\(x){
       mutate(x, SpeciesTotal = rowSums(select(x, matches("^Stop\\d{1,2}"))))
     })() |> 
+    mutate(AOU = as.integer(AOU)) |> 
     select(RouteDataID, AOU, Year, SpeciesTotal)
 }
   
 
 # load taxonomic data
 load_bird_species <- function(){ 
-  read_fwf("data/birds2022/SpeciesList.txt2", skip = 12) |> 
+  read_fwf("data/birds2023/SpeciesList.txt", skip = 10) |> 
     janitor::row_to_names(1) |> 
     slice(-1) |> 
     select(AOU, Order = ORDER, Family, Genus, Species) |> 
     mutate(
       AOU = as.integer(AOU),
-      Species = paste(Genus, Species))
+      Species = paste(Genus, Species)) |> 
+    # Add Northwestern Crow - now merged with American Crow but in data
+    add_row(AOU = 4890L, Order = "Passeriformes", Family = "Corvidae", Genus =  "Corvus", Species = "Corvus brachyrhynchos caurinus")
 }
 
 

### 2024-03-01 count_check_MS.qmd
update renv
new intro

@@
 ---
-title: "Tools for identifying unexpectedly low microfossil count sums"
+title: "Using the species abundance distribution to validate ecological and palaeoecological data"
 author:
@@
 abstract: |
-  Microfossil counts are a key data type in palaeoecology.
-Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so, and perhaps other aspects of the counts.
-This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions.
-Several assemblage datasets that fail these tests, and potentially have misreported counts or count sums, are identified.
+
+  Species community or assemblage counts are a key data type in ecology and palaeoecology. 
+  Sometimes, the data are archived as percentages.
+  Unfortunately, necessary metadata such as the count sum are not always archived with the percentages.
+  This paper uses insights from the ubiquitous shape of specied abundance distributions to develop methods to estimate the count sum of percentage data when it is missing.
+  The methods are demonstrated on a collection of archived percent data.   
+  The assumptions of the methods are tested using a large collection of ecological and palaeoecological count data.
+  Several datasets that fail the tests. 
+  These potentially have misreported data.
+  The consequences of misreported data for different types of analyses is discussed.
 date: today
@@
   error: false
+editor: 
+  markdown: 
+    wrap: sentence
 ---
@@
  tar_load("pollen2")
-# tar_load("owens_summ1")
-# tar_load("owens_summ")
-# tar_load("owens_est_summ")
  tar_load("diatom1_summ")
@@
 
-Six percent of papers published in _Molecular and Cellular Biology_ show evidence of inappropriate image duplication [@Bik2018] either due to error or, more rarely, misconduct.
-It would, especially after the spate of retractions in behavioural ecology [@Pennisi2021], be reckless to assume that the ecological and palaeoecological literature is devoid of errors and misconduct.
-The low rate of retractions in the ecological and geological literature [@Grieneisen2012] may partially reflect our limited ability to detect errors and misconduct rather than their prevalence.
+Counts of organisms are a key data type for ecology and palaeoecology.
+The sum of the counts is an important metric of data quality: larger count sums are associated with smaller uncertainties, both in relative abundance of taxa and any derived statistics such as transfer function reconstructions [@Heiri2001; @Payne2009].
+For raw count data, the count sum can be easily calculated, but some of the considerable amount of archived count data are only available as percentages.
+For some percent datasets, the count sum has not been archived and we might want to estimate it.
 
-Several numerical tools have been developed to identify questionable data.
-Deviations from the distribution of digits expected from the Newcomb-Benford law have been used to detect issues with scientific and financial data [@Barabesi2017].
-@Carlisle2017 identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data.
-@Brown2017 developed the granularity-related inconsistency of means (GRIM) test which assesses whether means of integer data are consistent with the reported sample size.
-However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data.
+While it is, in general, not possible to calculate the denominator used to calculate percentages, this paper aims to show that it is possible to estimate it for community or assemblage count data.
+Two insights make this possible.
+The first is that the raw counts are typically integers, giving a granularity [@Brown2017] to the percentage data.
+The second is that the raw counts can be represented by a species abundance distribution.
+The ubiquitous shape of species abundance distributions when plotted as the number of species against abundance is a hyperbolic curve with many rare and few abundant species [@McGill_2007].
+From this shape, which is central to ecological theory [@McGill_2007], two predictions can be made: that counts of an entire community or assemblage should have many rare taxa, typically including taxa occuring only once (hereafter singletons) and that the distribution of abundances should have a smooth and monotonic decline.
 
-A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often fifty for chironomids and several hundred for pollen and diatoms.
-This is important because larger count sums are associated with smaller uncertainties, both in relative abundance of taxa and derived statistics such as transfer function reconstructions [@Heiri2001; @Payne2009].
-There is a risk that count sums will be misreported.
-This could happen by mistake, especially with the use of secondary data.
-It could also happen knowingly, especially when microfossil preservation is poor or concentrations are low.
-This risk that count sums could be misreported is demonstrated by @Larocque-Tobler2016 who acknowledged that the chironomid count sums in @Larocque-Tobler2015 were not at least fifty head capsules as reported but as low as nineteen.
-@Telford2019 reports some other cases where count sums may be much lower than reported.
+This paper will use these insights to develop methods for estimating the count sum from percent data, and demonstrate them with some percent data.
+Then the paper tests the assumptions of the methods with a large collection of ecological and palaeoecological count data.
+This reveals that a some datasets have count data that differ markedly from the predictions based on species abundance distributions.
+These datasets might not accurately reflect the original community or assemblage.
 
-Given that count sums have been misreported, the possibility that other aspects of count data may be misreported needs to be considered.
+<!-- Six percent of papers published in *Molecular and Cellular Biology* show evidence of inappropriate image duplication [@Bik2018] either due to error or, more rarely, misconduct. -->
+<!-- It would, especially after the spate of retractions in behavioural ecology [@Pennisi2021], be reckless to assume that the ecological and palaeoecological literature is devoid of errors and misconduct.
+-->
+<!-- The low rate of retractions in the ecological and geological literature [@Grieneisen2012] may partially reflect our limited ability to detect errors and misconduct rather than their prevalence.
+-->
 
-The key insight that allows inference about community or assemblage count data is that they can be represented by a species abundance distribution.
-The ubiquitous shape of species abundance distributions when plotted as the number of species against abundance is a hyperbolic curve with many rare and few abundant species and is central to ecological theory [@McGill_2007].
-From this, two predictions can be made: that counts of an entire community or assemblage should have many rare taxa, and that the distribution of abundances should have a smooth and monotonic decline.
-Count data that differ markedly from these predictions might not accurately reflect the original community or assemblage.
+<!-- Several numerical tools have been developed to identify questionable data. -->
+<!-- Deviations from the distribution of digits expected from the Newcomb-Benford law have been used to detect issues with scientific and financial data [@Barabesi2017]. -->
+<!-- @Carlisle2017 identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data. -->
+<!-- @Brown2017 developed the granularity-related inconsistency of means (GRIM) test which assesses whether means of integer data are consistent with the reported sample size. -->
+<!-- However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data. -->
 
-This paper aims to develop simple methods to test the abundance of rare taxa and the smoothness of the species abundance distribution that can be used to flag cases where the count data might have been misreported.
-Next, as, regrettably, many palaeoecological data sets are archived as percent data, these tests are then extended to identify percent data where the count sum might be lower than reported.
-Finally, some complications and caveats are discussed.
+<!-- A common assertion in papers reporting microfossil assemblage data is that a minimum of *N* microfossils were counted, where *N* is often fifty for chironomids and several hundred for pollen and diatoms.
+-->
+
+<!-- There is a risk that count sums will be misreported. -->
+<!-- This could happen by mistake, especially with the use of secondary data. -->
+<!-- It could also happen knowingly, especially when microfossil preservation is poor or concentrations are low. -->
+<!-- This risk that count sums could be misreported is demonstrated by @Larocque-Tobler2016 who acknowledged that the chironomid count sums in @Larocque-Tobler2015 were not at least fifty head capsules as reported but as low as nineteen.
+-->
+<!-- @Telford2019 reports some other cases where count sums may be much lower than reported. -->
+
+<!-- Given that count sums have been misreported, the possibility that other aspects of count data may be misreported needs to be considered. -->
 
@@
 
+```{r}
+neotoma_date <- list.files("data/neotoma/", pattern = "sql$") |> 
+  str_extract("\\d{4}-\\d{1,2}-\\d{1,2}") |> 
+  ymd() |> 
+  max() |> 
+  format("%d %B %Y")
+```
+
 Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea, and other sources.
-A range of ecological and palaeoecological data were sought to allow for differences in typical count sums and species richness.
+A range of ecological and palaeoecological data were sought to allow for differences in typical count sums and species richness, and possible differences in species abundance distributions between taxonomic groups.
 
 Count data analysed include the 1) North American Breeding Bird survey [@bbsAssistant2], a very large, annual monitoring programme to track the status of North American bird populations.
-The survey reports all birds seen in point counts along ~40 km survey routes at the height of the breeding season. 
+The survey reports all birds seen in point counts along \~40 km survey routes at the height of the breeding season.
 2) The Brown University planktic foraminiferal database [@Prell1999].
@@
 4) The North American testate amoeba training set [@AMESBURY2018].
-5) Fossil pollen assemblage from the Neotoma database [@neotoma_2018] downloaded on `r targets::tar_meta(pollen_data) |> pull(time) |> format("%d %B %Y")`.
+5) Fossil pollen assemblage from the Neotoma database (`r neotoma_date` snapshot) [@neotoma_2018].
 
-Counts with with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded from all datasets.
+Counts with with small count sums (\< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded from all datasets.
 
@@
 
-Data sets with a many samples without singletons, a greatest common divisor greater than one, or an unusually high proportion, relative to the majority of pollen data set in the Neotoma database, of counts divisible by $k$ were flagged for closer inspection.
+Data sets with a many samples without singletons, a greatest common divisor greater than one, or an unusually high proportion of counts divisible by $k$, relative to the majority of pollen data set in the Neotoma database, were flagged for closer inspection.
 
@@
 In general, it is not possible to determine the sum from which percentages were calculated, but the properties of community and assemblage counts make it possible to estimate it.
-Given the expectation that the rarest taxon is a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
+Given the expectation that the rarest taxon is a singleton, the count sum *N* can be estimated as $1/p_{min} \times 100$ where *p~min~* is the percent abundance of the rarest taxon.
 This method will fail for assemblages without singletons.
-It is also expected that the percent _p_ is calculated from integer counts, therefore there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
-Possible values for _N_ can be found by a direct search over the range of plausible values of _N_.
-An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
+It is also expected that the percent *p* is calculated from integer counts, therefore there should be a count sum *N* such that $p_i/100\times N$ is, within rounding error, an integer for all *i* taxa.
+Possible values for *N* can be found by a direct search over the range of plausible values of *N*.
+An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of *N* except in cases where the greatest common divisor is greater than one.
 These tests for percent data are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.
@@
 
-All analyses were done in R v. `r getRversion()` [@R] and used the packages 
-extraDistr v. `r packageVersion("extraDistr")` [@extraDistr],
-numbers v. `r packageVersion("numbers")` [@numbers], 
-countSum v. `r packageVersion("countSum")` [@countSum], 
-neotoma2 v. `r packageVersion("neotoma2")` [@neotoma2], 
-bbsAssistant v. `r packageVersion("bbsAssistant")` [@bbsAssistant1], 
-and tidyverse v. `r packageVersion("tidyverse")` [@tidyverse].
-Packages targets v. `r packageVersion("targets")` [@targets], 
-quarto v. `r packageVersion("quarto")` [@bookdown1] and
-renv v. `r packageVersion("renv")` [@renv] were used to ensure reproducibility.
-Code to reproduce all the analyses is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms) and [zenodo.org](zenodo.org).
+All analyses were done in R v.
+`r getRversion()` [@R] and used the packages extraDistr v.
+`r packageVersion("extraDistr")` [@extraDistr], numbers v.
+`r packageVersion("numbers")` [@numbers], countSum v.
+`r packageVersion("countSum")` [@countSum], bbsAssistant v.
+`r packageVersion("bbsAssistant")` [@bbsAssistant1], and tidyverse v.
+`r packageVersion("tidyverse")` [@tidyverse].
+Packages targets v. `r packageVersion("targets")` [@targets], quarto v. `r packageVersion("quarto")` [@quarto] and renv v. `r packageVersion("renv")` [@renv] were used to ensure reproducibility.
+Code to reproduce all the analyses is archived at <https://github.com/richardjtelford/count.check.ms> and [zenodo.org](zenodo.org).
 All issues identified in the Neotoma database were reported to the maintainers.
@@
 While `r wildhorse$no_singletons * 100`% of assemblages in this dataset lack singletons, even more assemblage (`r wildhorse$no_zeros * 100`%) lack taxa with zero counts.
-This is in contrast to most (`r pollen_summ$has_zeros * 100`%) Neotoma pollen assemblages which have zero counts.  
+This is in contrast to most (`r pollen_summ$has_zeros * 100`%) Neotoma pollen assemblages which have zero counts.\
 These patterns could arise either when there is a high sampling effort in a low diversity site, or when the archived data do not include the entire pollen assemblage, instead focusing on the subset of abundant taxa of most interest to the authors.
@@
 These assemblages have between `r pollen1$min_taxa` and `r pollen1$max_taxa` taxa.
-The probability of all _n_ counts being divisible by _k_ is $k^{-n}$, assuming the distribution of counts is locally uniform.
-That is, for _k_ = 2, odd counts are as common as even counts.
+The probability of all *n* counts being divisible by *k* is $k^{-n}$, assuming the distribution of counts is locally uniform.
+That is, for *k* = 2, odd counts are as common as even counts.
 For an assemblage with `r pollen1$min_taxa` or `r pollen1$max_taxa` taxa, the probability of having all counts divisible by two by chance are 10^`r ceiling(- pollen1$min_taxa * log10(2))`^ and 10^`r ceiling(- pollen1$max_taxa * log10(2))`^ respectively.
 These probabilities are not so small that it is unrealistic to expect to find assemblages like these in the Neotoma database, but to find so many in one dataset is unexpected.
-Furthermore, the assumption made here is optimistic, in reality for _k_ = 2, more odd counts are expected ones are expected than even ones in most datasets, as typical rank abundance curves have more singletons than doubletons and so on, which makes finding assemblages with a greatest common divisor above one even more unlikely.
+Furthermore, the assumption made here is optimistic, in reality for *k* = 2, more odd counts are expected ones are expected than even ones in most datasets, as typical rank abundance curves have more singletons than doubletons and so on, which makes finding assemblages with a greatest common divisor above one even more unlikely.
 
 Dataset pollen2 has `r pollen2$n_sample` assemblages, of which `r words(pollen2$n_gcd2)` have a greatest common divisor of two and a further `r words(pollen2$n_gcd3)` with a greatest common divisor of three, one of which has `r pollen2$max_taxa3` taxa.
-The probability of having all counts divisible by three in such a taxonomically rich assemblage is 10^`r ceiling(- pollen2$max_taxa3 * log10(3))`.
+The probability of having all counts divisible by three in such a taxonomically rich assemblage is 10\^`r ceiling(- pollen2$max_taxa3 * log10(3))`.
 In addition to the assemblages with a greatest common divisor above one, this dataset includes several assemblages that have a higher proportion of counts divisible by two or three than is typical in assemblages from Neotoma (@fig-multiples-plot).
@@
 ## Unexpected percent data
+
 ```{r}
@@
 For example, each diatom cell has two valves which are the counting unit.
-These valves usually separate during processing, but paired valves are often found, and some taxa such as _Aulacoseira_ produce long chains of strongly bound valves.
+These valves usually separate during processing, but paired valves are often found, and some taxa such as *Aulacoseira* produce long chains of strongly bound valves.
 This might make singletons slightly less likely.
@@
 The methods in this paper may help estimate what the count sums are.
-If the actual count sums are materially below this, then action to correct the literature may be necessary, and if the omission is deliberate it 
-potentially constitutes misconduct [@Fanelli2013].
+If the actual count sums are materially below this, then action to correct the literature may be necessary, and if the omission is deliberate it potentially constitutes misconduct [@Fanelli2013].
 
@@
 
-# Acknowledgements{-}
+# Acknowledgements {.unnumbered}
 
@@
 
-# References{-}
+# References {.unnumbered}
